### 2021-10-28 introduction.tex
Abstract

\section{Introduction}

Programming geo-distributed replicated systems is hard given the complexity
of reasoning about constantly evolving states at different replicas.
Uncoordinated updates to replicas lead to states that are unlikely to
converge. A prominent approach to overcome this problem is to reorganize
the replicated state in terms of Commutative Replicated Data Types (CRDTs).
CRDTs require operations on the replicated state to commute, thereby
allowing such operations to be executed in different orders at replicas.
Applications composed of CRDTs therefore needn’t make assumptions about the
behavior of the underlying network; as long as the network eventually
delivers all in-flight messages, the replica states are guaranteed to
eventually converge. This guarantee is called Strong Eventual Consistency
(SEC). SEC is a sufficient condition to ensure eventual convergence of
replicated state. As such it provides a principled approach to build
distributed applications by composing CRDTs. It is however worthwhile to
note that SEC is not (proven to be) a necessary condition, and therefore
does not rule out the possibility of alternative approaches towards
convergent replicated state. Such an alternative approach is particularly
attractive if it frees application programming from the constrains of
commutativity.

The current work (SC MRDTs) proposes one such approach. The key idea behind
SC MRDTs is to decouple the orthogonal concerns of distributed applications
that are conflated by CRDTs, namely (1). Availability: the ability to
immediately apply updates to local replica without coordination, and (2).
Convergence: the ability to eventually apply updates at remote replicas in
a way that guarantees convergence. Note that while the Availability concern
explicitly rules out inter-replica coordination, Convergence doesn’t. In
other words, it is permissible for replicas to coordinate among themselves
to achieve convergent replication, as long as such coordination doesn’t
make a replica unavailable to its users.  SC MRDTs exploit this
observation. Like CRDTs, SC MRDTs allow updates from user-submitted
operations to be applied immediately to the local replica to minimize the
user-perceived latency. Unlike CRDTs however, SC MRDTs do not allow updates
to be applied at remote replica at arbitrary time and in arbitrary order.
Instead, the system underlying SC MRDTs carefully orchestrates the
distributed execution so as to merge the competing states at different
replicas in an order that guarantees convergence. Thus, any data type with
a merge operation qualifies to be a replicated data type without having to
re-engineer its operations around the commutativity principle.

### 2021-10-28 quark.tex
Abstract

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,10pt,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{xspace}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{ifpdf}
\usepackage{graphicx}
%\usepackage[usenames,dvipsnames]{color}
\usepackage{stmaryrd}
%\usepackage[numbers]{natbib}
\usepackage{amsthm}
\usepackage{listings}          % format code
\usepackage{wrapfig}
\usepackage{textcomp}
\usepackage{tabularx}
\usepackage{color}
\usepackage{url}
\usepackage{tikz}
\usepackage{multirow,array}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}

\input{mydefs.tex}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2022}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format must update the '\documentclass' and
%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% Title information
\title[Convergence Sans Commutativity]{Convergence Sans Commutativity:
Mergeability as a basis for Convergent Replicated Data Types}      %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{First1 Last1}
\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}                    %% \country is recommended
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}                   %% \country is recommended
}
\email{first2.last2@inst2a.com}         %% \email is recommended
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}                   %% \country is recommended
}
\email{first2.last2@inst2b.org}         %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
   Programming geo-distributed replicated systems is hard given the
   complexity of reasoning about constantly evolving states at different
   replicas. Uncoordinated updates to replicas lead to states that are
   unlikely to converge. In the absence of inter-replica coordination,
   commutativity has come to be accepted as the \emph{de facto} design
   principle to guarantee convergence. Unfortunately, designing replicated
   state updates as commutative operations is hard considering that most
   data types define operations that do not commute. Deriving a replicated
   variant of a data type often entails a creative re-engineering of its
   internal representation and algorithms so as to support commutative
   operations while retaining (close to) the original semantics. In this
   paper we propose an alternative approach to replicated data types that
   avoids such complex re-engineering. Our approach is fundamentally
   different in that it is based on the weaker notion of
   \emph{mergeability}. Unlike commutativity, mergeability only requires a
   data type to define \emph{a} merge semantics while leaving its existing
   definition intact. Notably, our approach leaves type and merge semantics
   completely unconstrained even as it guarantees eventual convergence and
   full availability in an asynchronous distributed setting. In other
   words, there is no need for the programmers to prove algebraic
   properties of their implementations in order to obtain the convergence
   and availability guarantees under a distributed execution. Such
   weakening of obligations is made possible by a novel distributed runtime
   that orchestrates distributed executions as per a structural
   well-formedness criterion, which we show is sufficient to guarantee
   convergence. We implement the runtime called \quark as a thin shim layer
   on top of an off-the-shelf distributed database, and use it to run
   mergeable replicated variants of several ordinary data types. Our
   evaluation categorically demonstrate the generality of our approach and
   the low overhead imposed by our runtime.

\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\input{introduction}

Text of paper \ldots

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
%\bibliography{bibfile}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2021-10-31 introduction.tex
Introduction added

@@
 
-Programming geo-distributed replicated systems is hard given the complexity
-of reasoning about constantly evolving states at different replicas.
-Uncoordinated updates to replicas lead to states that are unlikely to
-converge. A prominent approach to overcome this problem is to reorganize
-the replicated state in terms of Commutative Replicated Data Types (CRDTs).
-CRDTs require operations on the replicated state to commute, thereby
-allowing such operations to be executed in different orders at replicas.
-Applications composed of CRDTs therefore needn’t make assumptions about the
-behavior of the underlying network; as long as the network eventually
-delivers all in-flight messages, the replica states are guaranteed to
-eventually converge. This guarantee is called Strong Eventual Consistency
-(SEC). SEC is a sufficient condition to ensure eventual convergence of
-replicated state. As such it provides a principled approach to build
-distributed applications by composing CRDTs. It is however worthwhile to
-note that SEC is not (proven to be) a necessary condition, and therefore
-does not rule out the possibility of alternative approaches towards
-convergent replicated state. Such an alternative approach is particularly
-attractive if it frees application programming from the constrains of
-commutativity.
+Large-scale web services and decentralized applications often rely on
+geo-distributed state replication to meet their latency and availability
+needs. An application's state is replicated asynchronously in such a
+setting, meaning that operations are independently executed at each
+replica, and updates are applied asynchronously at other replicas after a
+possible network delay. Asynchronous execution complicates programming and
+reasoning about distributed applications as it induces the possibility of
+conflicting updates leading to non-convergence and application integrity
+violations. Two basic approaches have been proposed to address this
+problem. The first approach is to selectively strengthen the system
+consistency to pre-empt the conflicting updates. The second is to redefine
+the application state in terms of \emph{Replicated Data Types} (RDTs) that
+are specially engineered to handle conflicting updates. Strengthening
+system consistency necessarily entails inter-replica coordination,
+therefore applications prefer to utilize RDTs as much as possible,
+resorting to consistency strengthening only when it is necessary to maintain
+application integrity. The common design principle guiding the development
+of replicated data types is commutativity. The idea is that if the
+replicated state is only updated by commutative operations, then updates
+can be applied in any order and the replica states are still guaranteed to
+converge. Indeed, there exist common usecases in web applications that can
+be implemented using Commutative Replicated Data Types
+(CRDTs)~\cite{crdts}. For instance, a video view counter on a streaming
+application such as YouTube can be implemented as a Counter CRDT that
+supports commutative increments.
 
-The current work (SC MRDTs) proposes one such approach. The key idea behind
-SC MRDTs is to decouple the orthogonal concerns of distributed applications
-that are conflated by CRDTs, namely (1). Availability: the ability to
-immediately apply updates to local replica without coordination, and (2).
-Convergence: the ability to eventually apply updates at remote replicas in
-a way that guarantees convergence. Note that while the Availability concern
-explicitly rules out inter-replica coordination, Convergence doesn’t. In
-other words, it is permissible for replicas to coordinate among themselves
-to achieve convergent replication, as long as such coordination doesn’t
-make a replica unavailable to its users.  SC MRDTs exploit this
-observation. Like CRDTs, SC MRDTs allow updates from user-submitted
-operations to be applied immediately to the local replica to minimize the
-user-perceived latency. Unlike CRDTs however, SC MRDTs do not allow updates
-to be applied at remote replica at arbitrary time and in arbitrary order.
-Instead, the system underlying SC MRDTs carefully orchestrates the
-distributed execution so as to merge the competing states at different
-replicas in an order that guarantees convergence. Thus, any data type with
-a merge operation qualifies to be a replicated data type without having to
-re-engineer its operations around the commutativity principle.
+In general, however, commutativity is not a common occurrence among data
+types. Most common data type definitions come with at least a pair of
+operations that do not commute. For instance an \C{insert} and a \C{remove}
+operations on a set do not commute if both are for the same element.
+Likewise two \C{insert} operations for the same position in a list do not
+commute. Such non-commutative operations translate to conflicting updates
+in a distributed setting. For instance, concurrent \C{insert $k$} and
+\C{remove $k$} operations for an element $k$ in a replicated set are in
+conflict, and it is not clear what their cumulative result should
+be\footnote{Here, the \C{insert} $i$ is assumed to have witnessed a
+  \C{remove} $r'$ that is distinct from the earlier \C{remove} $r$. Both
+  $i$ and $r'$ are concurrent to $r$, hence their effects are not visible
+  during $r$'s execution.}. To use such a non-commutative data type in a
+replicated setting, creative re-engineering of its internal representation
+and algorithms is needed to turn it into a bonafide CRDT with a sensible
+distributed semantics. In the case of set data type, that could involve
+changing the internal representation to maintain two sets -- a set
+containing the current elements and a \emph{tombstone} set that contains
+every element that has ever been deleted from the set. A \C{remove
+$k$} operation removes $k$ from the element set and adds it to the
+tombstone set, and an \C{insert $k$} operation adds $k$ to the element set
+only if it is not present in the tombstone set (i.e., only if $k$ was not
+previously deleted from the set). This results in a CRDT set called
+\emph{two-phase set} where an element once removed cannot be re-added.
+
+A more complex re-engineering of the set data type could result in a set
+CRDT with a different (and arguably more useful) semantics.
+Sec.~\ref{sec:motivation} describes a \emph{remove-wins} CRDT set, which
+allows elements to be readded, but which prioritizes \C{remove}s over
+concurrent \C{insert}s. This transformation requires vector
+clocks~\cite{vectorclock} to keep track of casual ordering among
+\C{insert}s and \C{remove}s. In fact most CRDTs have carefully-crafted
+implementations involving non-trivial technical devices like vector
+clocks. For instance, Replicated Growable Array (RGA) -- a CRDT for
+collaborative editing~\cite{rga}, makes use of a variant of Lamport
+timestamps~\cite{lamportts}; TreeDoc, another collaborative-editing CRDT,
+makes use of \emph{dense linear orders}~\cite{treedoc}; and JSON CRDT -- a
+CRDT variant of the JSON storage format, implements transactional semantics
+and causal dependency sets~\cite{jsoncrdt}. Such complexity makes it very
+hard to reason about basic correctness properties of CRDTs, such as
+convergence (formally known as \emph{strong eventual
+consistency}~\cite{crdts}).  Indeed, formal verification of strong eventual
+consistency for a subset of the aforementioned CRDTs required considerable
+effort on behalf of multiple authors who are experts in the
+field~\cite{kleppmann2017}. The need for such advanced implementation and
+reasoning techniques makes the development of new CRDTs a research exercise
+to be undertaken only by experts.
+
+% hinders the widespread adoption of replicated data
+% types in software development.
+In this paper we propose an alternative approach to replicated data types
+that addresses the problems discussed above. Our approach differs from the
+CRDT approach in two fundamental ways. Firstly, we adopt a state-centric
+model of replication with \emph{mergeability} of states as the defining
+characteristic in contrast to CRDTs' operation-centric model characterized
+by commutativity. Unlike commutativity, mergeability does not require a
+full-scale re-engineering of the data type definition. Instead,
+a data type only has to define a merge semantics in form a \C{merge}
+function to reconcile concurrent versions of the type in presence of their
+common ancestor version. Notably, \C{merge} function is left completely
+unconstrained even as convergence is guaranteed on the corresponding
+\emph{mergeable replicated data type} (MRDT)~\cite{mrdt}. Such strong
+guarantees are possible due to the simplicity of state-based reasoning
+relative to the operation- or trace-based one. Note that state-centric
+model of replication has also been adapted to CRDTs, but such state-based
+CRDTs require the replicated state to be organized as a lattice to
+guarantee convergence -- a restriction we eliminate in our setting.
+
+The second distinguishing aspect of our approach is the utilization of a
+distributed runtime that orchestrates \emph{well-formed} distributed
+executions that are guaranteed to converge. Note that it is easy to
+construct a trivial runtime that guarantees convergence of all executions.
+Such a runtime would make an extensive use of inter-replica coordination to
+synchronize the execution of every operation and thus induce a
+sequentially-consistent behavior. This would however defeat the purpose of
+state replication as the latency incurred for synchronized execution of an
+operation is prohibitively high, and the availability of the system during
+the execution is correspondingly low. It is therefore important for a
+distributed runtime of RDTs to not synchronize the execution of RDT
+operations. Orchestrating a well-formed convergent execution without
+impacting latency and availability may seem impossible due to the CAP
+theorem~\cite{cap}, but it is not the case in fact. While CAP theorem does
+rule out linearizability, it does not prohibit enforcing weaker consistency
+constraints on distributed executions. In fact many CRDTs, including the
+aforementioned \emph{remove-wins} set, implicitly assume \emph{causal
+consistency}, which is weaker than linearizability but is stronger than the
+default \emph{eventual consistency}. While causal consistency alone cannot
+guarantee convergence, it sufficiently constrains the distributed execution
+for the CRDT implementations to guarantee convergence. In this respect our
+  approach differs in the sense that (a). It identifies a novel set of
+  constraints on distributed execution that guarantee convergence of an
+  MRDT \emph{regardless} of its merge semantics, and (b). It localizes
+  those constraints spatially and temporally into the runtime such that
+  per-operation latency and system-wide availability remain unaffected.
+Sec.~\ref{sec:semantics} explains our runtime in detail.
+
+\paragraph{Contributions} The contributions of this paper are summarized
+below:
+\begin{itemize}
+  \item We present a model of state replication for \emph{mergeable} data
+  types, i.e., data types equipped with a \C{merge} function, and identify a
+  structural well-formedness condition over distributed executions that is
+  sufficient to guarantee their convergence. This is in contrast to
+  existing models of replicated data types, where each type is obligated to
+  prove its convergence.
+
+  \item We formalize a distributed runtime called \quark that orchestrates
+  well-formed distributed executions. We prove the correctness of \quark,
+  and also demonstrate its \emph{progress} property, namely that no
+  distributed execution would ever get ``stuck''. Besides the manual proofs
+  for the theorems, we also provide the full formalization of the \quark
+  system and its properties in Alloy~\cite{alloy} and Ivy~\cite{ivy}. The
+  Ivy formalization of \quark has been automatically verified with help of
+  Z3~\cite{z3}. To the best of our knowledge, this is the first time an SMT
+  solver was used to verify the \emph{meta-theory} of a formal setup.
+
+  \item We realize an implementation of \quark as a light-weight shim layer
+  on top of Scylla -- an off-the-shelf distributed key-value
+  store~\cite{scylla}. An evaluation over several mergeable replicated data
+  types (MRDTs) demonstrate that \quark has \emph{no} effect on the latency
+  and availability of the system. To demonstrate the trade-offs not
+  captured by latency and availability, we invent a new metric called
+  \emph{staleness} and show that our approach does affect it negatively, albeit
+  within the tolerable limits.
+\end{itemize}

### 2021-11-12 introduction.tex
Motivation in progress

@@
 \section{Introduction}
+\label{sec:intro}
 
@@
 approach differs in the sense that (a). It identifies a novel set of
-  constraints on distributed execution that guarantee convergence of an
-  MRDT \emph{regardless} of its merge semantics, and (b). It localizes
-  those constraints spatially and temporally into the runtime such that
+constraints on distributed execution that guarantee the convergence of an
+MRDT \emph{regardless} of its merge semantics, and (b). It localizes those
+constraints spatially and temporally into the runtime such that
 per-operation latency and system-wide availability remain unaffected.
@@
   on top of Scylla -- an off-the-shelf distributed key-value
-  store~\cite{scylla}. An evaluation over several mergeable replicated data
-  types (MRDTs) demonstrate that \quark has \emph{no} effect on the latency
-  and availability of the system. To demonstrate the trade-offs not
-  captured by latency and availability, we invent a new metric called
-  \emph{staleness} and show that our approach does affect it negatively, albeit
-  within the tolerable limits.
+  store~\cite{scylla}. We do a thorough evaluation with a
+    collaborative-editing benchmark to demonstrate that \quark has
+    \emph{no} effect on the latency and availability of the system, while
+    slightly increasing the time to convergence.
 \end{itemize}

### 2021-11-12 motivation.tex
Motivation in progress

\section{Motivation}
\label{sec:motivation}

In this section we motivate our runtime-assisted approach to
convergent RDTs with help of a set data structure.
% demonstrates the need for run-time intervention to ensure convergence.

\subsection{Add-Wins Set RDT}

Fig.~\ref{fig:ocaml-set} shows a simplified interface of a set data
type in OCaml. The interface hides a reference to a set (\C{Set.t}),
which can be updated in-place via \C{insert} and \C{remove}
operations. As such \C{Set} is an ordinary data type primarily
designed for sequential execution. Under a concurrent execution with
an asynchronously replicated state, \C{Set} would exhibit anamolous
behaviours such as those in Fig.~\ref{fig:crdt-execs}.

Fig.~\ref{fig:crdt-execs-1} shows an anamolous execution with three
replicas -- $R_1$, $R_2$, and $R_3$, all of which start with a
singleton set containing the element $e$. A client connects to the
replica $R_3$ and executes a \C{remove $e$} operation, which is then
asynchronously propagated to other replicas. Some time after applying
$R_3$'s \C{remove} at $R_2$, another client connects to $R_2$ and
re-adds $e$ by issuing an \C{insert $e$} operation. Consequently, the
state at $R_2$ is again the singleton set $\{e\}$. Replica $R_3$
however receives $R_2$'s \C{insert} ahead of $R_1$'s \C{remove},
applies them in the same order, and ends up with an empty set. The
execution thus results in divergent replica states.

Note that if set \C{insert} and \C{remove} commute, then executing
them in different order at $R_2$ and $R_3$ would not have led to
divergence. As such, \C{Set} is not a CRDT due to the admittance of
non-commutative operations. Nonetheless \C{Set} can be transformed
into a CRDT by re-engineering its interface and operations. For
instance, the anomalous execution in Fig.~\ref{fig:crdt-execs-1} can
be pre-empted by ensuring that updates are only ever applied in the
causal order.  This can be done by extending \C{Set} with vector
clocks to keep track of the causal history of each operation. A set
\C{insert} (resp. \C{remove}) would now generate an \C{Insert} (resp.
\C{Remove}) \emph{effect} tagged with the vector clock of the origin
replica. Here, the vector clock simply records the sequence number of
last operation from each replica whose effect has been received and
applied at the current replica. When an effect is received at a
replica, it is buffered until the time all its causally-preceding
effects (as captured by the tagged vector clock) have already been
received and applied. This strategy would preempt the execution in
Fig.~\ref{fig:ctrdt-execs-1} by buffering $R_2$'s \C{insert} at $R_1$
until the causally-preceding \C{remove} of $R_1$ is received and
applied. An interface for such a \emph{causally-consistent} set RDT is
shown in Fig.~\ref{fig:cc-set}.

Unfortunately, \C{Set} data type of Fig.~\ref{fig:cc-set} still admits
divergent executions due to concurrent updates.
Fig.~\ref{fig:crdt-execs-2} describes one such execution. Here,
replicas $R_1$ and $R_2$ both start with a singleton set $\{e\}$.
Two distinct clients connect to $R_1$ and $R_2$ respectively and issue
two concurrent \C{remove $e$} operations. Later, another
client connects to $R_2$ and issues an \C{insert $e$} operation. The
effects of these operations are asynchronously applied at remote
replicas as shown in the figure, resulting in the divergent states at
$R_1$ and $R_2$.

Note that, unlike the execution in Fig.~\ref{fig:crdt-execs-1}, the
conflicting operations in Fig.~\ref{crdt-execs-2}, namely $R_1$'s
\C{remove} and $R_2$'s \C{insert}, are \emph{not} causally related,
hence their relative order is not determined by the sequential
specification of the data type. Forcing a causal relationship between
them requires synchronization between \C{insert}s and \C{remove}s,
which is expensive in an asynchronous distributed setting. It
therefore becomes inevitable to ascribe semantics to concurrent
executions to restore convergence. This is done by imposing an
\emph{arbitration order} among concurrent conflicting operations,
which are otherwise unordered. For example, in
Fig.~\ref{fig:crdt-execs-2}, we might let $R_2$'s \C{insert} override
$R_1$'s \C{remove} considering that \C{insert} is re-inserting an
element after a previous remove. Ordering concurrent removes ahead
of inserts uniformly on all replicas leads to an implementation of
\C{Set} RDT where concurrent (re-)insertions consistenty \emph{win}
over removals. Such an \emph{add-wins} set is useful, for instance, to
implement a shared shopping cart where two users can concurrently
remove an item from the cart, but if one of them re-adds it then the
item should be present in the final cart\footnote{This is in fact the
semantics of Amazon's shopping cart~\cite{dynamo}}.

Extending the \C{Set} implementation from Fig.~\ref{fig:cc-set} with
\emph{add-wins} semantics is however not trivial. The suggested
approach involves tracking element-wise causal dependencies between
the \C{remove}s and \C{insert}s by maintaining a vector clock
\emph{for each element} $e$ in the set~\cite{zawirski-thesis}. The
vector clock of $e$ records the sequence number of the last \C{insert
$e$} operation from each replica whose effect has been received and
applied at the current replica. The \C{Insert} and \C{Remove} effects
on $e$ will now be tagged with $e$'s vector clock. When an \C{Remove}
effect on $e$ is received at a replica, it is applied only if the
tagged vector clock is no less than $e$'s local vector clock, i.e.,
only if the arriving \C{Remove} has seen (i.e., causally succeeds) at
least those \C{insert $e$} operations the current replica is aware of.
Otherwise the effect is simply a no-op. This strategy would result in
convergent states despite the execution in Fig.~\ref{fig:crdt-execs-2}
as $R_2$'s \C{remove} effectively becomes a no-op at $R_1$ due to
there being at least one \C{insert} operation on $e$ that it hasn't
seen. The \C{Set} interface extended with element-wise vector clocks
is shown in Fig.~\ref{fig:rw-set}. The resultant set RDT is assumed to
be correct albeit a formal proof of convergence could not be found in
the literature.

The above exercise demonstrates the considerable ingenuity and effort
involved in deriving a convergent RDT out of such simple data type as
\C{Set}. Some of the effort can be mitigated by strengthening the
underlying system model insofar as it doesn't affect the latency and
availability of the application. For instance, a system that always
delivers messages in the causal order would automatically preempt the
execution in Fig.~\ref{fig:crdt-execs-1} without the need for
additional intervention on behalf of the developer. This is a
particularly attractive proposition considering that causal
consistency can be ``bolted on'' an existing implementation of an
eventually consistent system without weakening its
guarantees~\cite{bolton}. Unfortunately, such strengthening of the
system model would deliver no benefits to the developer if they still
have to reason about fine-grained causal dependencies to guarantee
convergence, such as in the case of Fig.~\ref{fig:crdt-execs-2}. As we
observed earlier, the execution in Fig.~\ref{fig:crdt-execs-2} seems
inevitable unless every pair of \C{insert} and \C{remove} operations
are synchronized, which, regrettably, is not a practical option. The
developer therefore seems to be stuck.

Fortunately, there is a way out of this problem and the solution
requires a switch in the perspective of replication from an
operation-centric view characterized by explicit effects to a
state-centric view characterized by explicit state merges. The
motivation for this switch is the key observation that it is possible
to induce an ordering over \emph{states} of replicas even as their
operations and effects remain unordered, as long as such states are
\emph{mergeable}. The ordering is enforced by synchronizing merges in
the background while the replica-local execution progresses
unhindered. Consequently, eventual convergence can be guaranteed
\emph{without} impacting the user-perceived latency of the operations.
Moreover, working with states instead of operations frees the latter
from having to conform to algebraic laws such as commutativity.  Put
together, this means that the clients of a replicated data type can
perform any action allowed by the sequential version of the type and
immediately see the results of their actions on the local replica
while being safe in the knowledge that replica states will eventually
converge.

To demonstrate the aforementioned intuitions we reconsider the
executions from Fig.~\ref{fig:crdt-execs}, this time in the
state-centric replication model. The cornerstone of the state-centric
model is a three-way \C{merge} function that merges concurrent
versions of the state in presence of their (lowest) common ancestor
version. In our running example, the state is a value of type
\C{Set.t}, hence \C{Set.merge} function would have the type signature:
\begin{center}
\C{Set.merge} : \C{Set.t} $\rightarrow$ \C{Set.t} $\rightarrow$ \C{Set.t}
$\rightarrow$ \C{Set.t}
\end{center}
The three arguments of \C{merge} correspond to the lowest common
ancestor (LCA) version and the two concurrent versions that
independently evolved from the ancestor version. The LCA is a causal
ancestor of concurrent versions, hence causal consistency is built
into the replication model. The result of set merge, intuitively, must
contain the common elements in the two concurrent versions along with
any newly added elements in either versions. Concretely:
\begin{ocaml}
let merge s s1 s2 = 
              (s1 $\cap$ s2) $\cup$ (s1 - s) $\cup$ (s2 - s)
\end{ocaml}
Equipped with the above set merge function, we can now consider the
equivalent execution of Fig.~\ref{fig:crdt-execs-1} in state-centric
model. The execution is shown in Fig.~\ref{fig:mrdt-execs-1}. The
initial version on all three replicas ($v_0$) is the singleton set
$\{e\}$. Applying operations to replicas creates new versions, e.g.,
$v_1$ on $R_3$. Changes can be propagated by merging versions, e.g.,
version $v_2$ on $R_2$ is a result of merging $v_0$ and $v_1$ in
presence of their LCA $v_0$\footnote{Versions $v_0$ and $v_1$ are not
  concurrent as the former is an ancestor of the latter. Merging $v_1$
  into $v_0$ is nonetheless possible as $v_1$ is ahead of $v_0$ in
  causal order. In Git parlance this is a \emph{fast forward} merge.}.
Likewise $v_4$ on $R_1$ is created by merging $v_3$ and $v_0$ in
presence of their LCA $v_0$. By the end of the execution, versions $v_4$
and $v_3$ at $R_1$ and $R_2$ (resp.) have witnessed the same set of
operations, hence are in agreement.

Convergence however is not an inherent virtue of the state-centric
replication model. Fig.~\ref{fig:mrdt-exec-2} shows the state-centric
analogue of the execution in Fig.~\ref{fig:crdt-exec-2}, which too
diverges.

### 2021-11-13 motivation.tex
Motivation text complete

@@
 ancestor (LCA) version and the two concurrent versions that
-independently evolved from the ancestor version. The LCA is a causal
+independently evolved from the LCA version. The LCA is a causal
 ancestor of concurrent versions, hence causal consistency is built
@@
 \begin{ocaml}
-let merge s s1 s2 = 
+  let merge s(*lca*) s1 s2 = 
               (s1 $\cap$ s2) $\cup$ (s1 - s) $\cup$ (s2 - s)
@@
 version $v_2$ on $R_2$ is a result of merging $v_0$ and $v_1$ in
-presence of their LCA $v_0$\footnote{Versions $v_0$ and $v_1$ are not
-  concurrent as the former is an ancestor of the latter. Merging $v_1$
-  into $v_0$ is nonetheless possible as $v_1$ is ahead of $v_0$ in
-  causal order. In Git parlance this is a \emph{fast forward} merge.}.
+presence of their lowest common ancestor (LCA) version
+$v_0$\footnote{Versions $v_0$ and $v_1$ are not concurrent as the
+  former is an ancestor of the latter. Merging $v_1$ into $v_0$ is
+  nonetheless possible as $v_1$ is ahead of $v_0$ in causal order. In
+  Git parlance this is a \emph{fast forward} merge.}.
 Likewise $v_4$ on $R_1$ is created by merging $v_3$ and $v_0$ in
-presence of their LCA $v_0$. By the end of the execution, versions $v_4$
-and $v_3$ at $R_1$ and $R_2$ (resp.) have witnessed the same set of
-operations, hence are in agreement.
+presence of their LCA $v_0$. By the end of the execution, versions
+$v_4$ and $v_3$ at $R_1$ and $R_2$ (resp.) have witnessed the same set
+of operations, hence are in agreement.
 
@@
 replication model. Fig.~\ref{fig:mrdt-exec-2} shows the state-centric
-analogue of the execution in Fig.~\ref{fig:crdt-exec-2}, which too
-diverges.
+analogue of the execution in Fig.~\ref{fig:crdt-exec-2} which
+diverges. Here $R_1$ and $R_2$ start with version $v_0 \,=\, \{e\}$.
+$R_2$ performs a \C{remove} and and an \C{insert} making versions
+$v_1$ and $v_2$ respectively. Simultaneously, $R_1$ performs a
+\C{remove} to make $v_3$. Replica $R_1$ now obtains $R_2$'s changes by
+merging versions $V_1$ and $v_2$ to make new versions $v_4$ and $v_5$
+respectively. LCAs for these merges are $v_0$ and $v_1$. Concurrently,
+$R_2$ obtains $R_1$'s changes by merging $v_3$ with $v_2$ (LCA =
+$v_0$) to make $v_6$.  Now $R_1$ and $R_2$ have same set of changes
+yet their final versions differ.
+
+Note that the anamolous execution in Fig.~\ref{fig:mrdt-execs-2} could
+have been avoided had the merges between $R_1$ and $R_2$ been
+linearized. Fig.~\ref{fig:mrdt-execs-3} shows an execution that only
+slightly differs from the one in Fig.~\ref{fig:mrdt-execs-2}. The
+difference is that the merges in Fig.~\ref{fig:mrdt-execs-3} happen
+linearly: first $R_1$ is merged into $R_2$ (bringing $R_1$'s
+\C{remove}), then $R_2$ into $R_1$ (bringing $R_2$'s \C{remove}),
+followed again by $R_2$ into $R_1$ (bringing $R_2$'s \C{insert}). As a
+result of such linearization, the final versions on $R_1$ and $R_2$
+converge to the singleton set $\{e\}$. Note that there exist other
+linearizations of merges; for instance, $R_1 \,\rightarrow\, R_2$
+merge could be ordered between the two $R_2 \rightarrow R_1$ merges.
+However, all linearizations result in the same final state $\{e\}$.
+Another important point to note is that only the merges are
+linearized; not the entire execution. In Fig.~\ref{fig:mrdt-execs-3},
+both $R_1$ and $R_2$'s \C{remove}s remove the same element $e$, which
+is not possible in a linearized execution. Leaving the execution
+unconstrained is crucial to satisfy the limits imposed by the CAP
+theorem on a highly-available partition-tolerant distributed system.
+
+In the context of Fig.~\ref{fig:mrdt-execs-3}, it is quite clear what
+linearization of merges means and how to enforce it via
+synchronization (for e.g., wrapping each merge within a global lock).
+In general however, the semantics of merge linearization isn't as cut
+and dried. For instance, consider the execution in
+Fig.~\ref{fig:mrdt-execs-4}. The four replicas involved in the
+execution start with version $v_0 \,=\, \{e\}$. The replicas perform
+local operations as shown in the figure to make versions $v_1$ to
+$v_5$. Next they perform a series of merges to propagate local
+changes. The merges can be ordered in time as following: first $R_2
+\rightarrow R_1$, then $R_3 \rightarrow R_1$, then $R_2 \rightarrow
+R_4$, and finally $R_3 \rightarrow R_4$. Despite linearly ordered in
+time, the merges nonetheless result in divergent execution. The
+problem here is that, although merges are executed linearly, the
+execution graph does not reflect this linearity; merges of that end in
+$R_1$ in Fig.~\ref{fig:mrdt-execs-4} are effectively concurrent with
+those that end in $R_4$. This shows that simply synchronizing the
+execution of merges does not necessarily result in convergence.
+
+Our key insight to overcome this impasse is a \emph{well-formedness}
+condition on execution graphs that ensures convergence of final
+states. To understand well-formedness, let us contrast the bad
+executions in Figs.~\ref{fig:mrdt-execs-2} and~\ref{fig:mrdt-execs-4}
+against the good execution in Fig.~\ref{fig:mrdt-execs-3}. Observe
+that in Fig.~\ref{fig:mrdt-execs-2}, \emph{every} pair of concurrent
+versions on $R_1$ and $R_2$ have a unique lowest common ancestor
+(LCA). For instance, LCA of $(v_5,v_6)$ is $v_6$, $(v_5,v_4)$ is
+$v_3$, $(v_1,v_2)$ is $v_0$ and so on. By contrast in
+Fig.~\ref{fig:mrdt-execs-2}, versions $v_5$ and $v_6$ have two LCAs,
+namely $v_2$ and $v_3$. Both these versions are common ancestors of
+$v_5$ and $v_6$, and both are \emph{lowest} in the sense that there do
+not exist versions lower (in the execution graph) than $v_2$ and $v_3$
+that are also common ancestors of $v_5$ and $v_6$. Likewise in
+Fig.~\ref{fig:mrdt-execs-4}, versions $v_7$ and $v_9$ have two LCAs --
+$v_2$ and $v_4$. Multiple LCAs is an indication that there exist
+merges prior in the execution graph that are effectively concurrent.
+Considering that our approach to convergence in state-centric model
+crucially relies on linearity of merges, presence of multiple LCAs
+opens up the possibility of divergence. We therefore define
+well-formed execution graphs as those where LCAs for every pair of
+concurrent versions is unique. As we prove in
+Sec.~\ref{sec:formalization}, enforcing this structural
+well-formedness condition indeed guarantees the convergence of
+distributed executions. In Sec.~\ref{fig:implementation}, we describe
+a distributed runtime called \quark whose primary purpose is to
+enforce the well-formedness condition on distributed executions. The
+runtime makes it possible to automatically promote ordinary data types
+equipped with a merge function to convergent replicated data types.
+Thus, extending the \C{Set} interface of Fig.~\ref{fig:ocaml-set} with
+the two-line \C{Set.merge} function shown above is sufficient to
+obtain a convergent replicated set data type.
+
+The following section formalizes the aforementioned intuitions with
+help of an abstract machine that generates only the well-formed
+executions.

### 2021-11-14 introduction.tex
Figures in motivation; edits to intro

@@
 
-In general, however, commutativity is not a common occurrence among data
-types. Most common data type definitions come with at least a pair of
-operations that do not commute. For instance an \C{insert} and a \C{remove}
-operations on a set do not commute if both are for the same element.
-Likewise two \C{insert} operations for the same position in a list do not
-commute. Such non-commutative operations translate to conflicting updates
-in a distributed setting. For instance, concurrent \C{insert $k$} and
-\C{remove $k$} operations for an element $k$ in a replicated set are in
-conflict, and it is not clear what their cumulative result should
-be\footnote{Here, the \C{insert} $i$ is assumed to have witnessed a
-  \C{remove} $r'$ that is distinct from the earlier \C{remove} $r$. Both
-  $i$ and $r'$ are concurrent to $r$, hence their effects are not visible
-  during $r$'s execution.}. To use such a non-commutative data type in a
-replicated setting, creative re-engineering of its internal representation
-and algorithms is needed to turn it into a bonafide CRDT with a sensible
-distributed semantics. In the case of set data type, that could involve
-changing the internal representation to maintain two sets -- a set
-containing the current elements and a \emph{tombstone} set that contains
-every element that has ever been deleted from the set. A \C{remove
-$k$} operation removes $k$ from the element set and adds it to the
-tombstone set, and an \C{insert $k$} operation adds $k$ to the element set
-only if it is not present in the tombstone set (i.e., only if $k$ was not
-previously deleted from the set). This results in a CRDT set called
-\emph{two-phase set} where an element once removed cannot be re-added.
-
-A more complex re-engineering of the set data type could result in a set
-CRDT with a different (and arguably more useful) semantics.
-Sec.~\ref{sec:motivation} describes a \emph{remove-wins} CRDT set, which
-allows elements to be readded, but which prioritizes \C{remove}s over
-concurrent \C{insert}s. This transformation requires vector
-clocks~\cite{vectorclock} to keep track of casual ordering among
-\C{insert}s and \C{remove}s. In fact most CRDTs have carefully-crafted
-implementations involving non-trivial technical devices like vector
-clocks. For instance, Replicated Growable Array (RGA) -- a CRDT for
-collaborative editing~\cite{rga}, makes use of a variant of Lamport
-timestamps~\cite{lamportts}; TreeDoc, another collaborative-editing CRDT,
-makes use of \emph{dense linear orders}~\cite{treedoc}; and JSON CRDT -- a
-CRDT variant of the JSON storage format, implements transactional semantics
-and causal dependency sets~\cite{jsoncrdt}. Such complexity makes it very
-hard to reason about basic correctness properties of CRDTs, such as
-convergence (formally known as \emph{strong eventual
-consistency}~\cite{crdts}).  Indeed, formal verification of strong eventual
-consistency for a subset of the aforementioned CRDTs required considerable
-effort on behalf of multiple authors who are experts in the
-field~\cite{kleppmann2017}. The need for such advanced implementation and
-reasoning techniques makes the development of new CRDTs a research exercise
-to be undertaken only by experts.
+In general, however, commutativity is not a common occurrence among
+data types. Most common data type definitions come with at least a
+pair of operations that do not commute. For instance an \C{add} and a
+\C{remove} operations on a set do not commute if both are for the same
+element.  Likewise two \C{insert} operations for the same position in
+a list do not commute. Such non-commutative operations translate to
+conflicting updates in a distributed setting, whose cumulative result
+cannot be determined by the sequential specification of the data type
+alone. To use a non-commutative data type in a replicated setting,
+creative re-engineering of its internal representation and algorithms
+is needed to turn it into a bonafide CRDT with a sensible distributed
+semantics.
+% In the case of set data type, that could involve changing the internal
+% representation to maintain two sets -- a set containing the current
+% elements and a \emph{tombstone} set that contains every element that
+% has ever been deleted from the set. A \C{remove $k$} operation removes
+% $k$ from the element set and adds it to the tombstone set, and an
+% \C{insert $k$} operation adds $k$ to the element set only if it is not
+% present in the tombstone set (i.e., only if $k$ was not previously
+% deleted from the set). This results in a CRDT set called
+% \emph{two-phase set} where an element once removed cannot be re-added. 
+% A more complex re-engineering of the set data type could result in a set
+% CRDT with a different (and arguably more useful) semantics.
+% Sec.~\ref{sec:motivation} describes a \emph{remove-wins} CRDT set, which
+% allows elements to be readded, but which prioritizes \C{remove}s over
+% concurrent \C{insert}s. This transformation requires vector
+% clocks~\cite{vectorclock} to keep track of casual ordering among
+% \C{insert}s and \C{remove}s.
+Indeed, most CRDTs have such carefully-crafted implementations that
+rely on non-trivial technical devices to keep track of causal
+dependencies and avoid or resolve conflicts. For instance, various
+CRDT variants of the set abstract data type rely on \emph{vector
+clocks} to identify conflicting \C{add}s and
+\C{remove}s~\cite{zawirski-thesis, zhang}, Replicated Growable Array
+(RGA) -- a CRDT for collaborative editing~\cite{rga} and JSON CRDT --
+a CRDT variant of the JSON storage format, both use \emph{Lamport
+timestamps}~\cite{rga, json-crdt}; and TreeDoc, another CRDT for
+collaborative editing, makes use of \emph{dense linear
+orders}~\cite{treedoc}. Such advanced implementations makes it hard to
+reason about basic correctness properties of CRDTs, such as
+convergence. Indeed, formal verification of \emph{strong eventual
+consistency} (i.e., eventual convergence) for few of the
+aforementioned CRDTs required considerable effort on behalf of
+multiple authors who are experts in the field~\cite{kleppmann2017}.
+The extraordinary effort and the expertise required to build CRDTs is
+a deterrent towards using type-safe abstractions with strong
+guarantees in distributed applications. To overcome this deterrent,
+there is a need for an alternative approach that lets developers reuse
+their sequential abstractions in a distributed setting with little to
+no additional overhead of reasoning about their correctness properties
+such as convergence.
 
@@
 % types in software development.
-In this paper we propose an alternative approach to replicated data types
-that addresses the problems discussed above. Our approach differs from the
-CRDT approach in two fundamental ways. Firstly, we adopt a state-centric
-model of replication with \emph{mergeability} of states as the defining
-characteristic in contrast to CRDTs' operation-centric model characterized
-by commutativity. Unlike commutativity, mergeability does not require a
-full-scale re-engineering of the data type definition. Instead,
-a data type only has to define a merge semantics in form a \C{merge}
-function to reconcile concurrent versions of the type in presence of their
-common ancestor version. Notably, \C{merge} function is left completely
-unconstrained even as convergence is guaranteed on the corresponding
-\emph{mergeable replicated data type} (MRDT)~\cite{mrdt}. Such strong
-guarantees are possible due to the simplicity of state-based reasoning
-relative to the operation- or trace-based one. Note that state-centric
-model of replication has also been adapted to CRDTs, but such state-based
-CRDTs require the replicated state to be organized as a lattice to
-guarantee convergence -- a restriction we eliminate in our setting.
+In this paper we propose such an alternative approach to convergent
+replicated data types.  
+% The key distinguishing characteristic of
+Our approach differs from the CRDT approach in
+two fundamental ways. Firstly, we adopt a state-centric model of
+replication with \emph{mergeability} of states as the defining
+characteristic in contrast to CRDTs' operation-centric model
+characterized by commutativity. Unlike commutativity, mergeability
+does not require a full-scale re-engineering of the data type
+definition. Instead, a data type only has to define a merge semantics
+in form a \C{merge} function to reconcile concurrent versions of the
+type in presence of their common ancestor version. Notably, \C{merge}
+function is left completely unconstrained even as convergence is
+guaranteed on the corresponding \emph{mergeable replicated data type}
+(MRDT)~\cite{mrdt}. Such strong guarantees are possible due to the
+simplicity of state-based reasoning relative to the operation- or
+trace-based one. Note that state-centric model of replication has also
+been adapted to CRDTs, but such state-based CRDTs require the
+replicated state to be organized as a lattice to guarantee convergence
+-- a restriction we eliminate in our setting.
 

### 2021-11-14 motivation.tex
Figures in motivation; edits to intro

@@
 convergent RDTs with help of a set data structure.
-% demonstrates the need for run-time intervention to ensure convergence.
 
-\subsection{Add-Wins Set RDT}
+\begin{figure}[ht]
+\centering
+\begin{ocaml}
+module Set: sig
+  type t (* Abstract type of the set*)    
+  type elt (* Abstract type of set elements *)
+  val add: elt -> unit (*Add an element *)
+  val remove: elt -> unit (* Remove an element *)
+end
+\end{ocaml}
+\caption{An OCaml interface to \C{Set} abstract data type}
+\label{fig:ocaml-set}
+\end{figure}
 
-Fig.~\ref{fig:ocaml-set} shows a simplified interface of a set data
-type in OCaml. The interface hides a reference to a set (\C{Set.t}),
-which can be updated in-place via \C{insert} and \C{remove}
-operations. As such \C{Set} is an ordinary data type primarily
-designed for sequential execution. Under a concurrent execution with
-an asynchronously replicated state, \C{Set} would exhibit anamolous
-behaviours such as those in Fig.~\ref{fig:crdt-execs}.
+\begin{figure}[ht]
+  \centering
+  \begin{subfigure}[t]{0.55\columnwidth}
+    \includegraphics[scale=0.35]{Figures/crdt-execs-1}
+    \caption{}
+    \label{fig:crdt-execs-1}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.44\columnwidth}
+    \includegraphics[scale=0.35]{Figures/crdt-execs-2}
+    \caption{}
+    \label{fig:crdt-execs-2}
+  \end{subfigure}
+\caption{Anamolous executions resulting from the Asynchronous
+  replication of \C{Set}. Dashed lines denote effect propagation.}
+\label{fig:crdt-execs}
+\end{figure}
+
+\noindent\paragraph{From Set ADT to Set RDT} Fig.~\ref{fig:ocaml-set}
+shows a simplified interface of \C{Set} (abstract) data type in OCaml.
+The interface hides a reference to a set (\C{Set.t}), which can be
+updated in-place via \C{add} and \C{remove} operations. \C{Set} is an
+ordinary data type meant for for sequential execution. Under a
+concurrent execution with an asynchronously replicated state, \C{Set}
+would exhibit anamolous behaviours such as those in
+Fig.~\ref{fig:crdt-execs}.
 
@@
 singleton set containing the element $e$. A client connects to the
-replica $R_3$ and executes a \C{remove $e$} operation, which is then
+replica $R_3$ and executes a \C{remove($e$)} operation, which is then
 asynchronously propagated to other replicas. Some time after applying
 $R_3$'s \C{remove} at $R_2$, another client connects to $R_2$ and
-re-adds $e$ by issuing an \C{insert $e$} operation. Consequently, the
+re-adds $e$ by issuing an \C{add($e$)} operation. Consequently, the
 state at $R_2$ is again the singleton set $\{e\}$. Replica $R_3$
-however receives $R_2$'s \C{insert} ahead of $R_1$'s \C{remove},
-applies them in the same order, and ends up with an empty set. The
-execution thus results in divergent replica states.
+however receives $R_2$'s \C{add} ahead of $R_1$'s \C{remove}, applies
+them in the same order, and ends up with an empty set. The execution
+thus results in divergent replica states.
 
-Note that if set \C{insert} and \C{remove} commute, then executing
+Note that if \C{Set.add} and \C{Set.remove} commute, then executing
 them in different order at $R_2$ and $R_3$ would not have led to
 divergence. As such, \C{Set} is not a CRDT due to the admittance of
-non-commutative operations. Nonetheless \C{Set} can be transformed
-into a CRDT by re-engineering its interface and operations. For
-instance, the anomalous execution in Fig.~\ref{fig:crdt-execs-1} can
-be pre-empted by ensuring that updates are only ever applied in the
-causal order.  This can be done by extending \C{Set} with vector
-clocks to keep track of the causal history of each operation. A set
-\C{insert} (resp. \C{remove}) would now generate an \C{Insert} (resp.
-\C{Remove}) \emph{effect} tagged with the vector clock of the origin
-replica. Here, the vector clock simply records the sequence number of
-last operation from each replica whose effect has been received and
-applied at the current replica. When an effect is received at a
-replica, it is buffered until the time all its causally-preceding
-effects (as captured by the tagged vector clock) have already been
-received and applied. This strategy would preempt the execution in
-Fig.~\ref{fig:ctrdt-execs-1} by buffering $R_2$'s \C{insert} at $R_1$
+non-commutative operations. Nonetheless, there exist approaches to
+transform \C{Set} into a CRDT by re-engineering its interface and
+operations~\cite{crdts, zawirski-thesis, zhang}. For instance, the
+anomalous execution in Fig.~\ref{fig:crdt-execs-1} can be pre-empted
+by ensuring that updates are only ever applied in the causal order.
+This can be done by extending \C{Set} with vector clocks to keep track
+of the causal history of each operation. A set \C{add} (resp.
+\C{remove}) would now generate an \C{Add} (resp.  \C{Remove})
+\emph{effect} tagged with the vector clock of the origin replica.
+Here, the vector clock simply records the sequence number of last
+operation from each replica whose effect has been received and applied
+at the current replica. When an effect is received at a replica, it is
+buffered until the time all its causally-preceding effects (as
+captured by the tagged vector clock) have already been received and
+applied. This strategy would preempt the execution in
+Fig.~\ref{fig:crdt-execs-1} by buffering $R_2$'s \C{add} at $R_1$
 until the causally-preceding \C{remove} of $R_1$ is received and
@@
 
-Unfortunately, \C{Set} data type of Fig.~\ref{fig:cc-set} still admits
-divergent executions due to concurrent updates.
-Fig.~\ref{fig:crdt-execs-2} describes one such execution. Here,
-replicas $R_1$ and $R_2$ both start with a singleton set $\{e\}$.
-Two distinct clients connect to $R_1$ and $R_2$ respectively and issue
-two concurrent \C{remove $e$} operations. Later, another
-client connects to $R_2$ and issues an \C{insert $e$} operation. The
-effects of these operations are asynchronously applied at remote
-replicas as shown in the figure, resulting in the divergent states at
-$R_1$ and $R_2$.
+\begin{figure}[ht]
+\centering
+\begin{ocaml}
+module Set: sig
+  type t      type elt
+  type r_id (* Replica Id *)
+  (* Vector Clock *)
+  type v_clock = (r_id,int) HashTable.t 
+  (* Set RDT Effects *)
+  type eff = Add of elt * v_clock 
+           | Remove of elt * v_clock
+  type buf = eff list (* pending effects *)
+  (* This replica's vector clock *)
+  val local_clock: v_clock 
+  val add: elt -> eff
+  val remove: elt -> eff
+  (* Apply an effect to the local state *)
+  val apply_eff: eff -> unit
+end
+\end{ocaml}
+\caption{An OCaml interface to \C{Set} replicated data type}
+\label{fig:cc-set}
+\end{figure}
+
+\noindent\paragraph{\emph{Add-Wins} Set CRDT} Unfortunately, \C{Set}
+data type of Fig.~\ref{fig:cc-set} still admits divergent executions
+due to concurrent updates.  Fig.~\ref{fig:crdt-execs-2} describes one
+such execution. Here, replicas $R_1$ and $R_2$ both start with a
+singleton set $\{e\}$.  Two distinct clients connect to $R_1$ and
+$R_2$ respectively and issue two concurrent \C{remove($e$)}
+operations. Later, another client connects to $R_2$ and issues an
+\C{add($e$)} operation. The effects of these operations are
+asynchronously applied at remote replicas as shown in the figure,
+resulting in the divergent states at $R_1$ and $R_2$.
 
 Note that, unlike the execution in Fig.~\ref{fig:crdt-execs-1}, the
-conflicting operations in Fig.~\ref{crdt-execs-2}, namely $R_1$'s
-\C{remove} and $R_2$'s \C{insert}, are \emph{not} causally related,
+conflicting operations in Fig.~\ref{fig:crdt-execs-2}, namely $R_1$'s
+\C{remove} and $R_2$'s \C{add}, are \emph{not} causally related,
 hence their relative order is not determined by the sequential
 specification of the data type. Forcing a causal relationship between
-them requires synchronization between \C{insert}s and \C{remove}s,
+them requires synchronization between \C{add}s and \C{remove}s,
 which is expensive in an asynchronous distributed setting. It
@@
 which are otherwise unordered. For example, in
-Fig.~\ref{fig:crdt-execs-2}, we might let $R_2$'s \C{insert} override
-$R_1$'s \C{remove} considering that \C{insert} is re-inserting an
+Fig.~\ref{fig:crdt-execs-2}, we might let $R_2$'s \C{add} override
+$R_1$'s \C{remove} considering that \C{add} is re-adding an
 element after a previous remove. Ordering concurrent removes ahead
-of inserts uniformly on all replicas leads to an implementation of
-\C{Set} RDT where concurrent (re-)insertions consistenty \emph{win}
-over removals. Such an \emph{add-wins} set is useful, for instance, to
+of adds uiniformly on all replicas leads to an implementation of
+\C{Set} RDT where concurrent (re-)adds consistenty \emph{win}
+over removes. Such an \emph{add-wins} set is useful, for instance, to
 implement a shared shopping cart where two users can concurrently
@@
 approach involves tracking element-wise causal dependencies between
-the \C{remove}s and \C{insert}s by maintaining a vector clock
-\emph{for each element} $e$ in the set~\cite{zawirski-thesis}. The
-vector clock of $e$ records the sequence number of the last \C{insert
-$e$} operation from each replica whose effect has been received and
-applied at the current replica. The \C{Insert} and \C{Remove} effects
-on $e$ will now be tagged with $e$'s vector clock. When an \C{Remove}
-effect on $e$ is received at a replica, it is applied only if the
-tagged vector clock is no less than $e$'s local vector clock, i.e.,
-only if the arriving \C{Remove} has seen (i.e., causally succeeds) at
-least those \C{insert $e$} operations the current replica is aware of.
-Otherwise the effect is simply a no-op. This strategy would result in
-convergent states despite the execution in Fig.~\ref{fig:crdt-execs-2}
-as $R_2$'s \C{remove} effectively becomes a no-op at $R_1$ due to
-there being at least one \C{insert} operation on $e$ that it hasn't
-seen. The \C{Set} interface extended with element-wise vector clocks
-is shown in Fig.~\ref{fig:rw-set}. The resultant set RDT is assumed to
-be correct albeit a formal proof of convergence could not be found in
-the literature.
+the \C{remove}s and \C{add}s by maintaining a vector clock \emph{for
+each element} $e$ in the set~\cite{zawirski-thesis}. The vector clock
+of $e$ records the sequence number of the last \C{add($e$)} operation
+from each replica whose effect has been received and applied at the
+current replica. The \C{Add} and \C{Remove} effects on $e$ will now be
+tagged with $e$'s vector clock. When an \C{Remove} effect on $e$ is
+received at a replica, it is applied only if the tagged vector clock
+is no less than $e$'s local vector clock, i.e., only if the arriving
+\C{Remove} has seen (i.e., causally succeeds) at least those
+\C{add($e$)} operations the current replica is aware of.  Otherwise
+the effect is simply a no-op. This strategy would result in convergent
+states despite the execution in Fig.~\ref{fig:crdt-execs-2} as $R_2$'s
+\C{remove} effectively becomes a no-op at $R_1$ due to there being at
+least one \C{add} operation on $e$ that it hasn't seen. The
+\emph{add-wins} \C{Set} interface is similar to Fig.~\ref{fig:cc-set},
+except that it requires an additional data structure to track
+element-wise vector clocks: 
+\begin{ocaml}
+  val e_clock: (elt, v_clock) HashTable.t
+\end{ocaml}
+The resultant set RDT is assumed to be correct albeit a formal proof
+of convergence could not be found in the literature.
 
-The above exercise demonstrates the considerable ingenuity and effort
-involved in deriving a convergent RDT out of such simple data type as
-\C{Set}. Some of the effort can be mitigated by strengthening the
-underlying system model insofar as it doesn't affect the latency and
-availability of the application. For instance, a system that always
-delivers messages in the causal order would automatically preempt the
-execution in Fig.~\ref{fig:crdt-execs-1} without the need for
-additional intervention on behalf of the developer. This is a
-particularly attractive proposition considering that causal
-consistency can be ``bolted on'' an existing implementation of an
-eventually consistent system without weakening its
-guarantees~\cite{bolton}. Unfortunately, such strengthening of the
+\noindent\paragraph{The problem with the CRDT approach} The above
+exercise demonstrates the considerable ingenuity and effort involved
+in deriving a convergent RDT out of such simple data type as \C{Set}.
+Such sophistication makes it quite hard for non-expert developers to
+build, or even customize existing replicated data types to suit the
+needs of their application. Some of the effort can be mitigated by
+strengthening the underlying system model insofar as it doesn't affect
+the latency and availability of the application. For instance, a
+system that always delivers messages in the causal order would
+automatically preempt the execution in Fig.~\ref{fig:crdt-execs-1}
+without the need for additional intervention on behalf of the
+developer. This is a particularly attractive proposition considering
+that causal consistency can be ``bolted on'' an existing
+implementation of an eventually consistent system without weakening
+its guarantees~\cite{bolton}. Unfortunately, such strengthening of the
 system model would deliver no benefits to the developer if they still
@@
 observed earlier, the execution in Fig.~\ref{fig:crdt-execs-2} seems
-inevitable unless every pair of \C{insert} and \C{remove} operations
-are synchronized, which, regrettably, is not a practical option. The
+inevitable unless every pair of \C{add} and \C{remove} operations are
+synchronized, which, regrettably, is not a practical option. The
 developer therefore seems to be stuck.
 
+\noindent\paragraph{Key observations guiding the solution}
 Fortunately, there is a way out of this problem and the solution
@@
 
-To demonstrate the aforementioned intuitions we reconsider the
-executions from Fig.~\ref{fig:crdt-execs}, this time in the
-state-centric replication model. The cornerstone of the state-centric
-model is a three-way \C{merge} function that merges concurrent
-versions of the state in presence of their (lowest) common ancestor
-version. In our running example, the state is a value of type
-\C{Set.t}, hence \C{Set.merge} function would have the type signature:
+\begin{figure}[ht]
+  \centering
+  \begin{subfigure}[t]{0.55\columnwidth}
+    \includegraphics[scale=0.35]{Figures/mrdt-execs-1}
+    \caption{}
+    \label{fig:mrdt-execs-1}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.44\columnwidth}
+    \includegraphics[scale=0.35]{Figures/mrdt-execs-2}
+    \caption{}
+    \label{fig:mrdt-execs-2}
+  \end{subfigure}
+\caption{Equivalent executions of Fig.~\ref{fig:crdt-execs} in
+state-centric replication model. Dashed lines now denote state merges.}
+\label{fig:mrdt-execs}
+\end{figure}
+
+\noindent\paragraph{A Mergeable Replicated Set} To demonstrate the
+aforementioned intuitions we reconsider the executions from
+Fig.~\ref{fig:crdt-execs}, this time in the state-centric replication
+model. The cornerstone of the state-centric model is a three-way
+\C{merge} function that merges concurrent versions of the state in
+presence of their (lowest) common ancestor version. In our running
+example, the state is a value of type \C{Set.t}, hence \C{Set.merge}
+function would have the type signature:
 \begin{center}
@@
 $v_4$ and $v_3$ at $R_1$ and $R_2$ (resp.) have witnessed the same set
-of operations, hence are in agreement.
+of operations, hence are in agreement.\GK{fix this.}
 
+\noindent\paragraph{Well-formed executions and the \quark runtime}
 Convergence however is not an inherent virtue of the state-centric
-replication model. Fig.~\ref{fig:mrdt-exec-2} shows the state-centric
-analogue of the execution in Fig.~\ref{fig:crdt-exec-2} which
+replication model. Fig.~\ref{fig:mrdt-execs-2} shows the state-centric
+analogue of the execution in Fig.~\ref{fig:crdt-execs-2} which
 diverges. Here $R_1$ and $R_2$ start with version $v_0 \,=\, \{e\}$.
-$R_2$ performs a \C{remove} and and an \C{insert} making versions
+$R_2$ performs a \C{remove} and and an \C{add} making versions
 $v_1$ and $v_2$ respectively. Simultaneously, $R_1$ performs a
@@
 \C{remove}), then $R_2$ into $R_1$ (bringing $R_2$'s \C{remove}),
-followed again by $R_2$ into $R_1$ (bringing $R_2$'s \C{insert}). As a
+followed again by $R_2$ into $R_1$ (bringing $R_2$'s \C{add}). As a
 result of such linearization, the final versions on $R_1$ and $R_2$
@@
 help of an abstract machine that generates only the well-formed
-executions.
+executions. \GK{fix this}

### 2021-11-16 introduction.tex
Sec 3.

@@
 \C{remove}s~\cite{zawirski-thesis, zhang}, Replicated Growable Array
-(RGA) -- a CRDT for collaborative editing~\cite{rga} and JSON CRDT --
+(RGA) -- a CRDT for collaborative editing~\cite{rga}, and JSON CRDT --
 a CRDT variant of the JSON storage format, both use \emph{Lamport

### 2021-11-16 motivation.tex
Sec 3.

@@
            | Remove of elt * v_clock
-  type buf = eff list (* pending effects *)
   (* This replica's vector clock *)
@@
   val remove: elt -> eff
+  val buf: eff list (* pending effects *)
   (* Apply an effect to the local state *)

### 2021-11-16 vc-semantics.tex
Sec 3.

\section{Semantics of State-Centric Replication}
\label{sec:vc-semantics}

\input{git-semantics}

In this section we present the formal semantics of our state-centric
replication scheme with linearized merges. As the informal development
from previous section suggests, our replication scheme is strongly
inspired by version control systems (VCS) such as Git~\cite{git}.  We
embrace this analogy in our formal development to manifest distributed
executions over replicated state with help of an abstract ``Git''
machine building a well-formed version history graph.  We show that
the version history graph thus generated has several desirable
properties including unique LCAs for every pair of versions, and
convergence of versions that include the same set of \emph{commits}.
We are however not concerned about the practical aspects of the system
yet; the subsequent sections gradually refine the abstract semantics
described here into a practical distributed system we call \quark. For
convenience, we refer to the abstract ``Git'' machine we describe here
also as \quark.

Fig.~\ref{fig:git-semantics} shows the operational semantics of the
\quark abstract machine. The machine admits the usual version control
operations, namely \rulelabel{Commit}, \rulelabel{Fork},
\rulelabel{Merge}, and \rulelabel{FastFwd}. Operation
\rulelabel{FastFwd} is a special case of merge which simply \emph{fast
forwards} a branch to a later version. A \emph{branch} is a linear
sequence of versions, which intuitively denotes the progressive
evolution of the state on a replica. The latest version of the branch,
called its \emph{head}, denotes the current state of the corresponding
replica. A new head version is created either by an
externally-initiated \emph{commit} (\rulelabel{Commit}) or by merging
the current head with a concurrent or causally-succeeding version from
a different branch (\rulelabel{Merge} and \rulelabel{FastFwd}
respectively). A new branch can be created by \emph{forking off} a
version on an existing branch (\rulelabel{Fork}). Although a version
control system admits more operations (e.g., ``rebase''), we observe
that these four basic actions are sufficient to capture the behavior
of an asynchronously replicated multi-versioned state machine.

The state of \quark abstract machine is a tuple $\Delta = (G,N,C,H,L)$, where:
\begin{itemize}
  \item $G = (V,E)$ is the version history DAG generated by the
    execution of the abstract machine. Vertices ($V$) of the graph are
    the set of all versions that ever existed during the execution.
    Edges ($E$) record the relationships between various versions. In
    particular, there are three kinds of edges corresponding to the
    three basic operations: commit ($\cedge$), merge ($\medge$), fork
    ($\fedge$). Fast forward, being a special case of merge, is also
    denoted by a merge edge ($\medge$). Existence of an edge $v_0
    \rightarrow  v_1$ denotes that versions $v_0$ and $v_1$ are related by
    some operation. For e.g., $v_0 \fedge v_1$ denotes that $v_1$ is a
    new version forked-off from the version $v_0$ using the
    \rulelabel{Fork} rule. As usual, \emph{path} relation is the
    reflexive transitive closure of the edge relation, i.e,
    $\reaches$.
%   \rulelabel{Fork}
%   rule (explained later) dictates that $v_0$ and $v_1$ are different
%   versions belonging to different branches, albeit denoting the

  \item $N : \texttt{Version} \rightarrow \texttt{Value}$ is a
    (partial) function (i.e., a map) that mas versions to their
    values.  We distinguish versions from values as different versions
    on different branches may store the same value (e.g., the same
    string ``hello''), yet need to be uniquely identified. The domain
    of values is left uninterpreted except for the requirement that
    it be \emph{mergeable}, i.e., define a three-way \C{merge}
    function.

  \item $C : \texttt{Version} \rightarrow \Pow{\texttt{CommitId}}$
    maps a version $v$ to the set of commit ids included in that
    version. Each commit event during the execution is uniquely
    identified by a commit id (analogous to the Git's commit hash).
    The set of commit ids $C(v)$ therefore denotes the commit events
    that contributed to the version $v$. Intuitively, this represents
    the set of user-initiated operations that affected the value of
    this version.

  \item $H : \texttt{Branch} \rightarrow \texttt{Version}$ is an
    injective (partial) function that maps each branch to its head
    version.

  \item $L: \texttt{Branch}\times\texttt{Branch} \rightarrow
    \texttt{Version}$ maps a pair of branches to the lowest common
    ancestor (LCA) version of their heads. We later prove the unique
    LCA property, so $L$ is indeed a (partial) function. Note that LCA
    of a pair of branches $b_1$ and $b_2$ need not necessarily lie on
    $b_1$ or $b_2$; it could also be a version on a different branch
    $b$. This happens, for e.g., if $b_1$ and $b_2$ alternatively
    merged the same version from $b$.
\end{itemize}

\paragraph{Notation} We let $v_{\odot}$ denote the initial version
(``root''), and $b_{\odot}$ denote the initial branch (``master'').
The execution of the abstract machine progresses by adding to the sets
$V$ and $E$, and updating the maps $N$, $C$, $H$, and $L$. We adopt
the usual update notation, e.g., $H[b \mapsto v]$ is a map $H'$ such
that $H'(b) = v$, and forall $b' \neq b$, $H'(b') = H(b')$.  Multiple
updates to a map are parsed left-associatively. Map $L$ is assumed to
be commutative, so $L[(b,b') \mapsto v]$ is equal to $L[(b,b') \mapsto
v][(b',b) \mapsto v]$; the former is used as a succinct replacement of
the latter. To update multiple bindings in $L$, we use the set
comprehension notation: given a branch $b$, $L[\{(b,b') \mapsto v
\,|\, \phi(b,b')\}]$ updates all bindings $(b,b')$ in $L$ to $v$,
where $b'$ is any branch such that $\phi(b,b')$ is true. Domain and
co-domain of a map $M$ is denoted $dom(M)$ and $codom(M)$
respectively.

\begin{definition}[Initial State and Version History Graph]
  The graph $G_{\odot} = (\{v_{\odot}\},\emptyset)$ is the initial
  version graph. The state $\Delta_{\odot}$ = $(G_{\odot},\,
  [v_{\odot} \mapsto b_{\odot}],\, [b_{\odot} \mapsto v_{\odot}],\,
  \emptyset)$ is the initial state.
\end{definition}

\noindent All executions of the abstract machine are assumed to start from the
initial state.

\begin{definition}[Ancestor]
  In version history DAG $G = (V,E)$, version $v_0 \in V$ is said
  to be a (causal) ancestor of $v_1 \in V$ iff $v_0 \reaches v_1$.
  Versions $v_0, v_1 \in V$ are causally related iff either $v_0
  \reaches v_1$ or $v_1 \reaches v_0$.
\end{definition}

% \begin{definition}[Common Ancestor] Common ancestor is defined for
%   versions and branches as following:
%   \begin{itemize} 
%     \item In $G = (V,E)$, a version $v\in V$ is called a common ancestor of
%       versions $v_1 \in V$ and $v_2 \in V$ iff $v \reaches v_1$ and $v \reaches
%       v_2$ in $G$.
%     \item In state $\Delta \,=\, ((V,E),N,C,H,L)$, $v\in V$ is called
%       a common ancestor of branches $b_1 \in dom(H)$ and $b_2 \in
%       dom(H)$ iff $v$ is a common ancestor of $H(b_1)$ and $H(b_2)$.
%   \end{itemize}
% \end{definition}

\begin{definition}[Lowest Common Ancestor (LCA)]
  In version history DAG $G = (V,E)$, a version $v\in$ is a lowest common
  ancestor of versions $v_1 \in V$ and $v_2 \in V$ iff:
  \begin{itemize}
    \item $v$ is a common ancestor of $v_1$ and $v_2$, i.e., $v
      \reaches v_1$ and $v \reaches v_2$, and
    \item There does not exist a $v'\in V$ such that $v'$ is a common
      ancestor of $v_1$ and $v_2$, and $v \reaches v'$.
  \end{itemize}
  LCA of a pair of branches is defined as the LCA of their heads.
\end{definition}

\paragraph{Rules} The rule \rulelabel{Commit}
(Fig.~\ref{fig:git-semantics}) describes committing a new version $v$
onto the branch $b$ updating its head. The value $n$ for the new
version is assumed to have been provided by whoever has invoked the
commit, e.g., the user. Intuitively, user invokes an RDT operation
(e.g., \C{add($e$)}) on the value of the current version to create the
new value $n$, and thus the new version $v$. A unique commit id $c$ is
assigned to this commit event and added to $C(v)$. The set $C(v)$
also contains all the commit ids from the previous version $H(b)$
since the new value $n$ is assumed to have been derived from the
previous value $N(H(b))$. The edge $H(b) \cedge v$ records this
dependency and also documents the progression of branch $b$. The LCA
map $L$ does not change as the only new edge is between the versions
of the same branch $b$.

\rulelabel{Fork} describes the semantics of forking a new branch $b'$
from the head of an existing branch $b$. The head of $b'$ is a new
version $v$ that shares the same commit set ($C(v)$) and value
($N(v)$) as its predecessor ($H(b)$). The lowest common ancestor (LCA)
of $b'$ and its parent $b$ is clearly the head of the parent $H(b)$ as
there does not exist a version $v$ lower than $H(b)$ that is an
ancestor of both $H(b)$ and $H(b')$. For every other branch $b''$,
$L(b',b'')$ is same as $L(b,b'')$. Fork operation could model, for
e.g., creating a new replica by forking off the current state of an
existing replica.

\rulelabel{Merge} describes the semantics of merging the head of a
branch $b'$ into $b$ resulting in an new version $v$ on $b$.
Intuitively, \rulelabel{Merge} models the information exchange between
replicas. The two pre-conditions specified using the strict superset
relation ($\supset$) require each of the merging versions, $H(b)$ and
$H(b')$, to include at least one commit not present in their common
ancestor $L(b,b')$. These conditions ensure that the merge is not
trivial (trivial merge is handled by \rulelabel{FastFwd}). The next
pre-condition is key to ensuring the uniqueness of LCAs and the
linearity of merges. It requires that, for every branch $b''$ in the
system, the LCA of $b''$ with merging branches $b$ and $b'$ be
causally related, i.e, $L(b,b'') \reaches L(b',b'') \disj L(b',b'')
\reaches L(b,b'')$. Fig.~\ref{fig:merge-precondition} helps visualize
this condition in the most general case when (\rom{1}). Branches $b$,
$b'$, and $b''$ are distinct, and (\rom{2}). Their LCAs $L(b,b'')$ and
$L(b',b'')$ lie on a distinct pair of branches not equal to $b$, $b'$,
and $b''$. In the figure, once you merge $b'$ into $b$, every version
$v$ that is an ancestor of $L(b',b'')$, i.e., $v \reaches L(b',b'')$,
will be  a common ancestor of $H(b)$ and $H(b'')$. Clearly,
$L(b',b'')$ is the lowest among such common ancestors. But the current
lowest common ancestor of $b$ and $b''$ is $L(b,b'')$. We therefore
end up with two lowest common ancestors -- $L(b',b'')$ and $L(b,b'')$,
\emph{unless} both are ancestrally related. Thus the pre-condition
$L(b,b'') \reaches L(b',b'') \disj L(b',b'') \reaches L(b,b'')$. If
$L(b',b'') \reaches L(b,b'')$, then $L(b,b'')$, the current LCA of $b$
and $b''$, is still the LCA after the merge. On the other hand, if
$L(b,b'') \reaches L(b',b'')$, then $L(b',b'')$ becomes the lowest
common ancestor of $b$ and $b''$ after the merge. Thus $L(b,b'')$
needs to be updated if and only if $L(b,b'') \reaches L(b',b'')$. The
conclusion of the \rulelabel{Merge} captures this update using the set
comprehension notation. It also updates the LCA of the merging
branches $L(b,b')$ to $H(b')$ since the head of $b'$ is merged into
$b$. Updates to the other components of the state (e.g., $C$) follow
the same rational as previours rules. Two edges are added to $E$ as
the new version $v$ is a descendent of the two merging versions.

Another notable aspect of the \rulelabel{Merge} rule is the invocation
of the \C{merge} function on the values of the  merging versions and
their common ancestor to derive the result of the merge. As explained
above, our development is parameterized on the domain of values for
which a \C{merge} function is defined. No futher constraints are
imposed on \C{merge}. We use an uncurried version of \C{merge} to
avoid clutter.

\rulelabel{FastFwd} generalizes merge to the case when the merging
version $H(b')$ is a descendent of the version $H(b)$. In this case
$H'(b)$, the new head of $b$, needs to have the same value and same
set of commits as $H(b')$. Other premises and conclusions are similar
to the \rulelabel{Merge} rule. Note that we don't need a separate rule
for fast forward merge if \C{merge} satisfies the invariant that
$\forall n, n'.~ \C{merge}(n,n,n') \,=\, \C{merge}(n,n',n) \,=\, n$.
Having a separate rule lets us elide this constraint.

\paragraph{Properties} We now formalize the notable properties of the
abstract machine and its executions\footnote{
  The manaul proofs of the theorems and their Ivy
  formalization~\cite{ivy} can be found in the supplementary material.
}.

\begin{lemma}[Uniqueness of LCA]
  \label{lem:lca-uniqueness}
  In every reachable state $\Delta = (G,N,C,H,L)$ of the abstract
  machine, every pair of branches $b_1, b_2 \in dom(H)$ has a unique
  LCA given by $L(b_1,b_2)$.
\end{lemma}

The intuition behind the proof is succinctly captured by
Fig.~\ref{fig:merge-precondition}, which is explained above.

\begin{lemma}[Commit sets grow monotonically]
  \label{lem:commit-monotonicity}
  In every reachable state $\Delta = ((V,E),N,C,H,L)$ of the abstract
  machine: Forall $v_1,v_2 \in V$, if $v_1 \reaches v_2$ then $C(v_1)
  \subseteq C(v_2)$.
\end{lemma}

Lemma~\ref{lem:commit-monotonicity} guarantees that merges never lose
a commit.

\begin{corollary}[Commit sets modulo LCA are disjoint]
  \label{lem:commit-disjointness}
  In every reachable state $\Delta = ((V,E),N,C,H,L)$ of the abstract
  machine: Forall distinct $b_1, b_2 \in dom(H)$, and $v_0, v_1, v_2
  \in V$ s.t.  $v_1 = H(b_1)$ and $v_2 = H(b_2)$ and $v_0 =
  L(b_1,b_2)$, the following is true: $(C(v_2) - C(v_0)) ~\cap~
  (C(v_1) - C(v_0)) = \emptyset$.
\end{corollary}

Corollary~\ref{lem:commit-disjointness} follows from
Lemmas~\ref{lem:lca-uniqueness} and~\ref{lem:commit-monotonicity}.

\begin{theorem}[{\bf Convergence}]
  \label{thm:convergence}
  In every reachable state $\Delta = ((V,E),N,C,H,L)$ of the abstract
  machine: Forall distinct $b_1, b_2 \in dom(H)$, and $v_1, v_2 \in V$
  such that $v_1 = H(b_1)$ and $v_2 = H(b_2)$, the following is true:
  $C(v_1) = C(v_2) \Rightarrow N(v_1) = N(v_2)$.
\end{theorem}

Theorem~\ref{thm:convergence} is the key result of this section. It
asserts that any two branches that witnessed the same set of commits
have the same value. Intuitively, this means that any two replicas
that witnessed the same set of user actions arrive at the same final
state \emph{regardless} of the order in which they are witnessed.

### 2021-11-17 concrete-semantics.tex
Quark concrete semantics added

\section{Concrete Semantics}
\label{sec:concrete-sem}

\input{quark-semantics}

### 2021-11-18 concrete-semantics.tex
Concrete semantics (Sec 4.)

@@
 \input{quark-semantics}
+
+The \quark abstract machine of the previous section deliberately
+ignores system-level concerns to focus on the semantics of
+version-controlled state replication. In this section we present the
+\quark \emph{distributed} machine that addresses the key system-level
+concerns and provides a blueprint for a practical version-controlled
+replicated state machine. We first reify the the one-to-one
+correspondence between replicas and branches we assumed informally in
+the previous section. This requires us to relax the assumption of the
+state $\Delta$ being shared synchronously across all replicas. Next we
+present an efficient method to track the version history with help
+of vector clocks. Finally, we outline an algorithm for garbage
+collecting older versions so that the version history need not grow
+unboundedly.
+
+Fig.~\ref{fig:quark-semantics} shows the operational semantics of the
+distributed machine. The key difference from the abstract semantics of
+Fig.~\ref{fig:git-semantics} is the presence of $\texttt{ReplicaId}$
+indexing the components of the system state. We thus admit the
+possibility of different replicas having different conceptions of the
+state. Another major difference is the use of version
+vectors~\cite{vectorclock} as identifiers and placeholders for
+versions. A version vector of a version $v$ records the sequence
+number of the last version from each branch that causally precedes $v$
+(as per Def.~\ref{def:ancestor}).  Concretely, version vector is a map
+from branches to natural numbers:
+\begin{center}
+  $\texttt{VersionVector}$ = $\texttt{Branch} \rightarrow \mathbb{N}$
+\end{center}
+
+The state of the distributed machine is the triple $\delta = (B,N,H)$.
+Components $B$ and $H$ are indexed by $\texttt{ReplicaId}$ to let us
+denote their replica-local copies. For instance, replica $i$'s copy of
+the \emph{head map} $H$ is given by $H\;i$, which we abbreviate to
+$H_i$ for notational convenience. Like $H$ in
+Fig.~\ref{fig:git-semantics}, $H_i$ maps each branch to the version
+vector of its head. Note that two replicas may be out of sync w.r.t
+the information in $H$ and $B$.  For instance, $H_i(b)$ may not be
+equal to $H_j(b)$ if replicas $i$ and $j$ are out of sync. On a
+replica $i$, \emph{branch map} $B_i$ gives the version vector
+corresponding to a particular sequence number on a branch. For
+instance, the version vector of the first version on branch $b$ is
+given by $B_i(b,1)$ on replica $i$. The value map $N$ maps version
+vectors to their values. We assume a single copy of $N$ to simplify
+the presentation. Generalizing Fig.~\ref{fig:quark-semantics} to allow
+for replica-local copies of $N$ is straightforward.
+
+Conspicuous by its absence in Fig.~\ref{fig:quark-semantics} is the
+LCA map $L$, which we previously used to track the LCA version for
+every pair of branches. The use of version vectors, coupled with the
+unique LCA guarantee, obviates the need for an LCA map.  We can
+instead identify the LCA of a pair of versions $v_1$ and $v_2$ by
+computing the \emph{greatest lower bound} (GLB) of their version
+vectors $t_1$ and $t_2$. Concretely:
+\begin{center}
+  $t_l \,=\, t_1 \sqcap t_2$
+\end{center}
+Where $t_l$ is the version vector of the LCA and $\sqcap$ is the GLB
+operator. Conversely, the version vector indentifying the result of a
+merge can be computed as the \emph{least upper bound} (LUB) of the two
+version vectors involved in the merge. Concretely:
+\begin{center}
+  $t_m \,=\, t_1 \sqcup t_2$
+\end{center}
+Where $t_1$ and $t_2$, and $t_m$ are the version vectors of the
+merging versions and the result of the merge, respectively. For
+technical reasons, however, we increment the component of $t_m$
+corresponding to the current branch $b$ to signify that this is a new
+version on $b$. So the actual version vector of the merge is $t_m' =
+t_m[b \mapsto t_m(b)+1]$. The GLB and LUB operations on version
+vectors are standard~\cite{vectorclock} -- GLB is computed by taking
+the component-wise minumum, and LUB by taking their maximum. The
+comparison of version vectors is also standard -- $t_1 < t_2$ iff
+$\forall b.~t_1(b) < t_2(b)$. Clearly, not all version vectors are
+comparable. We write $t_1 \lesseqgtr t_2$ if vectors $t_1$ and $t_2$
+are comparable.
+
+\paragraph{Notation and Conventions} We enforce a one-to-one mapping
+between replicas and branches by adopting the convention that branch
+$b_i$ always corresponds to replica $i$. Concretely this means that
+replica $i$ only ever creates new versions on branch $b_i$. Also,
+replica $i$ can only update its local copies of $H$ and $B$, i.e.,
+$H_i$ and $B_i$. For e.g., $H_i[b_i \mapsto t]$ updates the head of
+branch $b_i$ to $t$ on replica $i$.  Since $H_i$ is simply an
+abbreviation of $H\;i$ in the formalism, $H_i[b_i \mapsto t]$ actually
+expands to $H_i[i \mapsto b_i \mapsto t]$. We exploit this notation in
+Fig.~\ref{fig:quark-semantics}. 
+% When two replicas, $i$ and $j$, are involved in an update, e.g.,
+% \rulelabel{Fork}, we write $H_{\tuplee{i,j}}[b_j \mapsto t]$ to mean
+% $H[i \mapsto b_j \mapsto t][j \mapsto b_j \mapsto t]$.
+
+\paragraph{Rules} The transition relation $\qstepsto$ of the \quark
+distributed machine is defined in Fig.~\ref{fig:quark-semantics}.
+There is one rule corresponding to every rule in
+Fig.~\ref{fig:git-semantics} (abstract semantics). \rulelabel{Commit}
+describes replica $i$ committing a new version on the branch $b_i$
+with the given value $n$.  The vector $t'$ for the new version is
+obtained from that of the previous version $t$ by incrementing the
+component $b_i$. The sequence number of the new version on $b_i$ is
+$t'(b_i)$ and the branch map $B_i$ is updated to reflect that. Note
+that for all $j \neq i$, $H_j$ and $B_j$ remain unchanged indicating
+that other replicas are not (yet) aware of this commit.
+%
+\rulelabel{Fork} forks off a new branch $b_j$ from $b_i$. A new
+replica $j$ is assumed to take over $b_j$. The version vector $t'$ of
+the branch head now has a new component $b_j$ mapped to $1$ to denote
+this is the first version on $b_j$. Map $B_j$ is updated accordingly.
+Value $N(t')$ is same as its parent $N(t)$.
+
+\rulelabel{Merge} describes the semantics of replica $i$ merging a
+concurrent version from a (remote) replica $j$. The merging versions
+are heads of their respective branches $b_i$ and $b_j$.  Like its
+counterpart in Fig.~\ref{fig:git-semantics}, \rulelabel{Merge} insists
+that the LCAs of every other branch $b_k$ with $b_i$ and $b_j$ be
+causally related. This condition is however expressed in terms of
+version vectors with help of the GLB ($\sqcap$) and comparision
+($\lesseqgtr$) operators. For the LCA determination to be sound,
+the merging replica $i$ needs to have an accurate conception of the
+current version history. Futhermore, there cannot be a concurrent
+merge happening elsewhere that undermines the judgment of replica $i$
+(reg. the safety of $i \leftarrow j$ merge). These requirements are
+enforced by the premise $\forall (k\in dom(H)).~b_k \in dom(H_i) \conj
+H_i(b_k) = H_k(b_k)$, which insists that replica $i$'s knowledge of
+every other branch $k$ be current. This condition effectively
+linearizes merges by requiring a merge to either \emph{see} or
+\emph{be seen} by every other merge. In practice this is achieved
+through global coordination (Sec.~\ref{sec:implementation}). Note that
+\rulelabel{Merge} only pre-empts a concurrent \rulelabel{Merge}, not a
+concurrent \rulelabel{Commit}. A remote replica $k$ is allowed to keep
+committing new versions on to $b_k$ even as it remains unaware of the
+merge on replica $i$. Such leniency is imperative if the system were
+to retain the performance benefits of asynchronous replication.
+\rulelabel{FastFwd} is similar to \rulelabel{Merge}, so we don't
+discuss it separately.
+
+The rule \rulelabel{Sync} captures asynchronous communication between
+a pair of replicas ($i$ and $j$) to get one of them ($i$) up-to-date
+with the other ($j$).  Unlike the other rules, \rulelabel{Sync} does
+not extend the version history graph, and therefore has no counterpart
+in Fig.~\ref{fig:git-semantics}. It merely updates replica $i$'s
+knowledge of branch $b_k$ if replica $j$ happens to have later updates
+from $k$, i.e., $H_j(b_k)$ happens to be ahead of $H_i(b_k)$. The new
+versions on $b_k$ known to $j$ but not $i$ are then simply transferred
+to $i$.

### 2021-11-19 introduction.tex
Edits to abstract and introduction

@@
 In this paper we propose such an alternative approach to convergent
-replicated data types.  
-% The key distinguishing characteristic of
-Our approach differs from the CRDT approach in
-two fundamental ways. Firstly, we adopt a state-centric model of
-replication with \emph{mergeability} of states as the defining
-characteristic in contrast to CRDTs' operation-centric model
-characterized by commutativity. Unlike commutativity, mergeability
-does not require a full-scale re-engineering of the data type
-definition. Instead, a data type only has to define a merge semantics
-in form a \C{merge} function to reconcile concurrent versions of the
-type in presence of their common ancestor version. Notably, \C{merge}
-function is left completely unconstrained even as convergence is
-guaranteed on the corresponding \emph{mergeable replicated data type}
-(MRDT)~\cite{mrdt}. Such strong guarantees are possible due to the
-simplicity of state-based reasoning relative to the operation- or
-trace-based one. Note that state-centric model of replication has also
-been adapted to CRDTs, but such state-based CRDTs require the
-replicated state to be organized as a lattice to guarantee convergence
--- a restriction we eliminate in our setting.
+replicated data types. The key distinguishing characteristic of our
+approach is the utilization of a distributed runtime to orchestrate
+\emph{well-formed} distributed executions that are guaranteed to
+converge. We define well-formedness as a constraint over the
+\emph{states} that arise during the execution as opposed to the
+\emph{operations} that are involved in the execution. To allow for its
+enforcement, we adopt a state-centric model of replication based on
+version-controlled \emph{mergeable} states~\cite{mrdt} instead of an
+operation-centric model based on commutative operations. Unlike
+commutativity, mergeability does not require a data type definition to
+be refactored to suit distributed execution. Instead, the type
+definitions needs to be \emph{extended} with a \C{merge} function to
+reconcile concurrent versions of the type in presence of their common
+ancestor version. Notably, \C{merge} function is left completely
+unconstrained. Consequently, programmer can chose any merge semantics
+that make sense within the context of the data type's (sequential)
+semantics without risking the loss of convergence. Such strong
+guarantees are made possible by the simplicity of state-based
+reasoning combined with the run-time enforcement of well-formedness.
 
-The second distinguishing aspect of our approach is the utilization of a
-distributed runtime that orchestrates \emph{well-formed} distributed
-executions that are guaranteed to converge. Note that it is easy to
-construct a trivial runtime that guarantees convergence of all executions.
-Such a runtime would make an extensive use of inter-replica coordination to
-synchronize the execution of every operation and thus induce a
-sequentially-consistent behavior. This would however defeat the purpose of
-state replication as the latency incurred for synchronized execution of an
-operation is prohibitively high, and the availability of the system during
-the execution is correspondingly low. It is therefore important for a
+% combined with the simplicity
+% of state-based reasoning. Note that state-centric model of
+% replication has also been explored in the context of CRDTs. However,
+% state-based CRDTs require the replicated state to be organized as a
+% lattice -- a restriction we relax in our setting.
+
+Note that it is easy to construct a trivial runtime that guarantees
+convergence of all executions.  Such a runtime would make an extensive
+use of inter-replica coordination to synchronize the execution of
+every operation and thus induce a sequentially-consistent behavior.
+This would however defeat the purpose of state replication as the
+latency incurred for synchronized execution of an operation is
+prohibitively high, and the availability of the system during the
+execution is correspondingly low. It is therefore important for a
 distributed runtime of RDTs to not synchronize the execution of RDT
@@
 impacting latency and availability may seem impossible due to the CAP
-theorem~\cite{cap}, but it is not the case in fact. While CAP theorem does
-rule out linearizability, it does not prohibit enforcing weaker consistency
-constraints on distributed executions. In fact many CRDTs, including the
-aforementioned \emph{remove-wins} set, implicitly assume \emph{causal
-consistency}, which is weaker than linearizability but is stronger than the
-default \emph{eventual consistency}. While causal consistency alone cannot
-guarantee convergence, it sufficiently constrains the distributed execution
-for the CRDT implementations to guarantee convergence. In this respect our
-approach differs in the sense that (a). It identifies a novel set of
-constraints on distributed execution that guarantee the convergence of an
-MRDT \emph{regardless} of its merge semantics, and (b). It localizes those
-constraints spatially and temporally into the runtime such that
-per-operation latency and system-wide availability remain unaffected.
-Sec.~\ref{sec:semantics} explains our runtime in detail.
+theorem~\cite{cap}, but it is not the case in fact. While CAP theorem
+does rule out linearizability, it does not prohibit enforcing weaker
+consistency constraints on distributed executions. Our approach
+exploits this observation by (\rom{1}). Identifying a novel set of
+constraints on distributed executions that guarantee their
+convergence, and (\rom{2}). Localizing such constraints on
+(asynchronously executed) state merges such that per-operation latency
+and system-wide availability remain unaffected.
+Sec.~\ref{sec:concrete-sem} explains our runtime in detail.
+
+% In fact many CRDTs,
+% including the aforementioned \emph{remove-wins} set, implicitly assume
+% \emph{causal consistency}, which is weaker than linearizability but is
+% stronger than the default \emph{eventual consistency}. While causal
+% consistency alone cannot guarantee convergence, it sufficiently
+% constrains the distributed execution for the CRDT implementations to
+% guarantee convergence. In this respect our approach differs in the
+% sense that
 
@@
 \begin{itemize}
-  \item We present a model of state replication for \emph{mergeable} data
-  types, i.e., data types equipped with a \C{merge} function, and identify a
-  structural well-formedness condition over distributed executions that is
-  sufficient to guarantee their convergence. This is in contrast to
-  existing models of replicated data types, where each type is obligated to
-  prove its convergence.
+  \item We present a runtime-assisted approach to state replication
+    that sufficiently constrains distributed executions so as to
+    guarantee the convergence of replicated data types. This is in
+    contrast to the existing models of replication, where each type is
+    obligated to prove its convergence.
 
-  \item We formalize a distributed runtime called \quark that orchestrates
-  well-formed distributed executions. We prove the correctness of \quark,
-  and also demonstrate its \emph{progress} property, namely that no
-  distributed execution would ever get ``stuck''. Besides the manual proofs
-  for the theorems, we also provide the full formalization of the \quark
-  system and its properties in Alloy~\cite{alloy} and Ivy~\cite{ivy}. The
-  Ivy formalization of \quark has been automatically verified with help of
-  Z3~\cite{z3}. To the best of our knowledge, this is the first time an SMT
-  solver was used to verify the \emph{meta-theory} of a formal setup.
+  \item We formalize the aforementioned constraints in the context of
+    a an abstract machine (named \quark) that manifests distributed
+    executions as version history graphs (similar to Git~\cite{git}).
+    We prove the \emph{convergence} and \emph{progress} properties of
+    the \quark abstract machine that together guarantee its soundness.
+    The formalization has been mechanized in Ivy~\cite{ivy} and
+    \emph{automatically} verified with help of Z3~\cite{z3}. To the
+    best of our knowledge, this is the first time an SMT solver was
+    used to verify the \emph{meta-theory} of a formal system.
 
-  \item We realize an implementation of \quark as a light-weight shim layer
-  on top of Scylla -- an off-the-shelf distributed key-value
-  store~\cite{scylla}. We do a thorough evaluation with a
-    collaborative-editing benchmark to demonstrate that \quark has
-    \emph{no} effect on the latency and availability of the system, while
-    slightly increasing the time to convergence.
+  \item We formalize \quark distributed machine that implements the
+    semantics of \quark abstract machine in the context of an
+    asynchronous distributed system. The formalization addresses
+    several practical concerns of distribution and serves as a
+    blueprint for an efficient implementation.
+
+  \item We describe an implementation of \quark as a light-weight shim
+    layer on top of Scylla -- an off-the-shelf distributed key-value
+    store~\cite{scylla}. We evaluate the implementation on a
+    collaborative-editing benchmark to demonstrate novel tradeoffs
+    presented by our approach.
 \end{itemize}

### 2021-11-19 quark.tex
Introduction added
Motivation in progress
Motivation text complete
Sec 3.
Quark concrete semantics added
Concrete semantics (Sec 4.)
Edits to abstract and introduction

@@
 %% Title information
-\title[Convergence Sans Commutativity]{Convergence Sans Commutativity:
-Mergeability as a basis for Convergent Replicated Data Types}      %% [Short Title] is optional;
+\title[]{Runtime-Assisted Convergence in Replicated Data Types}
+%% [Short Title] is optional;
                                         %% when present, will be used in
@@
 \begin{abstract}
-   Programming geo-distributed replicated systems is hard given the
-   complexity of reasoning about constantly evolving states at different
-   replicas. Uncoordinated updates to replicas lead to states that are
-   unlikely to converge. In the absence of inter-replica coordination,
-   commutativity has come to be accepted as the \emph{de facto} design
-   principle to guarantee convergence. Unfortunately, designing replicated
-   state updates as commutative operations is hard considering that most
-   data types define operations that do not commute. Deriving a replicated
-   variant of a data type often entails a creative re-engineering of its
-   internal representation and algorithms so as to support commutative
-   operations while retaining (close to) the original semantics. In this
-   paper we propose an alternative approach to replicated data types that
-   avoids such complex re-engineering. Our approach is fundamentally
-   different in that it is based on the weaker notion of
-   \emph{mergeability}. Unlike commutativity, mergeability only requires a
-   data type to define \emph{a} merge semantics while leaving its existing
-   definition intact. Notably, our approach leaves type and merge semantics
-   completely unconstrained even as it guarantees eventual convergence and
-   full availability in an asynchronous distributed setting. In other
-   words, there is no need for the programmers to prove algebraic
-   properties of their implementations in order to obtain the convergence
-   and availability guarantees under a distributed execution. Such
-   weakening of obligations is made possible by a novel distributed runtime
-   that orchestrates distributed executions as per a structural
-   well-formedness criterion, which we show is sufficient to guarantee
-   convergence. We implement the runtime called \quark as a thin shim layer
-   on top of an off-the-shelf distributed database, and use it to run
-   mergeable replicated variants of several ordinary data types. Our
-   evaluation categorically demonstrate the generality of our approach and
-   the low overhead imposed by our runtime.
 
+  We propose a runtime-assisted approach to enforce convergence in
+  distributed executions of replicated data types. The key
+  distinguishing aspect of our approach is that it guarantees
+  convergence \emph{unconditionally} -- without requiring data
+  type operations to satisfy algebraic laws such as commutativity and
+  idempotence. Consequently, programmers are no longer obligated to
+  prove convergence on a per-type basis. Moreover, our approach lets
+  sequential data types be reused in a distributed setting by
+  extending their implementations rather than refactoring them.
+  The cornerstone of our approach is a distributed runtime that
+  orchestrates \emph{well-formed } executions that are guaranteed to
+  converge. Despite the utilization of a runtime, our approach comes
+  at no additional cost of per-operation latency and system-wide
+  availability. Instead, the tradeoff is the time required to achieve
+  convergence -- a metric we call \emph{staleness}. We quantify this
+  tradeoff via a thorough evaluation of a collaborative editing
+  application. and demonstrate a novel practical path towards
+  achieving high scalability through geo-replication.
 \end{abstract}
@@
 
-Text of paper \ldots
+\input{motivation}
+
+\input{vc-semantics}
+
+\input{concrete-semantics}
 

### 2021-11-19 abstract-semantics.tex
Implementation section

--- 2021-11-16 vc-semantics.tex
@@
 \begin{definition}[Ancestor]
+  \label{def:ancestor}
   In version history DAG $G = (V,E)$, version $v_0 \in V$ is said
@@
 $b$. Updates to the other components of the state (e.g., $C$) follow
-the same rational as previours rules. Two edges are added to $E$ as
+the same rationale as previours rules. Two edges are added to $E$ as
 the new version $v$ is a descendent of the two merging versions.

### 2021-11-19 concrete-semantics.tex
Edits to abstract and introduction
Implementation section

@@
 from $k$, i.e., $H_j(b_k)$ happens to be ahead of $H_i(b_k)$. The new
-versions on $b_k$ known to $j$ but not $i$ are then simply transferred
-to $i$.
+versions on $b_k$ known to $j$ but not $i$ are then simply replicated
+at $i$.
+
+\paragraph{Garbage Collection} One downside of the distributed
+semantics in Fig.~\ref{fig:quark-semantics} is that maps $B$ and $N$
+that together track the version history grow monotonically as the
+execution progresses and new versions are created. Fortunately, this
+is easy to address as the execution only needs version history
+inasmuch as it is required to obtain the LCA versions for
+\rulelabel{Merge}. Since LCA version vectors monotonically increase,
+older versions with vectors less than the least known LCA vector can
+simply be garbage collected. Moreover, a replica can make this
+decision locally without having to synchronize with its peers. In
+practice, applications may prefer to flush out older version history
+to a stable storage from where it can be re-created as necessary.

### 2021-11-19 implementation.tex
Implementation section

\section{Implementation}
\label{sec:implementation}

We implement \quark runtime as a lightweight shim layer on top of
Scylla -- an off-the-shelf distributed data store~\cite{scylla}. We
rely on Scylla for inter-replica communication, data replication,
persistence, and fault tolerance. \quark translates the high-level
mergeable data type implementations in OCaml to their low-level
representations in the backing store and organizes their well-formed
distributed executions. Fig.~\ref{fig:system} illustrates the overall
architecture.

The runtime \quark is a straightforward implementation the of
distributed machine \quark described in Sec.~\ref{sec:concrete-sem}.
We manifest each component of the state, namely the branch map $B$,
the value map $N$, and the head map $H$, as a column family (i.e., a
table) in scylla. The synchronization needed to linearize merges is
implemented with help of Scylla's support for conditional updates (CAS
operations) and expiring columns. The total order among merges is
enforced with help of \C{Quorum} reads and writes. Each user process
is assigned its own replica of the RDT.  Replica id is the
concatenation of its host name, process id, and Unix timestamp.
Version vectors are realized as associative lists and stored in Scylla
as blobs.

One key difference between the formalization and the implementation is
in the treatment of RDT values. Formalization assumes values to be
atomic with no sharing in between them. In practice, however, an RDT
value could be a linked data structure such as a binary tree, and two
such values could share a significant amount of internal structure.
Consequently, the size of the \emph{diff} between two consecutive
versions of a value could be asymptotically less then the size of the
data structure itself, in which case it unreasonable to transfer the
entire data structure over the network. To facilitate the efficient
computation of diffs between versions of data structures, we implement
a \emph{content-addressable store} as a key-value table in Scylla
where key is simply the SHA256 hash of the value. A linked data
structure is stored as a collection of nodes, where each node links to
the other by referring to its hash. The diff between two consecutive
versions of a data structure would simply manifest as new entries in
the content-addressable store reachable from the root of the new
version. The new entries, being new data, are automatically replicated
by Scylla, thus letting us reconstruct the new version at a remote
location. The root hash of the new version is obtained from the value
map $N$, which now maps version vectors to the \emph{hashes} of
values.

### 2021-11-19 introduction.tex
Implementation section

@@
 be refactored to suit distributed execution. Instead, the type
-definitions needs to be \emph{extended} with a \C{merge} function to
+definition only needs to be extended with a \C{merge} function to
 reconcile concurrent versions of the type in presence of their common
@@
   \item We formalize \quark distributed machine that implements the
-    semantics of \quark abstract machine in the context of an
+    semantics of the \quark abstract machine in the context of an
     asynchronous distributed system. The formalization addresses

### 2021-11-19 motivation.tex
Implementation section

@@
 
-\noindent\paragraph{\emph{Add-Wins} Set CRDT} Unfortunately, \C{Set}
+\noindent\paragraph{Add-Wins Set CRDT} Unfortunately, \C{Set}
 data type of Fig.~\ref{fig:cc-set} still admits divergent executions

### 2021-11-19 abstract-semantics.tex
Figures and Progress thm added.

@@
 \section{Semantics of State-Centric Replication}
-\label{sec:vc-semantics}
+\label{sec:abstract-sem}
 
@@
 
+\begin{figure}[ht]
+  \centering
+    \includegraphics[scale=0.4]{Figures/merge-precondition}
+\caption{Explanation for the premise $L(b,b'') \reaches L(b',b'') \disj L(b',b'')
+          \reaches L(b,b'')$ on the \rulelabel{Merge} rule.}
+\label{fig:merge-precondition}
+\end{figure}
+
 \rulelabel{Merge} describes the semantics of merging the head of a
@@
 $b$. Updates to the other components of the state (e.g., $C$) follow
-the same rationale as previours rules. Two edges are added to $E$ as
-the new version $v$ is a descendent of the two merging versions.
+the same rationale as previous rules. Two edges are added to $E$ as
+the new version $v$ is a descendant of the two merging versions.
 
@@
 above, our development is parameterized on the domain of values for
-which a \C{merge} function is defined. No futher constraints are
+which a \C{merge} function is defined. No further constraints are
 imposed on \C{merge}. We use an uncurried version of \C{merge} to
@@
 \rulelabel{FastFwd} generalizes merge to the case when the merging
-version $H(b')$ is a descendent of the version $H(b)$. In this case
+version $H(b')$ is a descendant of the version $H(b)$. In this case
 $H'(b)$, the new head of $b$, needs to have the same value and same
@@
 abstract machine and its executions\footnote{
-  The manaul proofs of the theorems and their Ivy
+  The manual proofs of the theorems and their Ivy
   formalization~\cite{ivy} can be found in the supplementary material.
@@
   \label{thm:convergence}
-  In every reachable state $\Delta = ((V,E),N,C,H,L)$ of the abstract
+  In every reachable state $\Delta$ = $((V,E),N,C,H,L)$ of the abstract
   machine: Forall distinct $b_1, b_2 \in dom(H)$, and $v_1, v_2 \in V$
@@
 state \emph{regardless} of the order in which they are witnessed.
+
+Convergence is vacuously true if the abstract machine never lets any
+merges to happen, i.e., if the premises of the \rulelabel{Merge} rule
+are too strong to be never true. We prove that this is not the case
+with help of the following theorem:
+
+\begin{theorem}[{\bf Progress}]
+  \label{thm:progress}
+  Every reachable state $\Delta = ((V,E),N,C,H,L)$ of the abstract
+  machine is either: 
+  \begin{itemize}
+    \item A ``quiescent'' state, where: $\forall b_1,b_2 \in
+      dom(H).~C(H(b_1)) = C(H(b_2))$, Or
+    \item An ``unstuck'' state, where there exist $b_1,b_2 \in dom(H)$
+      that satisfy the pre-conditions of \rulelabel{Merge} or
+      \rulelabel{FastFwd} rules, i.e., $b_1$ and $b_2$ are mergeable. 
+  \end{itemize}
+\end{theorem}
+
+Convergence and Progress together ensure the soundness of the \quark
+abstract machine.

### 2021-11-19 concrete-semantics.tex
Figures and Progress thm added.

@@
 Where $t_l$ is the version vector of the LCA and $\sqcap$ is the GLB
-operator. Conversely, the version vector indentifying the result of a
+operator. Conversely, the version vector identifying the result of a
 merge can be computed as the \emph{least upper bound} (LUB) of the two
@@
 vectors are standard~\cite{vectorclock} -- GLB is computed by taking
-the component-wise minumum, and LUB by taking their maximum. The
+the component-wise minimum, and LUB by taking their maximum. The
 comparison of version vectors is also standard -- $t_1 < t_2$ iff
@@
 causally related. This condition is however expressed in terms of
-version vectors with help of the GLB ($\sqcap$) and comparision
+version vectors with help of the GLB ($\sqcap$) and comparison
 ($\lesseqgtr$) operators. For the LCA determination to be sound,
 the merging replica $i$ needs to have an accurate conception of the
-current version history. Futhermore, there cannot be a concurrent
+current version history. Furthermore, there cannot be a concurrent
 merge happening elsewhere that undermines the judgment of replica $i$
@@
 through global coordination (Sec.~\ref{sec:implementation}). Note that
-\rulelabel{Merge} only pre-empts a concurrent \rulelabel{Merge}, not a
+\rulelabel{Merge} only preempts a concurrent \rulelabel{Merge}, not a
 concurrent \rulelabel{Commit}. A remote replica $k$ is allowed to keep
@@
 older versions with vectors less than the least known LCA vector can
-simply be garbage collected. Moreover, a replica can make this
+simply be garbage-collected. Moreover, a replica can make this
 decision locally without having to synchronize with its peers. In

### 2021-11-19 implementation.tex
Figures and Progress thm added.

@@
 the value map $N$, and the head map $H$, as a column family (i.e., a
-table) in scylla. The synchronization needed to linearize merges is
+table) in Scylla. The synchronization needed to linearize merges is
 implemented with help of Scylla's support for conditional updates (CAS

### 2021-11-19 motivation.tex
Figures and Progress thm added.

@@
     \caption{}
-    \label{fig:crdt-execs-1}
+    \label{fig:crdt-exec-1}
   \end{subfigure}
@@
     \caption{}
-    \label{fig:crdt-execs-2}
+    \label{fig:crdt-exec-2}
   \end{subfigure}
@@
 concurrent execution with an asynchronously replicated state, \C{Set}
-would exhibit anamolous behaviours such as those in
+would exhibit anamolous behaviors such as those in
 Fig.~\ref{fig:crdt-execs}.
 
-Fig.~\ref{fig:crdt-execs-1} shows an anamolous execution with three
+Fig.~\ref{fig:crdt-exec-1} shows an anamolous execution with three
 replicas -- $R_1$, $R_2$, and $R_3$, all of which start with a
@@
 operations~\cite{crdts, zawirski-thesis, zhang}. For instance, the
-anomalous execution in Fig.~\ref{fig:crdt-execs-1} can be pre-empted
+anomalous execution in Fig.~\ref{fig:crdt-exec-1} can be preempted
 by ensuring that updates are only ever applied in the causal order.
@@
 applied. This strategy would preempt the execution in
-Fig.~\ref{fig:crdt-execs-1} by buffering $R_2$'s \C{add} at $R_1$
+Fig.~\ref{fig:crdt-exec-1} by buffering $R_2$'s \C{add} at $R_1$
 until the causally-preceding \C{remove} of $R_1$ is received and
@@
 data type of Fig.~\ref{fig:cc-set} still admits divergent executions
-due to concurrent updates.  Fig.~\ref{fig:crdt-execs-2} describes one
+due to concurrent updates.  Fig.~\ref{fig:crdt-exec-2} describes one
 such execution. Here, replicas $R_1$ and $R_2$ both start with a
@@
 
-Note that, unlike the execution in Fig.~\ref{fig:crdt-execs-1}, the
-conflicting operations in Fig.~\ref{fig:crdt-execs-2}, namely $R_1$'s
+Note that, unlike the execution in Fig.~\ref{fig:crdt-exec-1}, the
+conflicting operations in Fig.~\ref{fig:crdt-exec-2}, namely $R_1$'s
 \C{remove} and $R_2$'s \C{add}, are \emph{not} causally related,
@@
 which are otherwise unordered. For example, in
-Fig.~\ref{fig:crdt-execs-2}, we might let $R_2$'s \C{add} override
+Fig.~\ref{fig:crdt-exec-2}, we might let $R_2$'s \C{add} override
 $R_1$'s \C{remove} considering that \C{add} is re-adding an
 element after a previous remove. Ordering concurrent removes ahead
-of adds uiniformly on all replicas leads to an implementation of
-\C{Set} RDT where concurrent (re-)adds consistenty \emph{win}
+of adds uniformly on all replicas leads to an implementation of
+\C{Set} RDT where concurrent (re-)adds consistently \emph{win}
 over removes. Such an \emph{add-wins} set is useful, for instance, to
@@
 the effect is simply a no-op. This strategy would result in convergent
-states despite the execution in Fig.~\ref{fig:crdt-execs-2} as $R_2$'s
+states despite the execution in Fig.~\ref{fig:crdt-exec-2} as $R_2$'s
 \C{remove} effectively becomes a no-op at $R_1$ due to there being at
@@
 system that always delivers messages in the causal order would
-automatically preempt the execution in Fig.~\ref{fig:crdt-execs-1}
+automatically preempt the execution in Fig.~\ref{fig:crdt-exec-1}
 without the need for additional intervention on behalf of the
@@
 have to reason about fine-grained causal dependencies to guarantee
-convergence, such as in the case of Fig.~\ref{fig:crdt-execs-2}. As we
-observed earlier, the execution in Fig.~\ref{fig:crdt-execs-2} seems
+convergence, such as in the case of Fig.~\ref{fig:crdt-exec-2}. As we
+observed earlier, the execution in Fig.~\ref{fig:crdt-exec-2} seems
 inevitable unless every pair of \C{add} and \C{remove} operations are
@@
     \caption{}
-    \label{fig:mrdt-execs-1}
+    \label{fig:mrdt-exec-1}
   \end{subfigure}
@@
     \caption{}
-    \label{fig:mrdt-execs-2}
+    \label{fig:mrdt-exec-2}
   \end{subfigure}
@@
 Equipped with the above set merge function, we can now consider the
-equivalent execution of Fig.~\ref{fig:crdt-execs-1} in state-centric
-model. The execution is shown in Fig.~\ref{fig:mrdt-execs-1}. The
+equivalent execution of Fig.~\ref{fig:crdt-exec-1} in state-centric
+model. The execution is shown in Fig.~\ref{fig:mrdt-exec-1}. The
 initial version on all three replicas ($v_0$) is the singleton set
 $\{e\}$. Applying operations to replicas creates new versions, e.g.,
-$v_1$ on $R_3$. Changes can be propagated by merging versions, e.g.,
-version $v_2$ on $R_2$ is a result of merging $v_0$ and $v_1$ in
-presence of their lowest common ancestor (LCA) version
-$v_0$\footnote{Versions $v_0$ and $v_1$ are not concurrent as the
+$v_1$ on $R_3$, and $v_2$ and $v_3$ on $R_2$. Changes can be propagated by merging versions, e.g.,
+version $v_2$ on $R_2$ is a result of merging $v_1$ into $v_0$ for
+which $v_0$ serves as the lowest common ancestor (LCA)
+  \footnote{Versions $v_0$ and $v_1$ are not concurrent as the
   former is an ancestor of the latter. Merging $v_1$ into $v_0$ is
@@
   Git parlance this is a \emph{fast forward} merge.}.
-Likewise $v_4$ on $R_1$ is created by merging $v_3$ and $v_0$ in
-presence of their LCA $v_0$. By the end of the execution, versions
-$v_4$ and $v_3$ at $R_1$ and $R_2$ (resp.) have witnessed the same set
-of operations, hence are in agreement.\GK{fix this.}
+Likewise $v_4$ on $R_1$ is created by merging $v_2$ into $v_0$ in
+presence of their LCA $v_0$. The next version $v_5$ on $R_1$ is the
+result of merging $R_2$'s $v_3$ into $R_1$'s $v_4$. The LCA for this
+merge is $v_2$. By the end of the execution, versions $v_5$ and $v_3$
+on $R_1$ and $R_2$ (resp.) have witnessed the same set of operations,
+hence are in agreement.
+
+\begin{figure}[ht]
+  \centering
+  \begin{subfigure}[t]{0.4\columnwidth}
+    \includegraphics[scale=0.35]{Figures/mrdt-execs-3}
+    \caption{}
+    \label{fig:mrdt-exec-3}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.57\columnwidth}
+    \includegraphics[scale=0.35]{Figures/mrdt-execs-4}
+    \caption{}
+    \label{fig:mrdt-exec-4}
+  \end{subfigure}
+  \caption{Executions demonstrating the difference between linearized
+  (former) and concurrent (latter) merges.}
+\label{fig:mrdt-execs-2}
+\end{figure}
 
@@
 Convergence however is not an inherent virtue of the state-centric
-replication model. Fig.~\ref{fig:mrdt-execs-2} shows the state-centric
-analogue of the execution in Fig.~\ref{fig:crdt-execs-2} which
+replication model. Fig.~\ref{fig:mrdt-exec-2} shows the state-centric
+analogue of the execution in Fig.~\ref{fig:crdt-exec-2} which
 diverges. Here $R_1$ and $R_2$ start with version $v_0 \,=\, \{e\}$.
@@
 \C{remove} to make $v_3$. Replica $R_1$ now obtains $R_2$'s changes by
-merging versions $V_1$ and $v_2$ to make new versions $v_4$ and $v_5$
+merging versions $v_1$ and $v_2$ to make new versions $v_4$ and $v_5$
 respectively. LCAs for these merges are $v_0$ and $v_1$. Concurrently,
@@
 
-Note that the anamolous execution in Fig.~\ref{fig:mrdt-execs-2} could
+Note that the anamolous execution in Fig.~\ref{fig:mrdt-exec-2} could
 have been avoided had the merges between $R_1$ and $R_2$ been
-linearized. Fig.~\ref{fig:mrdt-execs-3} shows an execution that only
-slightly differs from the one in Fig.~\ref{fig:mrdt-execs-2}. The
-difference is that the merges in Fig.~\ref{fig:mrdt-execs-3} happen
+linearized. Fig.~\ref{fig:mrdt-exec-3} shows an execution that only
+slightly differs from the one in Fig.~\ref{fig:mrdt-exec-2}. The
+difference is that the merges in Fig.~\ref{fig:mrdt-exec-3} happen
 linearly: first $R_1$ is merged into $R_2$ (bringing $R_1$'s
@@
 Another important point to note is that only the merges are
-linearized; not the entire execution. In Fig.~\ref{fig:mrdt-execs-3},
+linearized; not the entire execution. In Fig.~\ref{fig:mrdt-exec-3},
 both $R_1$ and $R_2$'s \C{remove}s remove the same element $e$, which
@@
 
-In the context of Fig.~\ref{fig:mrdt-execs-3}, it is quite clear what
+In the context of Fig.~\ref{fig:mrdt-exec-3}, it is quite clear what
 linearization of merges means and how to enforce it via
@@
 and dried. For instance, consider the execution in
-Fig.~\ref{fig:mrdt-execs-4}. The four replicas involved in the
+Fig.~\ref{fig:mrdt-exec-4}. The four replicas involved in the
 execution start with version $v_0 \,=\, \{e\}$. The replicas perform
 local operations as shown in the figure to make versions $v_1$ to
-$v_5$. Next they perform a series of merges to propagate local
+$v_3$. Next they perform a series of merges to propagate local
 changes. The merges can be ordered in time as following: first $R_2
-\rightarrow R_1$, then $R_3 \rightarrow R_1$, then $R_2 \rightarrow
-R_4$, and finally $R_3 \rightarrow R_4$. Despite linearly ordered in
-time, the merges nonetheless result in divergent execution. The
+\rightarrow R_1$ (merging $v_1$), then $R_3 \rightarrow R_1$ twice
+(merging $v_2$ and $v_3$ resp.), then $R_3 \rightarrow R_4$ (merging
+$v_3)$, and finally $R_2 \rightarrow R_4$ (merging $v_1$). These
+merges collectively propagate the effects of \C{add} and \C{remove}
+operations to $R_1$ and $R_2$. And despite being ordered in time, the
+they nonetheless result in a divergent execution ($v_5 \neq v_7$). The
 problem here is that, although merges are executed linearly, the
 execution graph does not reflect this linearity; merges of that end in
-$R_1$ in Fig.~\ref{fig:mrdt-execs-4} are effectively concurrent with
+$R_1$ in Fig.~\ref{fig:mrdt-exec-4} are effectively concurrent with
 those that end in $R_4$. This shows that simply synchronizing the
@@
 states. To understand well-formedness, let us contrast the bad
-executions in Figs.~\ref{fig:mrdt-execs-2} and~\ref{fig:mrdt-execs-4}
-against the good execution in Fig.~\ref{fig:mrdt-execs-3}. Observe
-that in Fig.~\ref{fig:mrdt-execs-2}, \emph{every} pair of concurrent
+executions in Figs.~\ref{fig:mrdt-exec-2} and~\ref{fig:mrdt-exec-4}
+against the good execution in Fig.~\ref{fig:mrdt-exec-3}. Observe that
+in Fig.~\ref{fig:mrdt-exec-2}, \emph{every} pair of concurrent
 versions on $R_1$ and $R_2$ have a unique lowest common ancestor
@@
 $v_3$, $(v_1,v_2)$ is $v_0$ and so on. By contrast in
-Fig.~\ref{fig:mrdt-execs-2}, versions $v_5$ and $v_6$ have two LCAs,
+Fig.~\ref{fig:mrdt-exec-2}, versions $v_5$ and $v_6$ have two LCAs,
 namely $v_2$ and $v_3$. Both these versions are common ancestors of
@@
 that are also common ancestors of $v_5$ and $v_6$. Likewise in
-Fig.~\ref{fig:mrdt-execs-4}, versions $v_7$ and $v_9$ have two LCAs --
+Fig.~\ref{fig:mrdt-exec-4}, versions $v_7$ and $v_9$ have two LCAs --
 $v_2$ and $v_4$. Multiple LCAs is an indication that there exist
@@
 concurrent versions is unique. As we prove in
-Sec.~\ref{sec:formalization}, enforcing this structural
+Sec.~\ref{sec:abstract-sem}, enforcing this structural
 well-formedness condition indeed guarantees the convergence of
-distributed executions. In Sec.~\ref{fig:implementation}, we describe
-a distributed runtime called \quark whose primary purpose is to
-enforce the well-formedness condition on distributed executions. The
-runtime makes it possible to automatically promote ordinary data types
-equipped with a merge function to convergent replicated data types.
+distributed executions. In Sec.~\ref{sec:concrete-sem}, we formalize a
+distributed runtime called \quark whose primary purpose is to enforce
+the well-formedness condition on distributed executions. In
+Sec.~\ref{sec:implementation}, we describe its implementation. The
+runtime makes it possible to automatically enforce convergence,
+letting us promote ordinary sequential data types to convergent
+replicated types by simply equipping them with a merge function.
 Thus, extending the \C{Set} interface of Fig.~\ref{fig:ocaml-set} with
@@
 obtain a convergent replicated set data type.
-
-The following section formalizes the aforementioned intuitions with
-help of an abstract machine that generates only the well-formed
-executions. \GK{fix this}

### 2021-11-20 evaluation.tex
Evaluation text in

\section{Evaluation}
\label{sec:eval}

We present an evaluation study of our runtime-assisted approach to
convergent RDTs with help of our prototype \quark implementation. The
benchmark we chose for the study is a collaborative document editing
application -- a common usecase addressed by several CRDT
proposals~\cite{rga, treedoc, lagoot, lseq}. However, unlike these
other approaches, we did not have to build a dedicated replicated data
type to represent collaboratively-edited documents; an ordinary
document format extended with a merge operation would suffice. While
many data structures exist to represent text documents (e.g.,
ropes~\cite{rope}), we decided to adopt the simplest representation of
a document as a list of characters. 
\begin{center}
\begin{ocaml}
        type doc = char list
\end{ocaml}
\end{center}
While being simple, the advantage of this presentation is that we can
simply reuse the three-way \C{List.merge} function of the list data
type to merge documents. \C{List.merge} is a simple implementation of
the GNU \C{diff3} algorithm~\cite{gnudiff} in ~60 lines of OCaml. We
thus adopt a straightforward approach to building a collaborative
document editor with the intention to keep the development effort low
enough to be easily replicated. The convergence guarantee of \quark
ensures that the simplicity of our implementation doesn't come at the
expense of correctness. We now describe an evaluation that
demonstrates that it also doesn't come at the expense of the editor
performance.

Our experiment setup consists of multiple collaborators simultaneously
editing a 10000+ line document obtained from the Canterbury
Corpus~\cite{canterbury}. Each user holds a replica of the document
and is assumed to be editing the document at the speed of 240
characters per minute or 1 character every 0.25s. At 6 characters per
word, this amounts to 40 words per minute, which is the average typing
speed of humans. Each edit is immediately persisted to the disk by
creating a new version in the backing store. Thus there are at least
as many versions of the document as there are edits. Such extensive
versioning may be considered excessive in practice and could be
disabled. Each user process runs a \quark thread that commits
user-generated document versions to the local branch, while
merging with the concurrent versions from the remote branches in the
background. Each merge is synchronized as described in previous
sections.

\quark's background merges however pose a new problem as they create
new versions on the local branch in the background while the user is
busy editing an older version. When the user attempts to write their
version of the document to the store, simply committing it would
effectively override the concurrent updates from other users obtained
via background merges. This problem is depicted in
Fig.~\ref{fig:local-merge}. The solution, fortunately, is
straightforward: we \emph{merge} the user-submitted value with the
(value of) the latest version on the local branch to create a new
version that includes the updates from either direction. The LCA for
this merge is simply the last version on the local branch read by the
user. This merge need not be synchronized as it doesn't alter the LCAs
between any two branches. The merged value can then returned
to the user as the result of the write.  Quark's \C{write} is a
function of type $\texttt{Value} \rightarrow \texttt{Value}$, where
$\texttt{Value}$ is \C{doc} in the current application.

To measure the impact of \quark runtime on user writes, we measure the
latency of the \C{write} operation, which includes for the time spent
merging the user version with the current version, and persisting the
resultant version to the store. To conduct the experiments, we deploy
a three-node cluster of \C{i3.large} machines in Amazon \C{us-west2}
data center. Each user connects to one of the machines and performs
1000 edits in succession, saving the document after each edit. We
progressively increase the number of concurrent users editing the
document from 3 to 300 and measure the impact of the increased
concurrency on write latency. We report three values of latency for
each run -- one each for the 10th percentile, 50th percentile, and
90th percentile.  Fig.~\ref{fig:latency} shows the results. As evident
from the figure, latency of \quark writes remain mostly constant,
increasing only slightly between $n=3$ to $n=300$.
Fig.~\ref{fig:latency} also plots the latency values for the baseline
``SC'' approach which achieves convergence by synchronizing each
operation, i.e., executing under strong consistency.

\quark system replicates the contents of each branch across all the
replicas as fast as the network allows. However, for a user $A$ to
see the changes made by the other user $B$, the changes have to be
reflected in $A$'s local version, which can only happen through a
merge operation. Since \quark synchronizes merge operations globally,
it induces additional delay before $A$ can see $B$'s changes. We call
this additional delay \emph{staleness} as with progression of time,
$B$'s version known to $A$ becomes increasingly stale. At the
system-level, an increase in staleness effectively delays the
convergence (but doesn't stop it, as proved by
Theorem~\ref{thm:progress}). To understand the staleness behavior in
\quark, we repeat the collaborative editing experiment, this time
measuring the staleness value at every merge. We do this by annotating
every version $v$ with the timestamp $t$ of its creation time. When
$v$ is is merged into a remote branch $b$ at a later time $t'$, the
difference $t' - t$ denotes the staleness of $v$ w.r.t the new version
on $b$. Multiple such staleness measurements are recorded for each
experiment and analyzed to compute the 10th, 50th, and 90th percentile
values. Fig.~\ref{fig:staleness} shows the results as we repeat
experiments with increasing number of concurrent users. While
staleness remains in the order of milliseconds with fewer ($\leq 9$)
concurrent users, it increases super-linearly as we increase the
number of collaborators by 3 until 30. While increased staleness is
inevitable in our approach due to synchronized merges, we believe the
increase can be contained by choosing the order of merges to avoid the
``starvation'' of some branches. Note that relaxing the
linearizability constraint on merges would completely eliminate the
staleness overhead, but the resultant Git-like system fails to
converge due to anomalous executions described in
Sec.~\ref{sec:motivation}\footnote{
  Git admits such anamolous version history graphs where two branches
  can have the same set of commits and yet differ in their final
  version. Appendix describes two such cases we observed on Github}.

Our experiments bring to the fore an inherent tradeoff among the
competing concerns of RDTs, namely (\rom{1}). The ease of achieving
convergence, (\rom{2}) Latency, and (\rom{3}).  Staleness. While CRDTs
try to optimize for latency and staleness, they require a significant
amount of development and verification effort to be expended to ensure
convergence~\cite{kleppmann2017}. In contrast, \quark lets developers
derive convergent-by-construction RDTs from ordinary data data types
that are optimized for latency, but incur a staleness overhead that
delays the time to convergence.

### 2021-11-20 implementation.tex
Implementation Fig added
Evaluation text in

@@
+\begin{figure}[ht]
+  \centering
+    \includegraphics[scale=0.35]{Figures/implementation2}
+\caption{\quark implementation architecture}
+\label{fig:implementation}
+\end{figure}
+
 \section{Implementation}
@@
 
-We implement \quark runtime as a lightweight shim layer on top of
-Scylla -- an off-the-shelf distributed data store~\cite{scylla}. We
-rely on Scylla for inter-replica communication, data replication,
-persistence, and fault tolerance. \quark translates the high-level
-mergeable data type implementations in OCaml to their low-level
-representations in the backing store and organizes their well-formed
-distributed executions. Fig.~\ref{fig:system} illustrates the overall
-architecture.
+We realize a prototype of \quark runtime as a lightweight shim layer
+on top of Scylla -- an off-the-shelf distributed data
+store~\cite{scylla}. We rely on Scylla for inter-replica
+communication, data replication, persistence, and fault tolerance.
+\quark translates the high-level mergeable data type implementations
+in OCaml to their low-level representations in the backing store and
+organizes their well-formed distributed executions.
+Fig.~\ref{fig:implementation} illustrates the overall architecture.
 

### 2021-11-20 evaluation.tex
Quark --> Carmot. Related work.

@@
 between any two branches. The merged value can then returned
-to the user as the result of the write.  Quark's \C{write} is a
+to the user as the result of the write.  \quark's \C{write} is a
 function of type $\texttt{Value} \rightarrow \texttt{Value}$, where
@@
 Sec.~\ref{sec:motivation}\footnote{
-  Git admits such anamolous version history graphs where two branches
-  can have the same set of commits and yet differ in their final
-  version. Appendix describes two such cases we observed on Github}.
+  Git admits anamolous version history graphs where two branches can
+  have the same set of commits and yet differ in their final version.
+  Appendix describes two such cases we observed on Github}.
 
 Our experiments bring to the fore an inherent tradeoff among the
-competing concerns of RDTs, namely (\rom{1}). The ease of achieving
+competing concerns of RDTs, namely (\rom{1}). The ease of programming
 convergence, (\rom{2}) Latency, and (\rom{3}).  Staleness. While CRDTs

### 2021-11-20 related.tex
Quark --> Carmot. Related work.

\section{Related Work}
\label{sec:related}

The idea of state-based replicated data types with three-way merge
functions was first proposed by Kaki et. al. in~\cite{mrdt},
where authors demonstrate an approach for deriving merge functions
from first principles for various data types. However, the resultant
\emph{mergeable replicated data types} (MRDTs) were never proved to be
convergent. As we demonstrate for the \C{Set} data type in
Sec.~\ref{sec:motivation}, simple merge functions with intuitive merge
semantics can nonetheless lead to divergent executions. The \quark
preempts such executions to guarantee convergence on MRDTs.

Concurrent Revisions~\cite{BBL+10} proposes versioned state for
shared-memory concurrent programs in order to guarantee determinism.
The fork-join concurrency they consider is however too restrictive for
the asynchronous distributed setting where replicas do not have a
single coherent view of the system state.  Furthermore the
hierarchical organization of the concurrent actors with a designated
\emph{server} is inconsistent with the fully-decentralized model of
execution adopted by several collaborative applications.

State-centric replication was also explored in the context of
CRDTs~\cite{crdts}. However, such state-based CRDTs require the
replicated state to be organized as a lattice with the merge function
acting as as a least upper bound operator. We eliminate this
restriction in our setting with help of the \quark runtime. A thorough
comparison of our approach with the operation-based CRDTs can be found
in Secs.~\ref{sec:introduction} and~\ref{sec:motivation}.

Several verification techniques, program analyses, and tools have been
proposed to reason about the program behavior in a weakly-consistent
distributed setting~\cite{bailis-vldb, alvaro-calm,
gotsman-popl16,redblue-atc, redblue-osdi, ecinec}. These techniques
treat replicated storage as a black box with a fixed pre-defined
consistency models. The focus is on assigning appropriate consistency
levels to operations so as to preserve application integrity. Such an
approach results in assigning sequential consistency (SC) to all
operations since the next weaker consistency model -- causal
consistency, is insufficient to guarantee convergence (as
Sec.~\ref{sec:motivation} demonstrates). Conversely, \quark cannot
reason about application-level invariants, such as $\C{balance}\ge 0$
in an banking application. Thus both approaches confer complimentary
benefits on application developers.

Finally, the implementation of \quark bears resemblance to
Quelea~\cite{pldi15}. Both these systems are implemented as shim
layers on top of off-the-shelf distributed stores, and both offload
low-level concerns to the underlying store. However, both these
systems serve orthogonal purposes. Quelea enforces high-level
invariants on applications built with CRDTs. Operations are assumed to
be commutative and applications are assumed to be convergent. In
contrast, \quark enforces convergence of data types that are otherwise
not guaranteed to converge.

### 2021-11-20 abstract-semantics.tex
Edits

@@
 from previous section suggests, our replication scheme is strongly
-inspired by version control systems (VCS) such as Git~\cite{git}.  We
+inspired by version control systems (VCS) such as Git.  We
 embrace this analogy in our formal development to manifest distributed

### 2021-11-20 concrete-semantics.tex
Edits

@@
 distributed machine is defined in Fig.~\ref{fig:quark-semantics}.
-There is one rule corresponding to every rule in
-Fig.~\ref{fig:git-semantics} (abstract semantics). \rulelabel{Commit}
-describes replica $i$ committing a new version on the branch $b_i$
-with the given value $n$.  The vector $t'$ for the new version is
-obtained from that of the previous version $t$ by incrementing the
-component $b_i$. The sequence number of the new version on $b_i$ is
-$t'(b_i)$ and the branch map $B_i$ is updated to reflect that. Note
-that for all $j \neq i$, $H_j$ and $B_j$ remain unchanged indicating
-that other replicas are not (yet) aware of this commit.
+Every transition rule of the abstract machine
+(Fig.~\ref{fig:git-semantics}) has a corresponding rule for the
+abstract machine. Fig.~\ref{fig:quark-semantics} however elides
+\rulelabel{Fork} and \rulelabel{FastFwd} in the interest of space.
+The rule \rulelabel{Commit} describes replica $i$ committing a new
+version on the branch $b_i$ with the given value $n$.  The vector $t'$
+for the new version is obtained from that of the previous version $t$
+by incrementing the component $b_i$. The sequence number of the new
+version on $b_i$ is $t'(b_i)$ and the branch map $B_i$ is updated to
+reflect that. Note that for all $j \neq i$, $H_j$ and $B_j$ remain
+unchanged indicating that other replicas are not (yet) aware of this
+commit.
 %
@@
 to retain the performance benefits of asynchronous replication.
-\rulelabel{FastFwd} is similar to \rulelabel{Merge}, so we don't
-discuss it separately.
+% \rulelabel{FastFwd} is similar to \rulelabel{Merge}, so we don't
+% discuss it separately.
 
@@
 
+\paragraph{Properties} The Unique LCA
+(Lemma~\ref{lem:lca-uniqueness}), Convergence
+(Theorem~\ref{thm:convergence}, and Progress
+(Theorem~\ref{thm:progress}) properties have been re-defined for the
+\quark distributed machine and automatically proven correct in
+Ivy~\cite{ivy}.
+
 \paragraph{Garbage Collection} One downside of the distributed
@@
 execution progresses and new versions are created. Fortunately, this
-is easy to address as the execution only needs version history
-inasmuch as it is required to obtain the LCA versions for
-\rulelabel{Merge}. Since LCA version vectors monotonically increase,
+is easy to address as the execution only needs finite version history
+to compute LCAs. Since LCA version vectors monotonically increase,
 older versions with vectors less than the least known LCA vector can

### 2021-11-20 evaluation.tex
Edits

@@
 application -- a common usecase addressed by several CRDT
-proposals~\cite{rga, treedoc, lagoot, lseq}. However, unlike these
+proposals~\cite{rga, treedoc, crdts}. However, unlike these
 other approaches, we did not have to build a dedicated replicated data
@@
 effectively override the concurrent updates from other users obtained
-via background merges. This problem is depicted in
-Fig.~\ref{fig:local-merge}. The solution, fortunately, is
-straightforward: we \emph{merge} the user-submitted value with the
-(value of) the latest version on the local branch to create a new
-version that includes the updates from either direction. The LCA for
-this merge is simply the last version on the local branch read by the
-user. This merge need not be synchronized as it doesn't alter the LCAs
-between any two branches. The merged value can then returned
-to the user as the result of the write.  \quark's \C{write} is a
-function of type $\texttt{Value} \rightarrow \texttt{Value}$, where
-$\texttt{Value}$ is \C{doc} in the current application.
+via background merges. The solution, fortunately, is straightforward:
+we \emph{merge} the user-submitted value with the (value of) the
+latest version on the local branch to create a new version that
+includes the updates from either direction. 
+% The LCA for this merge is
+% simply the last version on the local branch read by the user. This
+% merge need not be synchronized as it doesn't alter the LCAs between
+% any two branches. The merged value can then returned to the user as
+% the result of the write.  \quark's \C{write} is a function of type
+$\texttt{Value} \rightarrow \texttt{Value}$, where $\texttt{Value}$ is
+\C{doc} in the current application.
+
+\begin{figure}[ht]
+  \centering
+    \includegraphics[scale=0.3]{Figures/latency}
+\caption{Latency of collaborative editing operations under \quark vs
+  executing them with strong consistency (SC). }
+\label{fig:latency}
+\end{figure}
 
@@
 merging the user version with the current version, and persisting the
-resultant version to the store. To conduct the experiments, we deploy
-a three-node cluster of \C{i3.large} machines in Amazon \C{us-west2}
+resultant version to the store. We conduct the experiments on a
+three-node cluster of \C{i3.large} machines in Amazon \C{us-west2}
 data center. Each user connects to one of the machines and performs
@@
 progressively increase the number of concurrent users editing the
-document from 3 to 300 and measure the impact of the increased
-concurrency on write latency. We report three values of latency for
-each run -- one each for the 10th percentile, 50th percentile, and
-90th percentile.  Fig.~\ref{fig:latency} shows the results. As evident
-from the figure, latency of \quark writes remain mostly constant,
-increasing only slightly between $n=3$ to $n=300$.
-Fig.~\ref{fig:latency} also plots the latency values for the baseline
-``SC'' approach which achieves convergence by synchronizing each
-operation, i.e., executing under strong consistency.
+document from 3 to 60 and measure the impact of the increased
+concurrency on write latency. Fig.~\ref{fig:latency} shows the median
+latency values. As evident from the figure, latency of \quark writes
+constant with the median latency around 0.008s. The maximum latency
+ever measured for a \quark writes is 0.016s. Notably, the latency
+values are an order of magnitude less than the time between
+consecutive edits (0.25s), which lets users see their edits almost
+immediately. Fig.~\ref{fig:latency} also plots the latency values for
+the baseline ``SC'' approach which achieves convergence by
+synchronizing each operation, i.e., executing under strong
+consistency. The SC write latency increases super-linearly with the
+median value of 1.76s for 60 concurrent editors. The maximum latency
+measured for an SC write is 9.02s, which is considerably more than the
+inter-edit latency of 0.25s. The experiment confirms the theory that
+convergence does not come at the cost of latency in \quark.
+
+\begin{figure}[ht]
+  \centering
+    \includegraphics[scale=0.3]{Figures/staleness}
+\caption{Staleness increases as the number of concurrent editors
+  increase.}
+\label{fig:staleness}
+  \vspace*{-0.2in}
+\end{figure}
 
@@
 experiment and analyzed to compute the 10th, 50th, and 90th percentile
-values. Fig.~\ref{fig:staleness} shows the results as we repeat
-experiments with increasing number of concurrent users. While
+values. Fig.~\ref{fig:staleness} shows the results. While
 staleness remains in the order of milliseconds with fewer ($\leq 9$)
 concurrent users, it increases super-linearly as we increase the
-number of collaborators by 3 until 30. While increased staleness is
-inevitable in our approach due to synchronized merges, we believe the
-increase can be contained by choosing the order of merges to avoid the
-``starvation'' of some branches. Note that relaxing the
-linearizability constraint on merges would completely eliminate the
-staleness overhead, but the resultant Git-like system fails to
-converge due to anomalous executions described in
+number of collaborators in the multiples of 3 until 30. While
+increased staleness is inevitable in our approach due to synchronized
+merges, we believe the increase can be contained by choosing the order
+of merges to avoid the ``starvation'' of some branches. Note that
+relaxing the linearizability constraint on merges would completely
+eliminate the staleness overhead, but the resultant Git-like system
+fails to converge due to anomalous executions described in
 Sec.~\ref{sec:motivation}\footnote{
@@
   have the same set of commits and yet differ in their final version.
-  Appendix describes two such cases we observed on Github}.
+  Supplementary material describes two such cases we observed on
+  Github.}.
 

### 2021-11-20 implementation.tex
Edits

@@
 \label{fig:implementation}
+  \vspace*{-0.2in}
 \end{figure}
@@
 
-The runtime \quark is a straightforward implementation the of
-distributed machine \quark described in Sec.~\ref{sec:concrete-sem}.
+The implementation of \quark the is a straightforward realization of
+the \quark distributed machine described in Sec.~\ref{sec:concrete-sem}.
 We manifest each component of the state, namely the branch map $B$,
@@
 enforced with help of \C{Quorum} reads and writes. Each user process
-is assigned its own replica of the RDT.  Replica id is the
-concatenation of its host name, process id, and Unix timestamp.
-Version vectors are realized as associative lists and stored in Scylla
+is assigned its own replica of the RDT.  Version vectors are realized
+as associative lists and stored in Scylla
 as blobs.

### 2021-11-20 introduction.tex
Figures and Progress thm added.
Edits

@@
 \section{Introduction}
-\label{sec:intro}
+\label{sec:introduction}
 
@@
     a an abstract machine (named \quark) that manifests distributed
-    executions as version history graphs (similar to Git~\cite{git}).
+    executions as version history graphs (similar to Git).
     We prove the \emph{convergence} and \emph{progress} properties of
@@
     store~\cite{scylla}. We evaluate the implementation on a
-    collaborative-editing benchmark to demonstrate novel tradeoffs
+    collaborative-editing benchmark to demonstrate novel trade-offs
     presented by our approach.

### 2021-11-20 motivation.tex
Edits

@@
 implementation of an eventually consistent system without weakening
-its guarantees~\cite{bolton}. Unfortunately, such strengthening of the
+its guarantees~\cite{BoltOn}. Unfortunately, such strengthening of the
 system model would deliver no benefits to the developer if they still
@@
 \label{fig:mrdt-execs-2}
+\vspace*{-0.25in}
 \end{figure}

### 2021-11-20 quark.tex
Implementation section
Evaluation text in
Quark --> Carmot. Related work.
Edits

@@
 %% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
+\citestyle{acmnumeric}   %% For author/year citations
 
@@
   extending their implementations rather than refactoring them.
-  The cornerstone of our approach is a distributed runtime that
+  The novel component of our approach is a distributed runtime that
   orchestrates \emph{well-formed } executions that are guaranteed to
   converge. Despite the utilization of a runtime, our approach comes
-  at no additional cost of per-operation latency and system-wide
-  availability. Instead, the tradeoff is the time required to achieve
-  convergence -- a metric we call \emph{staleness}. We quantify this
-  tradeoff via a thorough evaluation of a collaborative editing
-  application. and demonstrate a novel practical path towards
-  achieving high scalability through geo-replication.
+  at no additional cost of latency and availability. We implement our
+  approach in a system called \quark and evaluate its performance over
+  a large-scale collaborative editing application. 
 \end{abstract}
@@
 
-\input{vc-semantics}
+\input{abstract-semantics}
 
 \input{concrete-semantics}
+
+\input{implementation}
+
+\input{evaluation}
+
+\input{related}
+
+\bibliography{all}
 
@@
 %% Appendix
-\appendix
-\section{Appendix}
+% \appendix
+% \section{Appendix}
 

### 2021-11-20 related.tex
Edits

@@
 The idea of state-based replicated data types with three-way merge
-functions was first proposed by Kaki et. al. in~\cite{mrdt},
-where authors demonstrate an approach for deriving merge functions
-from first principles for various data types. However, the resultant
-\emph{mergeable replicated data types} (MRDTs) were never proved to be
-convergent. As we demonstrate for the \C{Set} data type in
-Sec.~\ref{sec:motivation}, simple merge functions with intuitive merge
-semantics can nonetheless lead to divergent executions. The \quark
-preempts such executions to guarantee convergence on MRDTs.
+functions was first proposed in~\cite{mrdt}, where authors demonstrate
+an approach for deriving merge functions from first principles for
+various data types. However, the resultant \emph{mergeable replicated
+data types} (MRDTs) were never proved to be convergent. As we
+demonstrate for the \C{Set} data type in Sec.~\ref{sec:motivation},
+simple merge functions with intuitive merge semantics can nonetheless
+lead to divergent executions. The \quark preempts such executions to
+guarantee convergence on MRDTs.
 
@@
 reason about application-level invariants, such as $\C{balance}\ge 0$
-in an banking application. Thus both approaches confer complimentary
+in a banking application. Thus both approaches confer complimentary
 benefits on application developers.

### 2021-11-20 evaluation.tex
Edits

@@
 many data structures exist to represent text documents (e.g.,
-ropes~\cite{rope}), we decided to adopt the simplest representation of
-a document as a list of characters. 
+ropes~\cite{boehm95}), we decided to adopt the simplest representation
+of a document as a list of characters. 
 \begin{center}
@@
 % any two branches. The merged value can then returned to the user as
-% the result of the write.  \quark's \C{write} is a function of type
-$\texttt{Value} \rightarrow \texttt{Value}$, where $\texttt{Value}$ is
-\C{doc} in the current application.
+% the result of the write.
+Thus, \quark's \C{write} is a function of type $\texttt{Value}
+\rightarrow \texttt{Value}$, where $\texttt{Value}$ is \C{doc} in the
+current application.
 
@@
 latency values. As evident from the figure, latency of \quark writes
-constant with the median latency around 0.008s. The maximum latency
-ever measured for a \quark writes is 0.016s. Notably, the latency
-values are an order of magnitude less than the time between
-consecutive edits (0.25s), which lets users see their edits almost
-immediately. Fig.~\ref{fig:latency} also plots the latency values for
-the baseline ``SC'' approach which achieves convergence by
-synchronizing each operation, i.e., executing under strong
-consistency. The SC write latency increases super-linearly with the
-median value of 1.76s for 60 concurrent editors. The maximum latency
-measured for an SC write is 9.02s, which is considerably more than the
-inter-edit latency of 0.25s. The experiment confirms the theory that
-convergence does not come at the cost of latency in \quark.
+remain more-or-less constant with the median latency around 0.008s.
+The maximum latency ever measured for a \quark writes is 0.016s.
+Notably, latency remains an order of magnitude less than the time
+between consecutive edits (0.25s), making it hard to perceive.
+Fig.~\ref{fig:latency} also plots the latency values for the baseline
+``SC'' approach which achieves convergence by synchronizing each
+operation, i.e., executing under strong consistency (SC). The SC write
+latency increases super-linearly with the median value of 1.76s for 60
+concurrent editors. The maximum latency measured for an SC write is
+9.02s, which is considerably more than the inter-edit latency of
+0.25s. The experiment confirms the theory that convergence does not
+come at the cost of latency in
+\quark.
 
@@
 it induces additional delay before $A$ can see $B$'s changes. We call
-this additional delay \emph{staleness} as with progression of time,
-$B$'s version known to $A$ becomes increasingly stale. At the
+this additional delay \emph{staleness} as with the progression of
+time, $B$'s version known to $A$ becomes increasingly stale. At the
 system-level, an increase in staleness effectively delays the
-convergence (but doesn't stop it, as proved by
+convergence (but doesn't preempt it, as proved by
 Theorem~\ref{thm:progress}). To understand the staleness behavior in
@@
 on $b$. Multiple such staleness measurements are recorded for each
-experiment and analyzed to compute the 10th, 50th, and 90th percentile
+experiment to compute the 10th, 50th, and 90th percentile
 values. Fig.~\ref{fig:staleness} shows the results. While

### 2021-11-20 introduction.tex
Edits

@@
 theorem~\cite{cap}, but it is not the case in fact. While CAP theorem
-does rule out linearizability, it does not prohibit enforcing weaker
-consistency constraints on distributed executions. Our approach
-exploits this observation by (\rom{1}). Identifying a novel set of
-constraints on distributed executions that guarantee their
-convergence, and (\rom{2}). Localizing such constraints on
-(asynchronously executed) state merges such that per-operation latency
-and system-wide availability remain unaffected.
+does rule out linearizability (as perceived by the user), it does not
+prohibit enforcing weaker consistency constraints on distributed
+executions. Our approach exploits this observation by (\rom{1}).
+Identifying a novel set of constraints on distributed executions that
+guarantee their convergence, and (\rom{2}). Localizing such
+constraints on (asynchronously executed) state merges such that
+per-operation latency and system-wide availability remain unaffected.
 Sec.~\ref{sec:concrete-sem} explains our runtime in detail.

### 2021-11-20 related.tex
Edits

@@
 
-The idea of state-based replicated data types with three-way merge
-functions was first proposed in~\cite{mrdt}, where authors demonstrate
-an approach for deriving merge functions from first principles for
-various data types. However, the resultant \emph{mergeable replicated
-data types} (MRDTs) were never proved to be convergent. As we
-demonstrate for the \C{Set} data type in Sec.~\ref{sec:motivation},
-simple merge functions with intuitive merge semantics can nonetheless
-lead to divergent executions. The \quark preempts such executions to
-guarantee convergence on MRDTs.
+The idea of replicated data types with three-way merge functions was
+first proposed in~\cite{mrdt}, where authors demonstrate an approach
+for deriving merge functions from first principles for various data
+types. However, the resultant \emph{mergeable replicated data types}
+(MRDTs) were never proved to be convergent. As we demonstrate for
+the \C{Set} data type in Sec.~\ref{sec:motivation}, simple merge
+functions with intuitive merge semantics can nonetheless lead to
+divergent executions. \quark can thus be used with MRDTs to preempt
+such executions and guarantee their convergence.
 
@@
 layers on top of off-the-shelf distributed stores, and both offload
-low-level concerns to the underlying store. However, both these
-systems serve orthogonal purposes. Quelea enforces high-level
-invariants on applications built with CRDTs. Operations are assumed to
-be commutative and applications are assumed to be convergent. In
-contrast, \quark enforces convergence of data types that are otherwise
-not guaranteed to converge.
+low-level concerns to the underlying store. Quelea however enforces
+high-level invariants on applications built with CRDTs. Operations are
+assumed to be commutative and applications are assumed to be
+convergent. In contrast, \quark enforces convergence of data types
+that are otherwise not guaranteed to converge.

### 2022-01-23 abstract.tex
Committing last minute edits.

--- 2021-11-20 quark.tex
@@
-%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[sigplan,10pt,review,anonymous]{acmart}
-\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[acmsmall]{acmart}\settopmatter{}
-
-%\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{xspace}
-\usepackage{mathtools}
-\usepackage{mathpartir}
-\usepackage{ifpdf}
-\usepackage{graphicx}
-%\usepackage[usenames,dvipsnames]{color}
-\usepackage{stmaryrd}
-%\usepackage[numbers]{natbib}
-\usepackage{amsthm}
-\usepackage{listings}          % format code
-\usepackage{wrapfig}
-\usepackage{textcomp}
-\usepackage{tabularx}
-\usepackage{color}
-\usepackage{url}
-\usepackage{tikz}
-\usepackage{multirow,array}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{microtype}
-
-\input{mydefs.tex}
-
-%% Journal information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
-\acmArticle{1}
-\acmYear{2022}
-\acmMonth{1}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmnumeric}   %% For author/year citations
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Note: Authors migrating a paper from PACMPL format to traditional
-%% SIGPLAN proceedings format must update the '\documentclass' and
-%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-%% Title information
-\title[]{Runtime-Assisted Convergence in Replicated Data Types}
-%% [Short Title] is optional;
-                                        %% when present, will be used in
-                                        %% header instead of Full Title.
-%\titlenote{with title note}             %% \titlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-%\subtitle{Subtitle}                     %% \subtitle is optional
-%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-
-%% Author information
-%% Contents and number of authors suppressed with 'anonymous'.
-%% Each author should be introduced by \author, followed by
-%% \authornote (optional), \orcid (optional), \affiliation, and
-%% \email.
-%% An author may have multiple affiliations and/or emails; repeat the
-%% appropriate command.
-%% Many elements are not rendered, but should be provided for metadata
-%% extraction tools.
-
-%% Author with single affiliation.
-\author{First1 Last1}
-\authornote{with author1 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-  \position{Position1}
-  \department{Department1}              %% \department is recommended
-  \institution{Institution1}            %% \institution is required
-  \streetaddress{Street1 Address1}
-  \city{City1}
-  \state{State1}
-  \postcode{Post-Code1}
-  \country{Country1}                    %% \country is recommended
-}
-\email{first1.last1@inst1.edu}          %% \email is recommended
-
-%% Author with two affiliations and emails.
-\author{First2 Last2}
-\authornote{with author2 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}                   %% \country is recommended
-}
-\email{first2.last2@inst2a.com}         %% \email is recommended
-\affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}                   %% \country is recommended
-}
-\email{first2.last2@inst2b.org}         %% \email is recommended
-
-%% Abstract
-%% Note: \begin{abstract}...\end{abstract} environment must come
-%% before \maketitle command
-\begin{abstract}
-
-  We propose a runtime-assisted approach to enforce convergence in
-  distributed executions of replicated data types. The key
-  distinguishing aspect of our approach is that it guarantees
-  convergence \emph{unconditionally} -- without requiring data
-  type operations to satisfy algebraic laws such as commutativity and
-  idempotence. Consequently, programmers are no longer obligated to
-  prove convergence on a per-type basis. Moreover, our approach lets
-  sequential data types be reused in a distributed setting by
-  extending their implementations rather than refactoring them.
-  The novel component of our approach is a distributed runtime that
-  orchestrates \emph{well-formed } executions that are guaranteed to
-  converge. Despite the utilization of a runtime, our approach comes
-  at no additional cost of latency and availability. We implement our
-  approach in a system called \quark and evaluate its performance over
-  a large-scale collaborative editing application. 
-\end{abstract}
-
-%% 2012 ACM Computing Classification System (CSS) concepts
-%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-<concept>
-<concept_id>10003456.10003457.10003521.10003525</concept_id>
-<concept_desc>Social and professional topics~History of programming languages</concept_desc>
-<concept_significance>300</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
-%% End of generated code
-
-%% Keywords
-%% comma separated list
-\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
-
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
-\maketitle
-
-\input{introduction}
-
-\input{motivation}
-
-\input{abstract-semantics}
-
-\input{concrete-semantics}
-
-\input{implementation}
-
-\input{evaluation}
-
-\input{related}
-
-\bibliography{all}
-
-%% Acknowledgments
-\begin{acks}                            %% acks environment is optional
-                                        %% contents suppressed with 'anonymous'
-  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-  %% acknowledge financial support and will be used by metadata
-  %% extraction tools.
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
-\end{acks}
-
-%% Bibliography
-%\bibliography{bibfile}
-
-%% Appendix
-% \appendix
-% \section{Appendix}
-
-Text of appendix \ldots
-
-\end{document}
+We propose a runtime-assisted approach to enforce convergence in distributed executions of replicated data types.
+The key distinguishing aspect of our approach is that it guarantees convergence \emph{unconditionally} -- without requiring data type operations to satisfy algebraic laws such as commutativity and idempotence.
+Consequently, programmers are no longer obligated to prove convergence on a per-type basis.
+Moreover, our approach lets sequential data types be reused in a distributed setting by extending their implementations rather than refactoring them.
+The novel component of our approach is a distributed runtime that orchestrates \emph{well-formed } executions that are guaranteed to converge.
+Despite the utilization of a runtime, our approach comes at no additional cost of latency and availability.
+We implement our approach in a system called \quark and conduct a thorough evaluation of its performance over a large-scale collaborative editing application.

### 2022-02-09 pldi22_rebuttal.md
PLDI rebuttal

We thank the reviewers for their valuable feedback. We address major
concerns first, followed by reviewer-specific comments.

Use of Synchronization
----------------------

Merges in Carmot use synchronization. However, synchronization does
not trivially guarantee convergence. Our contribution is to show
*how* to use synchronization in a way that (a). Replicas are always
available for user actions, (b). No user action has to wait for a
remote replica to respond, and (c). Replicas eventually converge. To
demonstrate the subtlety in our use of synchronization, consider the
following trivial synchronizations:

1. Each operation is synchronized -- the system would violate
   aforementioned properties (a) and (b).
2. Merges ``stop the world'' to pull the latest version from all the
   branches -- violates properties (a) and (b) as no replica can
   progress when one replica is merging.
3. Only merges are synchronized; each merge is aware of, and has all
   the information from, the previous merges -- this would linearize
   merges *in time*, but nonethless results in ill-formed histories
   that violate property (c). Fig e.g., in Fig.5(b) merges could be
   ordered linearly in time, yet result in divergence.

Our key discovery is the LCA-based premise of Merge and FastFwd
rules that effectively ensure that LCAs in any history are organized
as a lattice. This inturn is based on our observation that multiple
LCAs inevitably lead to divergence in general case. We use
synchronization to enable the checks on LCAs and enforce the lattice
property (Corollary 3.6). Otherwise Synchronization in itself doesn't
constitute the solution.

Comparison with MRDTs and CRDTs
--------------------------------

Despite their best efforts to ensure convergence, Kaki et al's MRDT
merges [13] nonetheless result in divergence. For e.g., our
`Set.merge` in Sec.1 is same as Kaki et al's set merge, yet it results
in divergence as shown in Figs.4(b) and 5(b). The existence of these
counterexamples is in fact why they were not able to prove MRDT
convergence (source: personal communication). While convergence is
only a small aspect of full-functional correctness, ensuring it in the
context of CRDTs and MRDTs remains extremely challenging even for the
experts
([here](https://twitter.com/martinkl/status/1327025979454263297?lang=en)
is a recent example from CRDTs).  We hope our formal guarantees of
convergence would make RDTs accessible to non-experts.

Evaluation
------------

We have chosen to do one in-depth case study instead of experimenting
with a collection of RDTs as it lets us examine the tradeoffs more
closely. All the MRDT benchmarks from Kaki et al continue to work on
Carmot with no increase in latency and diff size measurements. Carmot
is in fact a fork of Quark -- the system from Kaki et al. Carmot
extends Quark with an more efficient network and storage layer (due to
Scylla), and a runtime component that is activated only during a merge
(happens in a background thread). **There is no additional overhead
associated with a user action**. When a user submits an action, it is
immediately executed and committed to the local version. Thus, insofar
as user interaction is concerned, Carmot is qualitatively no different
from any other CRDT system. Unfortunately, we couldn't find a
standardized CRDT system (akin to Quark for MRDTs) for us to do
comparisons against.

It is possible to give theoretical upper bound on staleness in
Carmot. Ring topology guarantees the uniqueness of LCAs (Lemma 3.4) by
construction (without a need for synchronization). If replicas are
organized as a ring, the maximum staleness is the sum of latencies on
the circumference of the ring. Hence if we have replicas distributed
across US-east and US-west, it is reasonable to expect staleness to be
in the order of 100ms. As it stands now, our implementation simply
reflects our formal model, which includes no such optimizations.
Lastly, we'd like to note that mere presence of synchronization doesn't
necessarily lead to unacceptable performance. Google docs, for
instance, relies on a centralized OT server to coordinate all its
edits and yet has acceptable performance.

Reviewer A
----------

* The concrete semantics of Sec.4 *refines* the abstract semantics of
  Sec. 3. The proof we did in Ivy is in fact a refinement proof (can be
  found in `quarkRefineImplement.ivy` in the supplementary material). We
  are incorrect in stating that all properties have been re-proved for
  the concrete semantics. The fact is that the properties follow
  because Sec.4 semantics is a refinement of Sec.3 semantics. We
  apologize for the confusion.

* For comments on Related Work and Evaluation, please see above.

Reviewer B
----------

* Useful merge functions exist for more complicated data structures.
  In [13] Kaki et al provide a constructive definition of well-formed
  merges with help of *relational decomposition*. Albeit
  computationally inefficient, relational decomposition leads to
  simple yet useful merges for a range of data structures. The
  problem, however, is that [13] lacks either a system or a proof that
  guarantees the convergence of resultant MRDTs. As demonstrated
  above, divergence is possible despite a sensible merge semantics.
  Carmot closes this gap.

Reviewer C
----------

* We hope your concerns on staleness and performance are addressed by
  our note on Evaluation above.

Reviewer D
-----------

* Every client forks off a branch which is effectively a replica, so
  there are multiple replicas per machine. With 30 clients, there are
  30 branches with 30 replicas distributed across 3 machines.

* Inclusion of a runtime imposes no additional overhead on the latency
  of individual operations. Given that Carmot runtime uses
  synchronization for merges, and the overhead of synchronization
  increases at least linearly with the number of replicas, one would
  expect the latency of each operation to grow at least linearly if it
  is affected by the presence of runtime. However, as Fig.10
  demonstrates, that's not the case. Hence the conclusion that our
  approach imposes no additional cost on latency and availability (our
  latency measurement accounts for availability too as it measures the
  time from the operation is submitted to the time a result is
  received. If the availability of the system goes down, then
  operations have to wait longer to find a replica, increasing their
  overall latency. Fig.10 shows this is not the case with Carmot.)

* We hope your other comments on performance are addressed by our note
  on Evaluation above.

### 2022-02-09 pldi22_rebuttal.md
PLDI Rebuttal edited

@@
 
-Merges in Carmot use synchronization. However, synchronization does
-not trivially guarantee convergence. Our contribution is to show
-*how* to use synchronization in a way that (a). Replicas are always
-available for user actions, (b). No user action has to wait for a
-remote replica to respond, and (c). Replicas eventually converge. To
-demonstrate the subtlety in our use of synchronization, consider the
-following trivial synchronizations:
+While merges in Carmot use synchronization, it is only one aspect of
+our overall approach. Our contribution is to show *how* to use
+synchronization in a way that (a). Replicas are always available for
+user actions, (b). No user action has to wait for a remote replica to
+respond, and (c). Replicas eventually converge. There are several ways
+one could use synchronization yet fail to achieve the aforementioned
+objectives. For e.g:
 
-1. Each operation is synchronized -- the system would violate
-   aforementioned properties (a) and (b).
-2. Merges ``stop the world'' to pull the latest version from all the
-   branches -- violates properties (a) and (b) as no replica can
-   progress when one replica is merging.
-3. Only merges are synchronized; each merge is aware of, and has all
-   the information from, the previous merges -- this would linearize
-   merges *in time*, but nonethless results in ill-formed histories
-   that violate property (c). Fig e.g., in Fig.5(b) merges could be
-   ordered linearly in time, yet result in divergence.
+1. Each user operation is synchronized: this leads to violation of
+   properties (a) and (b).
+2. Merges synchronize by ``stoping the world'' to pull the latest
+   version from all the branches. This too violates properties (a) and
+   (b) as no replica can progress when one replica is merging.
+3. Merges are synchronized with each other; each merge is aware of,
+   and has all the information from, the previous merges: this would
+   linearize merges *in time*, but nonethless results in ill-formed
+   histories that violate property (c). Fig e.g., in Fig.5(b) merges
+   could be ordered linearly in time, yet result in divergence.
 
-Our key discovery is the LCA-based premise of Merge and FastFwd
-rules that effectively ensure that LCAs in any history are organized
-as a lattice. This inturn is based on our observation that multiple
-LCAs inevitably lead to divergence in general case. We use
-synchronization to enable the checks on LCAs and enforce the lattice
-property (Corollary 3.6). Otherwise Synchronization in itself doesn't
-constitute the solution.
+As summarized by Reviewer B, our key discovery is the well-formedness
+condition that translates to LCA-based premises of Merge and FastFwd
+rules. These conditions effectively ensure that LCAs in any history
+DAG are organized as a lattice, which gives us the unique LCA property
+(Lemma 3.4), and thereby convergence (Theorem 3.7). The reasoning
+about LCAs is critical to convergence, and cannot be done away with.
+We use synchronization to perform the the checks on LCAs, which inturn
+guarantees convergence. Otherwise synchronization in itself doesn't
+guarantee convergence.
 
@@
 convergence (source: personal communication). While convergence is
-only a small aspect of full-functional correctness, ensuring it in the
-context of CRDTs and MRDTs remains extremely challenging even for the
-experts
-([here](https://twitter.com/martinkl/status/1327025979454263297?lang=en)
-is a recent example from CRDTs).  We hope our formal guarantees of
-convergence would make RDTs accessible to non-experts.
+only a small aspect of full-functional correctness, getting it right
+is not easy, even for the experts
+([Here](https://twitter.com/martinkl/status/1327025979454263297?lang=en)
+is a recent example from CRDTs. Another example is the queue
+persistent data structure from [Farinier et
+al](https://hal.inria.fr/hal-01099136v1/document), which turned out to
+admit the multiple LCA anomaly and (thereby) divergence). We hope
+our formal guarantees help the community advance the discussion on
+RDTs past convergence.
 
@@
 Scylla), and a runtime component that is activated only during a merge
-(happens in a background thread). **There is no additional overhead
-associated with a user action**. When a user submits an action, it is
+(happens in a background thread). **It is important to note that
+Carmot does not introduce any additional overhead on a user action to
+guarantee convergence**. When a user submits an action, it is
 immediately executed and committed to the local version. Thus, insofar
@@
 
-It is possible to give theoretical upper bound on staleness in
-Carmot. Ring topology guarantees the uniqueness of LCAs (Lemma 3.4) by
-construction (without a need for synchronization). If replicas are
-organized as a ring, the maximum staleness is the sum of latencies on
-the circumference of the ring. Hence if we have replicas distributed
-across US-east and US-west, it is reasonable to expect staleness to be
-in the order of 100ms. As it stands now, our implementation simply
-reflects our formal model, which includes no such optimizations.
-Lastly, we'd like to note that mere presence of synchronization doesn't
-necessarily lead to unacceptable performance. Google docs, for
-instance, relies on a centralized OT server to coordinate all its
-edits and yet has acceptable performance.
+It is possible to give theoretical upper bound on staleness in Carmot.
+We have proved (elsewhere) that a ring topology guarantees the
+uniqueness of LCAs by construction without a need for global
+synchronization. If replicas are organized as a ring, the maximum
+staleness is the sum of latencies on the circumference of the ring.
+Hence if we have replicas distributed across US-east and US-west, it
+is reasonable to expect staleness to be in the order of 100ms. As it
+stands now, our implementation simply reflects our formal model, which
+includes no such optimizations.  Lastly, we'd like to note that there
+exist real distributed applications that make use of synchronization
+in practice. Google docs, for instance, relies on a centralized OT
+server to coordinate all its edits yet remains usable.
 
@@
   problem, however, is that [13] lacks either a system or a proof that
-  guarantees the convergence of resultant MRDTs. As demonstrated
-  above, divergence is possible despite a sensible merge semantics.
-  Carmot closes this gap.
+  guarantees the convergence of resultant MRDTs. We show that
+  divergence is possible despite a sensible merge semantics, and close
+  this gap.
+
+* Git acknowledges the possibility of multiple LCAs in its
+  [documentation](https://git-scm.com/docs/merge-strategies). The
+  default merge strategy of Git is `recursive`, which recursively
+  merges LCAs if there are multiple of them. They admit that this is
+  merely a heuristic. We have found a pathological case where such
+  recursive merging could be a problem (Full example with
+  illustrations included in the appendix). We haven't contacted the
+  Git maintainers yet.
 
@@
 * Inclusion of a runtime imposes no additional overhead on the latency
-  of individual operations. Given that Carmot runtime uses
-  synchronization for merges, and the overhead of synchronization
-  increases at least linearly with the number of replicas, one would
-  expect the latency of each operation to grow at least linearly if it
-  is affected by the presence of runtime. However, as Fig.10
-  demonstrates, that's not the case. Hence the conclusion that our
+  of individual operations. Given that Carmot uses synchronization for
+  merges, and the overhead of synchronization increases at least
+  linearly with the number of replicas, one would expect the latency
+  of each operation to grow at least linearly *if* the runtime
+  interferes with the normal execution. Fig.10 categorically
+  demonstrates that this is not the case. Hence the conclusion that our
   approach imposes no additional cost on latency and availability (our
@@
 
-* We hope your other comments on performance are addressed by our note
-  on Evaluation above.
+* Being based on MRDTs, our approach *is* indeed comparable to CRDTs.
+  The absence of synchronization in CRDTs is a result of careful
+  engineering by domain experts [20, 25, 26, 11]. In the absence of
+  such expertise, the only alternative available to an application
+  developer is to synchronize every operation. Our approach presents a
+  third alternative -- a compromise where developer need not
+  synchronize every operation, yet obtain convergence.
+
+* Our claim in contribution bullet #2 pertains to the formalization of
+  the abstract machine and the theorems we prove on it. Mechanization
+  in Ivy, similar to mechanization in Coq, is intended to convey the
+  confidence in the soundness of our meta-theory. Our Ivy development
+  very closely resembles our formalization in Sections 3 and 4
+  (including the theorems). Lemmas 3.4 to 3.8 have been proved in Ivy.
+  In addition, we also proved that the concrete semantics of Sec. 4
+  refine the abstract semantics of Sec. 3. While Sledgehammer extends
+  an interactive proof assistant (Isabelle/HOL) with SMT support, our
+  mechanized verification in Ivy is wholly SMT-driven; there is no
+  interactive theorem proving involved (we have separately done manual
+  proofs however). We shall include more details on Ivy formalization,
+  and the comparison with Sledgehammer approach in the final version.
+
+* We hope your questions on merge synchronization are answered by the
+  note on Evaluation above.

### 2022-02-09 pldi22_rebuttal.md
PLDI Rebuttal Edits

@@
 We use synchronization to perform the the checks on LCAs, which inturn
-guarantees convergence. Otherwise synchronization in itself doesn't
-guarantee convergence.
+guarantees convergence. Otherwise synchronization in itself is not
+very useful.
+
+Conversely, synchronization is not necessarily an antithesis to
+practical adoption. Google docs, for instance, famously relies on a
+centralized OT server to coordinate all its edits.
 
@@
 stands now, our implementation simply reflects our formal model, which
-includes no such optimizations.  Lastly, we'd like to note that there
-exist real distributed applications that make use of synchronization
-in practice. Google docs, for instance, relies on a centralized OT
-server to coordinate all its edits yet remains usable.
+includes no such optimizations.
+
+Lastly, we'd like to point out that this work introduces staleness as
+a novel tradeoff against convergence. While trading off staleness may
+not be an optimal choice for every application, the fact that there
+exists such a tradeoff is, in our opinion, noteworthy. And if an
+application has to chose between latency and staleness, staleness is
+arguably the less disruptive choice.
 

### 2022-02-09 pldi22_rebuttal.md
PLDI Rebuttal Edits

@@
   merges with help of *relational decomposition*. Albeit
-  computationally inefficient, relational decomposition leads to
-  simple yet useful merges for a range of data structures. The
-  problem, however, is that [13] lacks either a system or a proof that
+  asymptotically sub-optimal, relational decomposition leads to simple
+  yet useful merges for a range of data structures. The problem,
+  however, is that [13] lacks either a system or a proof that
   guarantees the convergence of resultant MRDTs. We show that
@@
   default merge strategy of Git is `recursive`, which recursively
-  merges LCAs if there are multiple of them. They admit that this is
-  merely a heuristic. We have found a pathological case where such
-  recursive merging could be a problem (Full example with
-  illustrations included in the appendix). We haven't contacted the
-  Git maintainers yet.
+  merges LCAs if there are multiple of them. This however merely a
+  heuristic. We have found a pathological case where such recursive
+  merging could become a problem (Full example with illustrations
+  included in the appendix).
+  [Others](http://r6.ca/blog/20110416T204742Z.html) have also found
+  cases where Git merge is inconsistent. We haven't contacted the Git
+  maintainers yet.
 
@@
 
-* We hope your concerns on staleness and performance are addressed by
-  our note on Evaluation above.
+* We hope the reviewer's concerns on staleness and performance are
+  addressed by our note on Evaluation above.
 

### 2022-02-09 pldi22_rebuttal.md
PLDI Rebuttal Edits

@@
    linearize merges *in time*, but nonethless results in ill-formed
-   histories that violate property (c). Fig e.g., in Fig.5(b) merges
+   histories that violate property (c). For e.g., in Fig.5(b) merges
    could be ordered linearly in time, yet result in divergence.
@@
 with a collection of RDTs as it lets us examine the tradeoffs more
-closely. All the MRDT benchmarks from Kaki et al continue to work on
-Carmot with no increase in latency and diff size measurements. Carmot
-is in fact a fork of Quark -- the system from Kaki et al. Carmot
-extends Quark with an more efficient network and storage layer (due to
-Scylla), and a runtime component that is activated only during a merge
-(happens in a background thread). **It is important to note that
-Carmot does not introduce any additional overhead on a user action to
-guarantee convergence**. When a user submits an action, it is
-immediately executed and committed to the local version. Thus, insofar
-as user interaction is concerned, Carmot is qualitatively no different
-from any other CRDT system. Unfortunately, we couldn't find a
-standardized CRDT system (akin to Quark for MRDTs) for us to do
+closely. All the MRDT benchmarks from Kaki et al [13] continue to work
+on Carmot with no increase in latency and diff size measurements.
+Carmot is in fact a fork of Quark -- the system from Kaki et al.
+Carmot extends Quark with an more efficient network and storage layer
+(due to Scylla), and a runtime component that is activated only during
+a merge (happens in a background thread). **It is important to note
+that Carmot does not introduce any additional overhead on a user
+action to guarantee convergence**. When a user submits an action, it
+is immediately executed and committed to the local version. Thus,
+insofar as user interaction is concerned, Carmot is qualitatively no
+different from any other CRDT system. Unfortunately, we couldn't find
+a standardized CRDT system (akin to Quark for MRDTs) for us to do
 comparisons against.

### 2022-03-15 abstract-semantics.tex
Incorporating reviewer comments

@@
 sequence of versions, which intuitively denotes the progressive
-evolution of the state on a replica. The latest version of the branch,
-called its \emph{head}, denotes the current state of the corresponding
-replica. A new head version is created either by an
+evolution of the state on a replica. There is thus a one-to-one
+mapping between replicas and branches. The latest version of the
+branch, called its \emph{head}, denotes the current state of the
+corresponding replica. A new head version is created either by an
 externally-initiated \emph{commit} (\rulelabel{Commit}) or by merging
@@
 
-The state of \quark abstract machine is a tuple $\Delta = (G,N,C,H,L)$, where:
+The state of \quark abstract machine is a tuple $\Delta \;=\; $ 
+$(G,N,C,H,L)$, where:
 \begin{itemize}
@@
 \begin{definition}[Initial State and Version History Graph]
+\label{def:init-state}
   The graph $G_{\odot} = (\{v_{\odot}\},\emptyset)$ is the initial
@@
 (e.g., \C{add($e$)}) on the value of the current version to create the
-new value $n$, and thus the new version $v$. A unique commit id $c$ is
+new value $n$, and thus the new version $v$. A unique commit id $i$ is
 assigned to this commit event and added to $C(v)$. The set $C(v)$
@@
 \paragraph{Properties} We now formalize the notable properties of the
-abstract machine and its executions\footnote{
-  The manual proofs of the theorems and their Ivy
-  formalization~\cite{ivy} can be found in the supplementary material.
-}.
+abstract machine and its executions.
 
@@
   \label{thm:progress}
-  Every reachable state $\Delta = ((V,E),N,C,H,L)$ of the abstract
+  Every reachable state $\Delta = ((V,E),\\N,C,H,L)$ of the abstract
   machine is either: 
@@
 abstract machine.
+
+\noindent\paragraph{Ivy formalization} Appendix contains the manual
+proofs of the aforementioned theorems. In addition we also formalized
+the \quark abstract machine in Ivy~\cite{ivy}, and verified its
+properties automatically with help of Z3~\cite{z3}. Ivy is a
+multi-modal verification tool, whose primary purpose is to verify the
+design and implementations of distributed protocols. The key
+differentiating aspect of Ivy is its onus on predictable automation
+and decidable reasoning.  Automated proofs in Ivy are restricted to
+logical fragments for which the tool is a decision procedure. %Towards
+% this end, Ivy checks the verification conditions for syntactic
+% confirmity with known decidable fragments, and provides feedback to
+% the user if the checks fail. The feedback helps users reorganize their
+% formal development or reformulate their queries until the reasoning
+% becomes decidable. Ivy uses induction to prove the properties of
+% infinite-state systems. The failure to discharge a proof results in a
+% counterexample, which is either a real counterexample (i.e., a bug in
+% the system) or a counterexample to induction. In the latter case, the
+% counterexample serves as a guide for the user to formulate the
+% inductive invariants required to discharge the proof. 
+Consequently, the creative aspect of proving a theorem in Ivy lies in
+discovering decidable abstractions that characterize the system being
+reasoned about, and crafting the inductive invariants required to
+discharge the automated proof. This is in contrast to theorem provers
+such as Coq and Isabelle, and their extensions such as
+Sledgehammer~\cite{sledge}, where automation is used on a best-effort
+basis to aid interactive theorem proving.
+
+Our formalization of the \quark abstract machine in Ivy follows the
+formal development in Fig.~\ref{fig:git-semantics} but uses slightly
+different abstractions to obtain decidable reasoning. Maps $C$, $N$,
+$H$, and $L$ are modeled as relations instead of functions since
+reasoning about quantifier-bound function application is in general
+undecidable. $V$ and $E$ are relations by definition. Their initial
+values are as specified by $\Delta_{\odot}$ in
+Def.~\ref{def:init-state}. The system admits four \emph{actions}
+corresponding to the four transition rules in
+Fig.~\ref{fig:git-semantics}. Each action ``updates'' the system state
+$\Delta = ((V,E),N,C,H,L)$ as defined by the corresponding rule in
+Fig.~\ref{fig:git-semantics}.
+Commit, version, and branch identifiers are modeled as totally ordered
+sets of uninterpreted values from which new values can be chosen
+sequentially. This lets us capture these identifiers as monotonically
+increasing integers while avoiding undecidable reasoning with
+quantifier-bound integer arithmetic.
+Lemmas/Theorems~\ref{lem:lca-uniqueness} to~\ref{thm:progress} are
+asserted as inductive properties of the \quark abstract machine. Ivy
+succesfully verifies each of these properties with help of Z3, thus
+certifying the soundness of \quark's meta-theory.

### 2022-03-15 concrete-semantics.tex
Incorporating reviewer comments

@@
 expands to $H_i[i \mapsto b_i \mapsto t]$. We exploit this notation in
-Fig.~\ref{fig:quark-semantics}. 
+Fig.~\ref{fig:quark-semantics}. We let $\delta_{\odot}$ denote the
+initial concrete state which is defined on the similar lines as the
+initial abstract state (Def.~\ref{def:init-state}).
 % When two replicas, $i$ and $j$, are involved in an update, e.g.,
@@
 
-\paragraph{Properties} The Unique LCA
-(Lemma~\ref{lem:lca-uniqueness}), Convergence
-(Theorem~\ref{thm:convergence}, and Progress
-(Theorem~\ref{thm:progress}) properties have been re-defined for the
-\quark distributed machine and automatically proven correct in
-Ivy~\cite{ivy}.
+\paragraph{Refinement} The following theorem asserts that the \quark
+distributed machine refines the abstract machine described in
+Sec.~\ref{sec:abstract-sem}:
+\begin{theorem}[{\bf Refinement}]
+  \label{thm:refinement}
+  There exists a refinement relation $R$ relating abstract and
+  concrete states such that:
+  \begin{itemize}
+    \item $R$ relates initial abstract and concrete states, i.e.,
+      $R(\Delta_{\odot},\delta_{\odot})$.
+    \item Forall $\delta,\delta',\Delta$, if $R(\delta,\Delta)$ and
+      $\delta \qstepsto \delta'$, then there exists a $\Delta'$ such
+      that $\Delta \stepsto \Delta'$ and $R(\delta',\Delta')$.
+  \end{itemize}
+\end{theorem}
+The proof is done in Ivy. The refinement relation $R$ essentially
+relates the version vectors of concrete semantics to the version
+histories of abstract semantics. A \emph{happens-before} relation has
+been defined as a ghost state in concrete semantics to help us
+establish the refinement relation in Ivy.
 

### 2022-03-15 evaluation.tex
Incorporating reviewer comments

@@
 
-We present an evaluation study of our runtime-assisted approach to
-convergent RDTs with help of our prototype \quark implementation. The
-benchmark we chose for the study is a collaborative document editing
+In this section we present an emperical evaluation of \quark on two
+case studies. First is a collaborative document editing
 application -- a common usecase addressed by several CRDT
-proposals~\cite{rga, treedoc, crdts}. However, unlike these
-other approaches, we did not have to build a dedicated replicated data
-type to represent collaboratively-edited documents; an ordinary
+proposals~\cite{rga, treedoc, crdts}. Second is a replicated key-value
+store implemented using a mergeable Red-Black tree data structure.
+
+\subsection{Collaborative Editing}
+
+\quark's MRDT approach obviates the need to build a dedicated
+replicated data type for collaboratively-edited documents; an ordinary
 document format extended with a merge operation would suffice. While
@@
 type to merge documents. \C{List.merge} is a simple implementation of
-the GNU \C{diff3} algorithm~\cite{gnudiff} in ~60 lines of OCaml. We
-thus adopt a straightforward approach to building a collaborative
-document editor with the intention to keep the development effort low
-enough to be easily replicated. The convergence guarantee of \quark
-ensures that the simplicity of our implementation doesn't come at the
-expense of correctness. We now describe an evaluation that
-demonstrates that it also doesn't come at the expense of the editor
-performance.
+list merge algorithm (in ~60 lines of OCaml) inspired by the GNU
+\C{diff3} algorithm~\cite{gnudiff}. We thus adopt a straightforward
+approach to building a collaborative document editor with the
+intention to keep the development effort low enough to be easily
+replicated. The convergence guarantee of \quark ensures that the
+simplicity of our implementation doesn't come at the expense of
+correctness. The aim of the experimental evaluation is to quantify the
+impact of \quark on the performance.
 
@@
 latest version on the local branch to create a new version that
-includes the updates from either direction. 
+includes the updates from either direction. Since this merge is fully
+confined to the local replica, it is guaranteed to not affect the LCA
+of the local branch with respect to a remote branch. To the external
+world, it appears as if the local branch has simply committed a new
+version.
 % The LCA for this merge is
@@
 
+\noindent\paragraph{Latency} To measure the impact of \quark runtime on user
+writes, we measure the latency of the \C{write} operation, which
+includes the time spent merging the user version with the current
+version, and persisting the resultant version to the store. We conduct
+the experiments on a three-node cluster of \C{i3.large} machines in
+Amazon \C{us-west2} data center. Each user connects to one of the
+machines, forks a new branch, and performs 1000 edits in succession,
+saving the document after each edit. We progressively increase the
+number of concurrent users editing the document from 3 to 200 and
+measure the impact of the increased concurrency on write latency.
+Fig.~\ref{fig:latency} shows the 10th, 50th (median), and 90th
+percentile latency values. The median and 10th percentile latencies
+remain more-or-less constant with a slight increase between 3 and 60
+concurrent users. This demonstrates that \quark has negligible effect
+on the per-operation latency, which is expected. The 90th percentile
+latency, however, shows no clear pattern by the virtue of being more
+susceptible to transient system conditions. Nonetheless, the maximum
+value of 90th percentile latency measured (32ms) remains well below
+the the time between consecutive edits (0.25s), making it hard to
+perceive by a human user.
+
 \begin{figure}[ht]
   \centering
-    \includegraphics[scale=0.3]{Figures/latency}
-\caption{Latency of collaborative editing operations under \quark vs
-  executing them with strong consistency (SC). }
+    \includegraphics[scale=0.4]{Figures/monkey_latency}
+\caption{Latency of writes to a shared document under \quark. }
 \label{fig:latency}
@@
 
-To measure the impact of \quark runtime on user writes, we measure the
-latency of the \C{write} operation, which includes for the time spent
-merging the user version with the current version, and persisting the
-resultant version to the store. We conduct the experiments on a
-three-node cluster of \C{i3.large} machines in Amazon \C{us-west2}
-data center. Each user connects to one of the machines and performs
-1000 edits in succession, saving the document after each edit. We
-progressively increase the number of concurrent users editing the
-document from 3 to 60 and measure the impact of the increased
-concurrency on write latency. Fig.~\ref{fig:latency} shows the median
-latency values. As evident from the figure, latency of \quark writes
-remain more-or-less constant with the median latency around 0.008s.
-The maximum latency ever measured for a \quark writes is 0.016s.
-Notably, latency remains an order of magnitude less than the time
-between consecutive edits (0.25s), making it hard to perceive.
-Fig.~\ref{fig:latency} also plots the latency values for the baseline
-``SC'' approach which achieves convergence by synchronizing each
-operation, i.e., executing under strong consistency (SC). The SC write
-latency increases super-linearly with the median value of 1.76s for 60
-concurrent editors. The maximum latency measured for an SC write is
-9.02s, which is considerably more than the inter-edit latency of
-0.25s. The experiment confirms the theory that convergence does not
-come at the cost of latency in
-\quark.
+For comparison against a baseline, we have implemented an ``SC''
+approach which achieves convergence by synchronizing each operation,
+i.e., executing it under strong consistency (SC). The SC
+implementation shares most of its code with \quark with the only
+change being that it wraps commits instead of merges inside a lock.
+As with \quark, we measured the write latency of the SC implementation
+while increasing the number of concurrent users ($n$).  The median SC
+write latency increases super-linearly from 10ms for $n=3$ to 63s
+(i.e., >1m) for $n=60$, which is considerably more than the inter-edit
+latency of 0.25s.
 
-\begin{figure}[ht]
-  \centering
-    \includegraphics[scale=0.3]{Figures/staleness}
-\caption{Staleness increases as the number of concurrent editors
-  increase.}
-\label{fig:staleness}
-  \vspace*{-0.2in}
-\end{figure}
-
-\quark system replicates the contents of each branch across all the
-replicas as fast as the network allows. However, for a user $A$ to
-see the changes made by the other user $B$, the changes have to be
+\noindent\paragraph{Staleness} Our implementation of \quark relies on
+Scylla to replicate the contents of each branch across all the
+replicas as fast as the network allows. However, for a user $A$ to see
+the changes made by the other user $B$, the changes have to be
 reflected in $A$'s local version, which can only happen through a
@@
 convergence (but doesn't preempt it, as proved by
-Theorem~\ref{thm:progress}). To understand the staleness behavior in
-\quark, we repeat the collaborative editing experiment, this time
-measuring the staleness value at every merge. We do this by annotating
-every version $v$ with the timestamp $t$ of its creation time. When
-$v$ is is merged into a remote branch $b$ at a later time $t'$, the
-difference $t' - t$ denotes the staleness of $v$ w.r.t the new version
-on $b$. Multiple such staleness measurements are recorded for each
-experiment to compute the 10th, 50th, and 90th percentile
-values. Fig.~\ref{fig:staleness} shows the results. While
-staleness remains in the order of milliseconds with fewer ($\leq 9$)
-concurrent users, it increases super-linearly as we increase the
-number of collaborators in the multiples of 3 until 30. While
-increased staleness is inevitable in our approach due to synchronized
-merges, we believe the increase can be contained by choosing the order
-of merges to avoid the ``starvation'' of some branches. Note that
-relaxing the linearizability constraint on merges would completely
-eliminate the staleness overhead, but the resultant Git-like system
-fails to converge due to anomalous executions described in
-Sec.~\ref{sec:motivation}\footnote{
-  Git admits anamolous version history graphs where two branches can
-  have the same set of commits and yet differ in their final version.
-  Supplementary material describes two such cases we observed on
-  Github.}.
+Theorem~\ref{thm:progress}).
+
+\begin{figure}[ht]
+  \centering
+    \includegraphics[scale=0.4]{Figures/monkey_staleness}
+\caption{Staleness increases as the number of concurrent editors
+  increase.}
+\label{fig:monkey-staleness}
+  \vspace*{-0.2in}
+\end{figure}
+
+To understand the effect of \quark on
+staleness, we quantify and measure it along with latency in the
+experiment setup described above. Staleness is defined as the time
+taken for a version committed on one replica to be merged into a
+concurrent version on a remote replica. To measure staleness, we
+annotate every version $v$ with the timestamp $t$ of capturing the
+wall clock time of its creation. When $v$ is is merged into a remote
+branch $b$ at a later time $t'$, the difference $t' - t$ denotes the
+staleness of $v$ w.r.t the new version on $b$. One such staleness
+measurement is recorded for every merge that ever happens during the
+experiment. We compute 10th, 50th, and 90th percentiles of staleness
+values thus obtained. Fig.~\ref{fig:monkey-staleness} shows the
+results. As evident, staleness increases steadily with the increasing
+number of concurrent replicas, which is expected considering that
+merges are synchronized, and increasing the number of replicas reduces
+the number of opportunities for a replica to merge. The increase is
+roughly linear in the number of replicas as our implementation passes
+the lock around in a round-robin fashion, thus increasing the lock
+latency in proportion to the number of replicas. Nonetheless, the 90th
+percentile staleness values remain low -- in the order of 100ms with
+the number of replicas under 30, and in the order of 1s with number of
+replicas under 60. While further optimizations might reduce staleness,
+a non-trivial staleness overhead is inevitable in \quark due to our
+use of synchronized merges to guarantee convergence.
+
+% Our approach thus
+% introduces staleness as novel tradeoff against convergence. While
+% trading off staleness may not be an optimal choice for every
+% application, it is arguably a less disruptive choice if an application
+% has to choose between latency and staleness.
+% Sec.~\ref{sec:motivation}\footnote{
+%   Git admits anamolous version history graphs where two branches can
+%   have the same set of commits and yet differ in their final version.
+%   Supplementary material describes two such cases we observed on
+%   Github.}.
+
+\subsection{Key-Value Store}
+
+We repeat our experiments on a mergeable key-value store application
+we implemented using a Red-Black tree MRDT. Latency results are shown
+in Fig.~\ref{fig:rb-latency}, and staleness results in
+Fig.~\ref{fig:rb-staleness}.
+
+\begin{figure}[ht]
+  \centering
+    \includegraphics[scale=0.4]{Figures/rbmonkey_latency}
+  \caption{Latency for Key-Value store operations under \quark}
+\label{fig:rb-latency}
+  \vspace*{-0.2in}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+    \includegraphics[scale=0.4]{Figures/rbmonkey_staleness}
+  \caption{Key-Value store staleness under \quark}
+\label{fig:rb-staleness}
+  \vspace*{-0.2in}
+\end{figure}
 

### 2022-03-15 implementation.tex
Incorporating reviewer comments

@@
 
-We realize a prototype of \quark runtime as a lightweight shim layer
-on top of Scylla -- an off-the-shelf distributed data
+We realize a prototype of \quark runtime for MRDTs as a lightweight
+shim layer on top of Scylla -- an off-the-shelf distributed data
 store~\cite{scylla}. We rely on Scylla for inter-replica
 communication, data replication, persistence, and fault tolerance.
-\quark translates the high-level mergeable data type implementations
-in OCaml to their low-level representations in the backing store and
-organizes their well-formed distributed executions.
+\quark translates the high-level MRDT implementations in OCaml to
+their low-level representations in the backing store and orchestrates
+their well-formed distributed executions.
 Fig.~\ref{fig:implementation} illustrates the overall architecture.
 
-The implementation of \quark the is a straightforward realization of
-the \quark distributed machine described in Sec.~\ref{sec:concrete-sem}.
+The implementation of \quark is a straightforward realization of the
+\quark distributed machine described in Sec.~\ref{sec:concrete-sem}.
 We manifest each component of the state, namely the branch map $B$,
@@
 enforced with help of \C{Quorum} reads and writes. Each user process
-is assigned its own replica of the RDT.  Version vectors are realized
-as associative lists and stored in Scylla
-as blobs.
+is assigned its own replica of the MRDT.  Version vectors are realized
+as associative lists and stored in Scylla as blobs.
 
 One key difference between the formalization and the implementation is
-in the treatment of RDT values. Formalization assumes values to be
-atomic with no sharing in between them. In practice, however, an RDT
-value could be a linked data structure such as a binary tree, and two
+in the treatment of MRDT values. Formalization assumes values to be
+atomic with no sharing in between them. In practice, however, an MRDT
+could be a linked data structure such as a binary tree, and two
 such values could share a significant amount of internal structure.

### 2022-03-15 introduction.tex
Committing last minute edits.
Incorporating reviewer comments

--- 2021-11-20 introduction.tex
@@
 
-% hinders the widespread adoption of replicated data
-% types in software development.
-In this paper we propose such an alternative approach to convergent
-replicated data types. The key distinguishing characteristic of our
-approach is the utilization of a distributed runtime to orchestrate
-\emph{well-formed} distributed executions that are guaranteed to
-converge. We define well-formedness as a constraint over the
-\emph{states} that arise during the execution as opposed to the
-\emph{operations} that are involved in the execution. To allow for its
-enforcement, we adopt a state-centric model of replication based on
-version-controlled \emph{mergeable} states~\cite{mrdt} instead of an
-operation-centric model based on commutative operations. Unlike
-commutativity, mergeability does not require a data type definition to
-be refactored to suit distributed execution. Instead, the type
-definition only needs to be extended with a \C{merge} function to
-reconcile concurrent versions of the type in presence of their common
-ancestor version. Notably, \C{merge} function is left completely
-unconstrained. Consequently, programmer can chose any merge semantics
-that make sense within the context of the data type's (sequential)
-semantics without risking the loss of convergence. Such strong
-guarantees are made possible by the simplicity of state-based
-reasoning combined with the run-time enforcement of well-formedness.
-
+% In this paper we propose such an alternative approach to convergent
+% replicated data types. The key distinguishing characteristic of our
+% approach is the utilization of a distributed runtime to orchestrate
+% \emph{well-formed} distributed executions that are guaranteed to
+% converge. We define well-formedness as a constraint over the
+% \emph{states} that arise during the execution as opposed to the
+% \emph{operations} that are involved in the execution. To allow for its
+% enforcement, we adopt a state-centric model of replication based on
+% version-controlled \emph{mergeable} states~\cite{mrdt} instead of an
+% operation-centric model based on commutative operations. Unlike
+% commutativity, mergeability does not require a data type definition to
+% be refactored to suit distributed execution. 
+In this paper we describe an alternative approach to building
+convergent replicated data types that realizes the aforementioned
+virtues. Our approach is based on Mergeable Replicated Data Types
+(MRDTs)~\cite{mrdt} -- an alternative take on RDTs that is inspired by
+the Git version control system. MRDTs adopt a state-centric model of
+replication based on version-controlled \emph{mergeable} states
+instead of an operation-centric model based on commutative operations.
+Unlike commutativity, mergeability does not require a data type
+definition to be refactored to suit distributed execution. Instead,
+the type definition only needs to be extended with a \C{merge}
+function to reconcile concurrent versions of the type in presence of
+their common ancestor version. However, mergeability itself doesn't
+guarantee convergence; there exist MRDTs with well-defined and
+intuitive merge semantics that nonetheless admit divergent executions.
+While constraining the data type semantics to conform to algebraic
+laws such as monotonicity might restore convergence, it would lead
+MRDTs down the path of CRDTs, which has known problems described
+above. We therefore propose an alternative approach to convergence --
+one that is based on runtime orchestration rather than static
+enforcement. In particular, we extend MRDTs with a distributed runtime
+that orchestrates only the \emph{well-formed} executions which are
+guaranteed to converge. Notably, our approach guarantees convergence
+of MRDTs \emph{regardless} of their merge semantics, thus obviating
+the need to constrain the type semantics or their implementations. The
+key insight behind our approach is the observation that a data type's
+\C{merge} is a complete function, and thefore can be safely invoked to
+merge concurrent versions of the type notwithstanding their individual
+operation histories. Thus, a replica's current version
+can always be made available for user operations with the guarantee
+that it can be safely merged with a concurrent version at a later
+point of time. The ability to arbitrarily delay merges gives us the
+opportunity to orchestrate them in a way that is \emph{effectively}
+linear, which inturn guarantees convergence.
 % combined with the simplicity
@@
 execution is correspondingly low. It is therefore important for a
-distributed runtime of RDTs to not synchronize the execution of RDT
-operations. Orchestrating a well-formed convergent execution without
-impacting latency and availability may seem impossible due to the CAP
-theorem~\cite{cap}, but it is not the case in fact. While CAP theorem
-does rule out linearizability, it does not prohibit enforcing weaker
-consistency constraints on distributed executions. Our approach
+distributed runtime of RDTs to \emph{not} interfere with the execution
+of RDT operations. Orchestrating a well-formed convergent execution
+without impacting latency and availability may seem impossible due to
+the CAP theorem~\cite{cap}, but it is not the case in fact. While CAP
+theorem does rule out linearizability, it does not prohibit enforcing
+weaker consistency constraints on distributed executions. Our approach
 exploits this observation by (\rom{1}).  Identifying a novel set of
@@
 (asynchronously executed) state merges such that per-operation latency
-and system-wide availability remain unaffected.
-Sec.~\ref{sec:concrete-sem} explains our runtime in detail.
+and system-wide availability remain unaffected. The downside of
+interfering with merges, however, is that the replicas may now take
+longer to converge -- a phenomenon we quantify using a metric called
+\emph{staleness} (Sec.~\ref{sec:eval}). Our approach thus brings to
+fore a fundamental tradeoff that was never explored in the context of
+RDTs before.
 
-% In fact many CRDTs,
-% including the aforementioned \emph{remove-wins} set, implicitly assume
-% \emph{causal consistency}, which is weaker than linearizability but is
-% stronger than the default \emph{eventual consistency}. While causal
-% consistency alone cannot guarantee convergence, it sufficiently
-% constrains the distributed execution for the CRDT implementations to
-% guarantee convergence. In this respect our approach differs in the
-% sense that
+%Sec.~\ref{sec:motivation} explains these insights in detail.
 
@@
 \begin{itemize}
-  \item We present a runtime-assisted approach to state replication
-    that sufficiently constrains distributed executions so as to
-    guarantee the convergence of replicated data types. This is in
-    contrast to the existing models of replication, where each type is
-    obligated to prove its convergence.
+  \item We present a runtime-assisted approach to MRDT state
+    replication that sufficiently constrains distributed executions so
+    as to guarantee their convergence. This is in contrast to the
+    existing models of replication, where each data type is obligated
+    to prove its convergence by discharging algebraic proof
+    obligations.
 
   \item We formalize the aforementioned constraints in the context of
-    a an abstract machine (named \quark) that manifests distributed
-    executions as version history graphs (similar to Git).
-    We prove the \emph{convergence} and \emph{progress} properties of
-    the \quark abstract machine that together guarantee its soundness.
-    The formalization has been mechanized in Ivy~\cite{ivy} and
-    \emph{automatically} verified with help of Z3~\cite{z3}. To the
-    best of our knowledge, this is the first time an SMT solver was
-    used to verify the \emph{meta-theory} of a formal system.
+    an abstract machine (named \quark) that manifests distributed
+    executions as version history graphs similar to Git.  We prove the
+    \emph{convergence} and \emph{progress} properties of the \quark
+    abstract machine that together guarantee its soundness. The
+    formalization has also been mechanized in Ivy~\cite{ivy} and
+    automatically verified with help of Z3~\cite{z3}. To the best of
+    our knowledge, this is the first time the meta-theory of a formal
+    is proven completely inside an SMT solver.
 
@@
     layer on top of Scylla -- an off-the-shelf distributed key-value
-    store~\cite{scylla}. We evaluate the implementation on a
-    collaborative-editing benchmark to demonstrate novel trade-offs
-    presented by our approach.
+    store~\cite{scylla}. We describe an extensive evaluation to
+    quantify the staleness metric and novel tradeoffs in the context
+    of case studies involving a data structure benchmark and a
+    collaborative editing application. 
 \end{itemize}

### 2022-03-15 motivation.tex
Incorporating reviewer comments

@@
 
-\noindent\paragraph{Key observations guiding the solution}
-Fortunately, there is a way out of this problem and the solution
-requires a switch in the perspective of replication from an
-operation-centric view characterized by explicit effects to a
-state-centric view characterized by explicit state merges. The
-motivation for this switch is the key observation that it is possible
-to induce an ordering over \emph{states} of replicas even as their
-operations and effects remain unordered, as long as such states are
-\emph{mergeable}. The ordering is enforced by synchronizing merges in
-the background while the replica-local execution progresses
-unhindered. Consequently, eventual convergence can be guaranteed
-\emph{without} impacting the user-perceived latency of the operations.
-Moreover, working with states instead of operations frees the latter
-from having to conform to algebraic laws such as commutativity.  Put
-together, this means that the clients of a replicated data type can
-perform any action allowed by the sequential version of the type and
-immediately see the results of their actions on the local replica
-while being safe in the knowledge that replica states will eventually
-converge.
-
 \begin{figure}[ht]
@@
 
-\noindent\paragraph{A Mergeable Replicated Set} To demonstrate the
-aforementioned intuitions we reconsider the executions from
-Fig.~\ref{fig:crdt-execs}, this time in the state-centric replication
-model. The cornerstone of the state-centric model is a three-way
-\C{merge} function that merges concurrent versions of the state in
-presence of their (lowest) common ancestor version. In our running
-example, the state is a value of type \C{Set.t}, hence \C{Set.merge}
-function would have the type signature:
+\noindent\paragraph{The MRDT Approach} Mergeable Replicated Data Types
+(MRDTs) implement a state-centric model of replication that is
+inspired by the Git version control system. Like in Git, the state
+evolves in terms of versions, and concurrent versions of the state can
+be merged. The semantics of merge depends on the type of the state, so
+each MRDT is required to be equipped with a three-way \C{merge}
+function that merges concurrent versions of that type in presence of
+their (lowest) common ancestor version. In our running example, the
+state is a value of type \C{Set.t}, hence \C{Set.merge} function would
+have the type signature:
 \begin{center}
@@
 \end{ocaml}
-Equipped with the above set merge function, we can now consider the
-equivalent execution of Fig.~\ref{fig:crdt-exec-1} in state-centric
-model. The execution is shown in Fig.~\ref{fig:mrdt-exec-1}. The
-initial version on all three replicas ($v_0$) is the singleton set
+Extending \C{Set} ADT with the above merge function results in a
+\C{Set} MRDT.
+
+Let us now reconsider the executions in Fig.~\ref{fig:crdt-execs},
+this time on \C{Set} MRDT. The equivalent executions are shown in
+Fig.~\ref{fig:mrdt-execs}. Due to causal consistency being built into
+the model, the divergent execution of Fig.~\ref{fig:crdt-exec-1} is
+preempted in favor of the convergent execution shown in
+Fig.~\ref{fig:mrdt-exec-1}. The execution manifests as follows.  The
+initial version ($v_0$) on all three replicas  is the singleton set
 $\{e\}$. Applying operations to replicas creates new versions, e.g.,
-$v_1$ on $R_3$, and $v_2$ and $v_3$ on $R_2$. Changes can be propagated by merging versions, e.g.,
-version $v_2$ on $R_2$ is a result of merging $v_1$ into $v_0$ for
-which $v_0$ serves as the lowest common ancestor (LCA)
+$v_1$ on $R_3$, and $v_2$ and $v_3$ on $R_2$. Changes can be
+propagated by merging versions, e.g., version $v_2$ on $R_2$ is a
+result of merging $v_1$ into $v_0$ for which $v_0$ serves as the
+lowest common ancestor (LCA)
   \footnote{Versions $v_0$ and $v_1$ are not concurrent as the
@@
 both $R_1$ and $R_2$'s \C{remove}s remove the same element $e$, which
-is not possible in a linearized execution. Leaving the execution
-unconstrained is crucial to satisfy the limits imposed by the CAP
-theorem on a highly-available partition-tolerant distributed system.
+is not possible in a linearized execution. %Leaving the execution
+%unconstrained is crucial to satisfy the limits imposed by the CAP
+%theorem on a highly-available partition-tolerant distributed system.
 
@@
 merges collectively propagate the effects of \C{add} and \C{remove}
-operations to $R_1$ and $R_2$. And despite being ordered in time, the
+operations to $R_1$ and $R_2$. And despite being ordered in time,
 they nonetheless result in a divergent execution ($v_5 \neq v_7$). The
 problem here is that, although merges are executed linearly, the
-execution graph does not reflect this linearity; merges of that end in
+execution graph does not reflect this linearity; merges that end in
 $R_1$ in Fig.~\ref{fig:mrdt-exec-4} are effectively concurrent with
 those that end in $R_4$. This shows that simply synchronizing the
-execution of merges does not necessarily result in convergence.
+execution of merges does \emph{not} result in convergence.
 
-Our key insight to overcome this impasse is a \emph{well-formedness}
+Our key insight to overcome this impasse is a novel \emph{well-formedness}
 condition on execution graphs that ensures convergence of final
@@
 against the good execution in Fig.~\ref{fig:mrdt-exec-3}. Observe that
-in Fig.~\ref{fig:mrdt-exec-2}, \emph{every} pair of concurrent
+in Fig.~\ref{fig:mrdt-exec-3}, \emph{every} pair of concurrent
 versions on $R_1$ and $R_2$ have a unique lowest common ancestor
-(LCA). For instance, LCA of $(v_5,v_6)$ is $v_6$, $(v_5,v_4)$ is
+(LCA). For instance, LCA of $(v_5,v_6)$ is $v_5$, $(v_5,v_4)$ is
 $v_3$, $(v_1,v_2)$ is $v_0$ and so on. By contrast in
@@
 that are also common ancestors of $v_5$ and $v_6$. Likewise in
-Fig.~\ref{fig:mrdt-exec-4}, versions $v_7$ and $v_9$ have two LCAs --
-$v_2$ and $v_4$. Multiple LCAs is an indication that there exist
+Fig.~\ref{fig:mrdt-exec-4}, versions $v_6$ and $v_8$ have two LCAs --
+$v_1$ and $v_3$. Multiple LCAs is an indication that there exist
 merges prior in the execution graph that are effectively concurrent.

### 2022-03-15 quark.tex
Edits
Incorporating reviewer comments

@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[sigplan,10pt,review,anonymous]{acmart}
-\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigplan,screen]{acmart}
+%\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% For final camera-ready submission, w/ required CCS and ACM Reference
@@
 %% use defaults for review submission.
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
-\acmArticle{1}
-\acmYear{2022}
-\acmMonth{1}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
+%%\acmJournal{PACMPL}
+%%\acmVolume{1}
+%%\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
+%%\acmArticle{1}
+%%\acmYear{2022}
+%%\acmMonth{1}
+%%\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
+%%\startPage{1}
+\acmConference[PLDI]{ACM Conference}{June 2022}{San Diego, USA}%
 
@@
 %% Author with single affiliation.
-\author{First1 Last1}
-\authornote{with author1 note}          %% \authornote is optional;
+\author{Gowtham Kaki}
+%\authornote{with author1 note}          %% \authornote is optional;
                                         %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
 \affiliation{
-  \position{Position1}
-  \department{Department1}              %% \department is recommended
-  \institution{Institution1}            %% \institution is required
-  \streetaddress{Street1 Address1}
-  \city{City1}
-  \state{State1}
-  \postcode{Post-Code1}
-  \country{Country1}                    %% \country is recommended
+%  \position{Position1}
+%  \department{Department1}              %% \department is recommended
+  \institution{University of Colorado Boulder}            %% \institution is required
+%  \streetaddress{Street1 Address1}
+%  \city{City1}
+%  \state{State1}
+%  \postcode{Post-Code1}
+  \country{USA}                    %% \country is recommended
 }
-\email{first1.last1@inst1.edu}          %% \email is recommended
+%\email{first1.last1@inst1.edu}          %% \email is recommended
 
-%% Author with two affiliations and emails.
-\author{First2 Last2}
-\authornote{with author2 note}          %% \authornote is optional;
+%% Second Author
+\author{Prasanth Prahladan}
+%\authornote{with author2 note}          %% \authornote is optional;
                                         %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
 \affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}                   %% \country is recommended
+%  \position{Position2a}
+%  \department{Department2a}             %% \department is recommended
+  \institution{University of Colorado Boulder}           %% \institution is required
+%  \streetaddress{Street2a Address2a}
+%  \city{City2a}
+%  \state{State2a}
+%  \postcode{Post-Code2a}
+  \country{USA}                   %% \country is recommended
 }
-\email{first2.last2@inst2a.com}         %% \email is recommended
+%\email{first2.last2@inst2a.com}         %% \email is recommended
+
+%% Third Author
+\author{Nicholas Lewchenko}
+%\authornote{with author2 note}          %% \authornote is optional;
+                                        %% can be repeated if necessary
+%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
 \affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}                   %% \country is recommended
+%  \position{Position2a}
+%  \department{Department2a}             %% \department is recommended
+  \institution{University of Colorado Boulder}           %% \institution is required
+%  \streetaddress{Street2a Address2a}
+%  \city{City2a}
+%  \state{State2a}
+%  \postcode{Post-Code2a}
+  \country{USA}                   %% \country is recommended
 }
-\email{first2.last2@inst2b.org}         %% \email is recommended
+%
 
@@
 \begin{abstract}
-
   We propose a runtime-assisted approach to enforce convergence in
@@
   converge. Despite the utilization of a runtime, our approach comes
-  at no additional cost of latency and availability. We implement our
-  approach in a system called \quark and evaluate its performance over
-  a large-scale collaborative editing application. 
+  at no additional cost of latency and availability. Instead, we
+  introduce a novel tradeoff against a metric called \emph{staleness},
+  which roughly corresponds to the time taken for replicas to
+  converge. We implement our approach in a system called \quark and
+  conduct a thorough evaluation of its tradeoffs in the context of two
+  case studies.
 \end{abstract}
@@
 <concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
+<concept_id>10010147.10010919.10010177</concept_id>
+<concept_desc>Computing methodologies~Distributed programming languages</concept_desc>
 <concept_significance>500</concept_significance>
@@
 <concept>
-<concept_id>10003456.10003457.10003521.10003525</concept_id>
-<concept_desc>Social and professional topics~History of programming languages</concept_desc>
+<concept_id>10010520.10010575.10010578</concept_id>
+<concept_desc>Computer systems organization~Availability</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011074.10011099.10011692</concept_id>
+<concept_desc>Software and its engineering~Formal software verification</concept_desc>
 <concept_significance>300</concept_significance>
@@
 
-\ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
+\ccsdesc[500]{Computing methodologies~Distributed programming languages}
+\ccsdesc[500]{Computer systems organization~Availability}
+\ccsdesc[300]{Software and its engineering~Formal software verification}
 %% End of generated code
@@
 %% comma separated list
-\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+\keywords{Replication, MRDT, CRDT, Runtime, Convergence, Collaborative
+Editing}  %% \keywords are mandatory in final camera-ready submission
 
@@
 %% Acknowledgments
-\begin{acks}                            %% acks environment is optional
+%\begin{acks}                            %% acks environment is optional
                                         %% contents suppressed with 'anonymous'
@@
   %% extraction tools.
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
-\end{acks}
+%\end{acks}
 
@@
 
-Text of appendix \ldots
-
 \end{document}

### 2022-03-21 abstract-semantics.tex
Edits for shepherd review and camera-ready

@@
   \item $N : \texttt{Version} \rightarrow \texttt{Value}$ is a
-    (partial) function (i.e., a map) that mas versions to their
+    (partial) function (i.e., a map) that maps versions to their
     values.  We distinguish versions from values as different versions
@@
 the usual update notation, e.g., $H[b \mapsto v]$ is a map $H'$ such
-that $H'(b) = v$, and forall $b' \neq b$, $H'(b') = H(b')$.  Multiple
+that $H'(b) = v$, and for all $b' \neq b$, $H'(b') = H(b')$.  Multiple
 updates to a map are parsed left-associatively. Map $L$ is assumed to
@@
   In every reachable state $\Delta = ((V,E),N,C,H,L)$ of the abstract
-  machine: Forall $v_1,v_2 \in V$, if $v_1 \reaches v_2$ then $C(v_1)
+  machine: For all $v_1,v_2 \in V$, if $v_1 \reaches v_2$ then $C(v_1)
   \subseteq C(v_2)$.
@@
   In every reachable state $\Delta = ((V,E),N,C,H,L)$ of the abstract
-  machine: Forall distinct $b_1, b_2 \in dom(H)$, and $v_0, v_1, v_2
+  machine: For all distinct $b_1, b_2 \in dom(H)$, and $v_0, v_1, v_2
   \in V$ s.t.  $v_1 = H(b_1)$ and $v_2 = H(b_2)$ and $v_0 =
@@
   In every reachable state $\Delta$ = $((V,E),N,C,H,L)$ of the abstract
-  machine: Forall distinct $b_1, b_2 \in dom(H)$, and $v_1, v_2 \in V$
+  machine: For all distinct $b_1, b_2 \in dom(H)$, and $v_1, v_2 \in V$
   such that $v_1 = H(b_1)$ and $v_2 = H(b_2)$, the following is true:

### 2022-03-21 concrete-semantics.tex
Edits for shepherd review and camera-ready

@@
 concerns and provides a blueprint for a practical version-controlled
-replicated state machine. We first reify the the one-to-one
+replicated state machine. We first reify the one-to-one
 correspondence between replicas and branches we assumed informally in
@@
 state $\Delta$ being shared synchronously across all replicas. Next we
-present an efficient method to track the version history with help
+present an efficient method to track the version history with the help
 of vector clocks. Finally, we outline an algorithm for garbage
@@
 denote their replica-local copies. For instance, replica $i$'s copy of
-the \emph{head map} $H$ is given by $H\;i$, which we abbreviate to
+the \emph{head map} $H$ is given by $(H\;i)$, which we abbreviate to
 $H_i$ for notational convenience. Like $H$ in

### 2022-03-21 evaluation.tex
Edits for shepherd review and camera-ready

@@
 
-In this section we present an emperical evaluation of \quark on two
+In this section we present an empirical evaluation of \quark on two
 case studies. First is a collaborative document editing
@@
 versioning may be considered excessive in practice and could be
-disabled. Each user process runs a \quark thread that commits
-user-generated document versions to the local branch, while
-merging with the concurrent versions from the remote branches in the
-background. Each merge is synchronized as described in previous
-sections.
-
-\quark's background merges however pose a new problem as they create
-new versions on the local branch in the background while the user is
-busy editing an older version. When the user attempts to write their
-version of the document to the store, simply committing it would
-effectively override the concurrent updates from other users obtained
-via background merges. The solution, fortunately, is straightforward:
-we \emph{merge} the user-submitted value with the (value of) the
-latest version on the local branch to create a new version that
-includes the updates from either direction. Since this merge is fully
-confined to the local replica, it is guaranteed to not affect the LCA
-of the local branch with respect to a remote branch. To the external
-world, it appears as if the local branch has simply committed a new
-version.
-% The LCA for this merge is
-% simply the last version on the local branch read by the user. This
-% merge need not be synchronized as it doesn't alter the LCAs between
-% any two branches. The merged value can then returned to the user as
-% the result of the write.
-Thus, \quark's \C{write} is a function of type $\texttt{Value}
-\rightarrow \texttt{Value}$, where $\texttt{Value}$ is \C{doc} in the
-current application.
+disabled.
 
@@
 version, and persisting the resultant version to the store. We conduct
-the experiments on a three-node cluster of \C{i3.large} machines in
-Amazon \C{us-west2} data center. Each user connects to one of the
-machines, forks a new branch, and performs 1000 edits in succession,
-saving the document after each edit. We progressively increase the
-number of concurrent users editing the document from 3 to 200 and
-measure the impact of the increased concurrency on write latency.
-Fig.~\ref{fig:latency} shows the 10th, 50th (median), and 90th
-percentile latency values. The median and 10th percentile latencies
-remain more-or-less constant with a slight increase between 3 and 60
-concurrent users. This demonstrates that \quark has negligible effect
-on the per-operation latency, which is expected. The 90th percentile
-latency, however, shows no clear pattern by the virtue of being more
-susceptible to transient system conditions. Nonetheless, the maximum
+the experiments on a three-node cluster of \C{i3.large} machines
+located in Amazon \C{us-west2} data center. Each user connects to one
+of the machines, forks a new branch, and performs 1000 edits in
+succession, saving the document after each edit. We progressively
+increase the number of concurrent users editing the document from 3 to
+60 and measure the impact of the increased concurrency on write
+latency.  Fig.~\ref{fig:latency} shows the 10th, 50th (median), and
+90th percentile latency values. The median and 10th percentile
+latencies remain more-or-less constant with a slight increase between
+3 and 60 concurrent users. This is expected considering that \quark
+does not constrain the execution of user operations. The slight
+increase, we confirmed, is due to the increased latency of concurrent
+database writes. The 90th percentile latencies, however, show an
+initial increase before adopting a pattern similar to median
+latencies. We attribute this behavior to an idiosyncrasy of our
+multi-threading implementation, which is more likely to affect
+measurements at the extreme. In particular, we use OCaml v4.12
+extended with Lightweight Threads (LWT) library, which offers
+concurrency but not parallelism.  At lower number of replicas, the
+merge thread's lock requests succeed more often, resulting in the
+process spending more time merging than editing. We expect this
+pattern to smoothen out in OCaml v5.0, which introduces true
+multi-core parallelism. Notwithstanding this idiosyncrasy, the maximum
 value of 90th percentile latency measured (32ms) remains well below
@@
 \label{fig:latency}
+\vspace*{-0.2in}
 \end{figure}
@@
 \label{fig:monkey-staleness}
-  \vspace*{-0.2in}
+% \vspace*{-0.2in}
 \end{figure}
@@
 concurrent version on a remote replica. To measure staleness, we
-annotate every version $v$ with the timestamp $t$ of capturing the
-wall clock time of its creation. When $v$ is is merged into a remote
-branch $b$ at a later time $t'$, the difference $t' - t$ denotes the
-staleness of $v$ w.r.t the new version on $b$. One such staleness
-measurement is recorded for every merge that ever happens during the
-experiment. We compute 10th, 50th, and 90th percentiles of staleness
-values thus obtained. Fig.~\ref{fig:monkey-staleness} shows the
-results. As evident, staleness increases steadily with the increasing
-number of concurrent replicas, which is expected considering that
-merges are synchronized, and increasing the number of replicas reduces
-the number of opportunities for a replica to merge. The increase is
-roughly linear in the number of replicas as our implementation passes
-the lock around in a round-robin fashion, thus increasing the lock
-latency in proportion to the number of replicas. Nonetheless, the 90th
+annotate every version $v$ with the timestamp $t$ of the wall clock
+time of its creation. When $v$ is is merged into a remote branch $b$
+at a later time $t'$, the difference $t' - t$ denotes the staleness of
+$v$ w.r.t the new version on $b$. One such staleness measurement is
+recorded for every merge that ever happens during the experiment. We
+compute 10th, 50th, and 90th percentiles of staleness values thus
+obtained. Fig.~\ref{fig:monkey-staleness} shows the results. As
+evident, staleness increases steadily with the increasing number of
+concurrent replicas, which is expected considering that merges are
+synchronized, and increasing the number of replicas reduces the number
+of opportunities for a replica to merge. The increase is roughly
+linear in the number of replicas as our implementation passes the lock
+around in a round-robin fashion. Despite the steady increase, the 90th
 percentile staleness values remain low -- in the order of 100ms with
@@
 
-We repeat our experiments on a mergeable key-value store application
-we implemented using a Red-Black tree MRDT. Latency results are shown
-in Fig.~\ref{fig:rb-latency}, and staleness results in
-Fig.~\ref{fig:rb-staleness}.
+Our second case study involves a mergeable key-value store, which,
+unlike a text editor, is expected to work in the background supporting
+applications such as e-Commerce. The store is mergeable in the sense
+that it can merge conflicting values assigned to equal keys using
+value-specific merge functions. For instance, an e-Commerce
+application might store shopping carts as values, in which case the
+merge function on shopping carts is used to merge concurrent
+conflicting updates to a user's shopping cart. The key-value store is
+implemented using a Red-Black tree MRDT~\cite{mrdt}, which is a
+standard Red-Black tree implementation extended with a merge function.
+The merge function implements set merge semantics
+(Sec.~\ref{sec:motivation}) for concurrent insertions and deletions of
+keys, and defers to the value merge function for concurrent updates
+against a key. Concretely, \C{RBTree} is an OCaml functor
+parameterized on the \C{Key} and \C{Value} modules, where the latter
+is required to provide a \C{merge} function. We use integer keys and
+mergeable counter values in our experiments.
 
-\begin{figure}[ht]
-  \centering
-    \includegraphics[scale=0.4]{Figures/rbmonkey_latency}
-  \caption{Latency for Key-Value store operations under \quark}
-\label{fig:rb-latency}
-  \vspace*{-0.2in}
-\end{figure}
+% \begin{figure}[ht]
+%   \centering
+%     \includegraphics[scale=0.4]{Figures/rbmonkey_latency}
+%   \caption{Latency of operations on a key-value store implemented
+%   using a Red-Black tree MRDT.}
+% \label{fig:rb-latency}
+%   \vspace*{-0.2in}
+% \end{figure}
 
@@
     \includegraphics[scale=0.4]{Figures/rbmonkey_staleness}
-  \caption{Key-Value store staleness under \quark}
+  \caption{Staleness of key-value store merge operations.}
 \label{fig:rb-staleness}
-  \vspace*{-0.2in}
+% \vspace*{-0.2in}
 \end{figure}
 
+We use the same experimental setup as before -- an increasing number
+of concurrent processes perform a random operation on the local
+version of the tree while merging with the remote versions in the
+background.  Each process forks off a branch with an initial version
+of an \C{RBTree} of size 1000, performs an insertion, updatation, or
+deletion with probabilities of 0.5, 0.25, and 0.25. A new version of
+the tree is committed after each operation. Each process performs 1000
+operations with an inter-operation delay of 100ms. We measure the
+latency of each tree operation and the staleness of each merge.
+Latency measurements follow the same pattern as those for
+collaborative editing (Fig.~\ref{fig:latency}), and have been elided.
+The median latency remains around 50ms for most of the experiment.
+Staleness measurements are plotted in Fig.~\ref{fig:rb-staleness}. As
+before, staleness increases steadily in proportion to the increasing
+number of concurrent processes. The 90th percentile staleness exceeds
+1s for 15 concurrent processes, and 7s for 54 concurrent processes.
+Unlike the collaborative applications, however, the number of replicas
+in a conventional web application tends to be low, which could limit
+staleness to a tolerable range. For instance, in the e-Commerce
+application described above, 3 replicas of the shopping cart can be
+kept in sync with each other with a 90th percentile lag (staleness) of
+30ms.
+
 Our experiments bring to the fore an inherent tradeoff among the
-competing concerns of RDTs, namely (\rom{1}). The ease of programming
-convergence, (\rom{2}) Latency, and (\rom{3}).  Staleness. While CRDTs
+competing concerns of Convergent RDTs, namely (\rom{1}). The ease of
+programming, (\rom{2}) Latency, and (\rom{3}).  Staleness. While CRDTs
 try to optimize for latency and staleness, they require a significant
 amount of development and verification effort to be expended to ensure
-convergence~\cite{kleppmann2017}. In contrast, \quark lets developers
-derive convergent-by-construction RDTs from ordinary data data types
-that are optimized for latency, but incur a staleness overhead that
+convergence. In contrast, \quark lets developers derive
+convergent-by-construction MRDTs from ordinary data types that are
+optimized for latency, but incur a non-trivial staleness overhead that
 delays the time to convergence.

### 2022-03-21 implementation.tex
Edits for shepherd review and camera-ready

@@
-\begin{figure}[ht]
-  \centering
-    \includegraphics[scale=0.35]{Figures/implementation2}
-\caption{\quark implementation architecture}
-\label{fig:implementation}
-  \vspace*{-0.2in}
-\end{figure}
-
 \section{Implementation}
@@
 
-The implementation of \quark is a straightforward realization of the
-\quark distributed machine described in Sec.~\ref{sec:concrete-sem}.
-We manifest each component of the state, namely the branch map $B$,
-the value map $N$, and the head map $H$, as a column family (i.e., a
+The implementation of \quark largely follows the design of \quark
+distributed machine described in Sec.~\ref{sec:concrete-sem}.  We
+manifest each component of the state, namely the branch map $B$, the
+value map $N$, and the head map $H$, as a column family (i.e., a
 table) in Scylla. The synchronization needed to linearize merges is
@@
 enforced with help of \C{Quorum} reads and writes. Each user process
-is assigned its own replica of the MRDT.  Version vectors are realized
-as associative lists and stored in Scylla as blobs.
+is assigned its own branch containing a replica of the MRDT.  Version
+vectors are realized as associative lists and stored in Scylla as
+blobs.
 
-One key difference between the formalization and the implementation is
-in the treatment of MRDT values. Formalization assumes values to be
-atomic with no sharing in between them. In practice, however, an MRDT
-could be a linked data structure such as a binary tree, and two
-such values could share a significant amount of internal structure.
-Consequently, the size of the \emph{diff} between two consecutive
-versions of a value could be asymptotically less then the size of the
-data structure itself, in which case it unreasonable to transfer the
-entire data structure over the network. To facilitate the efficient
-computation of diffs between versions of data structures, we implement
-a \emph{content-addressable store} as a key-value table in Scylla
-where key is simply the SHA256 hash of the value. A linked data
-structure is stored as a collection of nodes, where each node links to
-the other by referring to its hash. The diff between two consecutive
-versions of a data structure would simply manifest as new entries in
-the content-addressable store reachable from the root of the new
-version. The new entries, being new data, are automatically replicated
-by Scylla, thus letting us reconstruct the new version at a remote
-location. The root hash of the new version is obtained from the value
-map $N$, which now maps version vectors to the \emph{hashes} of
-values.
+\begin{figure}[ht]
+  \centering
+    \includegraphics[scale=0.35]{Figures/implementation2}
+\caption{\quark implementation architecture}
+\label{fig:implementation}
+  \vspace*{-0.2in}
+\end{figure}
+
+The implementation however differs from the formalization in two
+significant ways. First is in the treatment of MRDT values.
+Formalization assumes values to be atomic with no sharing in between
+them. In practice, however, an MRDT could be a linked data structure
+such as a binary tree, and two such values could share a significant
+amount of internal structure.  Consequently, the size of the
+\emph{diff} between two consecutive versions of a value could be
+asymptotically less then the size of the data structure itself, in
+which case it unreasonable to transfer the entire data structure over
+the network. To facilitate the efficient computation of diffs between
+versions of data structures, we implement a \emph{content-addressable
+store} as a key-value table in Scylla where key is simply the SHA256
+hash of the value. A linked data structure is stored as a collection
+of nodes, where each node links to the other by referring to its hash.
+The diff between two consecutive versions of a data structure would
+simply manifest as new entries in the content-addressable store
+reachable from the root of the new version. The new entries, being new
+data, are automatically replicated by Scylla, thus letting us
+reconstruct the new version at a remote location. The root hash of the
+new version is obtained from the value map $N$, which now maps version
+vectors to the \emph{hashes} of values.
+
+The second significant difference between the implementation and the
+formalization is the timing of merges. Operational Semantics in
+Sec.~\ref{sec:abstract-sem} and Sec.~\ref{sec:concrete-sem} interleave
+commits and merges such that only one of them executes at any given
+time. Since commits are initiated by the user in practice,
+interleaving them with strongly consistent merges increases their
+latency as perceived by the user. For the user-perceived latency to
+remain unaffected, it is important that a replica be always available
+to execute user requests. \quark ensures this by handling user
+requests in a foreground thread that is always allowed to commit to
+the local branch. Merges, on the other hand, are relegated to the
+background. A background thread constantly scans the remote branches
+for new versions, and if there are any, merges them into the latest
+version on the local branch after checking the necessary preconditions
+(Sec.~\ref{sec:abstract-sem}).
+
+\quark's background merges however pose a new problem as they create
+new versions on the local branch in the background while a user
+operation is manipulating an older version in the foreground. When
+the user attempts to write their version to the store, simply
+committing it would effectively override the concurrent updates from
+other users obtained via background merges. The solution, fortunately,
+is straightforward: \quark \emph{merges} the user-submitted version
+with the latest version on the local branch to create a new version
+that includes the updates from either direction. Since this merge is
+fully confined to the local replica, it is guaranteed to not affect
+the LCA of the local branch in relation to any remote branch. To the
+external world, it appears as if the local branch has simply committed
+a new version that was derived from the older version. Since the
+commit operation now entails a merge, the merged version has to be
+returned to the user as the result of the commit.  Consequently the
+\C{write} operation exposed by \quark is a function of type
+$\texttt{Value} \rightarrow \texttt{Value}$, where $\texttt{Value}$ is
+any MRDT.

### 2022-03-21 introduction.tex
Edits for shepherd review and camera-ready

@@
 can be applied in any order and the replica states are still guaranteed to
-converge. Indeed, there exist common usecases in web applications that can
+converge. Indeed, there exist common use cases in web applications that can
 be implemented using Commutative Replicated Data Types
@@
 data types. Most common data type definitions come with at least a
-pair of operations that do not commute. For instance an \C{add} and a
+pair of operations that do not commute. For instance \C{add} and
 \C{remove} operations on a set do not commute if both are for the same
@@
 
-% In this paper we propose such an alternative approach to convergent
-% replicated data types. The key distinguishing characteristic of our
-% approach is the utilization of a distributed runtime to orchestrate
-% \emph{well-formed} distributed executions that are guaranteed to
-% converge. We define well-formedness as a constraint over the
-% \emph{states} that arise during the execution as opposed to the
-% \emph{operations} that are involved in the execution. To allow for its
-% enforcement, we adopt a state-centric model of replication based on
-% version-controlled \emph{mergeable} states~\cite{mrdt} instead of an
-% operation-centric model based on commutative operations. Unlike
-% commutativity, mergeability does not require a data type definition to
-% be refactored to suit distributed execution. 
 In this paper we describe an alternative approach to building
@@
 While constraining the data type semantics to conform to algebraic
-laws such as monotonicity might restore convergence, it would lead
-MRDTs down the path of CRDTs, which has known problems described
-above. We therefore propose an alternative approach to convergence --
-one that is based on runtime orchestration rather than static
-enforcement. In particular, we extend MRDTs with a distributed runtime
-that orchestrates only the \emph{well-formed} executions which are
-guaranteed to converge. Notably, our approach guarantees convergence
-of MRDTs \emph{regardless} of their merge semantics, thus obviating
-the need to constrain the type semantics or their implementations. The
-key insight behind our approach is the observation that a data type's
-\C{merge} is a complete function, and thefore can be safely invoked to
-merge concurrent versions of the type notwithstanding their individual
-operation histories. Thus, a replica's current version
-can always be made available for user operations with the guarantee
-that it can be safely merged with a concurrent version at a later
-point of time. The ability to arbitrarily delay merges gives us the
-opportunity to orchestrate them in a way that is \emph{effectively}
-linear, which inturn guarantees convergence.
-% combined with the simplicity
-% of state-based reasoning. Note that state-centric model of
-% replication has also been explored in the context of CRDTs. However,
-% state-based CRDTs require the replicated state to be organized as a
-% lattice -- a restriction we relax in our setting.
+laws such as monotonicity might restore convergence, it would burden
+the developer with the task of enforcing such constraints, which we
+would like to avoid. We therefore propose an alternative approach to
+convergence -- one that is based on runtime orchestration rather than
+static enforcement. In particular, we extend MRDTs with a distributed
+runtime that orchestrates only the \emph{well-formed} executions which
+are guaranteed to converge. Notably, our approach guarantees
+convergence of MRDTs \emph{regardless} of their merge semantics, thus
+obviating the need to constrain the type semantics or their
+implementations. The key insight behind our approach is the
+observation that a data type's \C{merge} is a complete function, and
+thefore can be safely invoked to merge concurrent versions of the type
+notwithstanding their individual operation histories. Thus, a
+replica's current version can always be made available for user
+operations with the guarantee that it can be safely merged with a
+concurrent version at a later point of time. The ability to
+arbitrarily delay merges gives us the opportunity to orchestrate them
+in a way that is \emph{effectively} linear, which inturn guarantees
+convergence.
 
@@
 of RDT operations. Orchestrating a well-formed convergent execution
-without impacting latency and availability may seem impossible due to
-the CAP theorem~\cite{cap}, but it is not the case in fact. While CAP
-theorem does rule out linearizability, it does not prohibit enforcing
-weaker consistency constraints on distributed executions. Our approach
-exploits this observation by (\rom{1}).  Identifying a novel set of
-constraints on distributed executions that guarantee their
-convergence, and (\rom{2}). Localizing such constraints on
-(asynchronously executed) state merges such that per-operation latency
-and system-wide availability remain unaffected. The downside of
-interfering with merges, however, is that the replicas may now take
-longer to converge -- a phenomenon we quantify using a metric called
-\emph{staleness} (Sec.~\ref{sec:eval}). Our approach thus brings to
-fore a fundamental tradeoff that was never explored in the context of
-RDTs before.
-
-%Sec.~\ref{sec:motivation} explains these insights in detail.
+without impacting latency and availability is possible despite the the
+CAP theorem~\cite{cap}. While CAP theorem does rule out
+linearizability, it does not prohibit enforcing weaker consistency
+constraints on distributed executions. Our approach exploits this
+observation by (\rom{1}).  Identifying a novel set of constraints on
+distributed executions that guarantee their convergence, and
+(\rom{2}). Localizing such constraints on (asynchronously executed)
+state merges such that per-operation latency and system-wide
+availability remain unaffected. The downside of interfering with
+merges, however, is that the replicas may now take longer to converge
+-- a phenomenon we quantify using a metric called \emph{staleness}
+(Sec.~\ref{sec:eval}). Our approach thus brings to fore a fundamental
+tradeoff that has not been explored in the context of RDTs.
 
@@
     formalization has also been mechanized in Ivy~\cite{ivy} and
-    automatically verified with help of Z3~\cite{z3}. To the best of
-    our knowledge, this is the first time the meta-theory of a formal
-    is proven completely inside an SMT solver.
+    automatically verified with the help of Z3~\cite{z3}. To the best
+    of our knowledge, this is the first time the meta-theory of a
+    formal system is proven completely inside an SMT solver.
 
@@
 \end{itemize}
+
+While our focus in this paper is on RDTs, we believe our contributions
+are also meaningful in the context of the long line of work on
+programming models for shared-memory concurrency. In particular, this
+work generalizes concurrent revisions~\cite{BBL+10} beyond fork-join
+parallelism and server-client paradigm. Our approach demonstrates how
+causal coherency-based semantics can be implemented more generally in
+decentralized systems while guaranteeing the convergence of concurrent
+executions.

### 2022-03-21 motivation.tex
Edits for shepherd review and camera-ready

@@
 updated in-place via \C{add} and \C{remove} operations. \C{Set} is an
-ordinary data type meant for for sequential execution. Under a
+ordinary data type meant for sequential execution. Under a
 concurrent execution with an asynchronously replicated state, \C{Set}
@@
 current replica. The \C{Add} and \C{Remove} effects on $e$ will now be
-tagged with $e$'s vector clock. When an \C{Remove} effect on $e$ is
+tagged with $e$'s vector clock. When a \C{Remove} effect on $e$ is
 received at a replica, it is applied only if the tagged vector clock
@@
 diverges. Here $R_1$ and $R_2$ start with version $v_0 \,=\, \{e\}$.
-$R_2$ performs a \C{remove} and and an \C{add} making versions
+$R_2$ performs a \C{remove} and an \C{add} making versions
 $v_1$ and $v_2$ respectively. Simultaneously, $R_1$ performs a
@@
 merges collectively propagate the effects of \C{add} and \C{remove}
-operations to $R_1$ and $R_2$. And despite being ordered in time,
-they nonetheless result in a divergent execution ($v_5 \neq v_7$). The
+operations to $R_1$ and $R_4$. And despite being ordered in time,
+they nonetheless result in a divergent execution ($v_6 \neq v_8$). The
 problem here is that, although merges are executed linearly, the
@@
 merges prior in the execution graph that are effectively concurrent.
-Considering that our approach to convergence in state-centric model
-crucially relies on linearity of merges, presence of multiple LCAs
-opens up the possibility of divergence. We therefore define
-well-formed execution graphs as those where LCAs for every pair of
-concurrent versions is unique. As we prove in
-Sec.~\ref{sec:abstract-sem}, enforcing this structural
-well-formedness condition indeed guarantees the convergence of
-distributed executions. In Sec.~\ref{sec:concrete-sem}, we formalize a
-distributed runtime called \quark whose primary purpose is to enforce
-the well-formedness condition on distributed executions. In
-Sec.~\ref{sec:implementation}, we describe its implementation. The
-runtime makes it possible to automatically enforce convergence,
-letting us promote ordinary sequential data types to convergent
-replicated types by simply equipping them with a merge function.
-Thus, extending the \C{Set} interface of Fig.~\ref{fig:ocaml-set} with
-the two-line \C{Set.merge} function shown above is sufficient to
-obtain a convergent replicated set data type.
+Considering that our approach to convergence in the state-centric
+model crucially relies on the linearity of merges, presence of
+multiple LCAs opens up the possibility of divergence. We therefore
+define well-formed execution graphs as those where LCAs for every pair
+of concurrent versions is unique. As we prove in
+Sec.~\ref{sec:abstract-sem}, enforcing this structural well-formedness
+condition indeed guarantees the convergence of distributed executions.
+In Sec.~\ref{sec:concrete-sem}, we formalize a distributed runtime
+called \quark whose primary purpose is to enforce the well-formedness
+condition on distributed executions. In Sec.~\ref{sec:implementation},
+we describe its implementation. The runtime makes it possible to
+automatically enforce convergence, letting us promote ordinary
+sequential data types to convergent replicated types by simply
+equipping them with a merge function.  Thus, extending the \C{Set}
+interface of Fig.~\ref{fig:ocaml-set} with the two-line \C{Set.merge}
+function shown above is sufficient to obtain a convergent replicated
+set data type.

### 2022-03-21 quark.tex
Edits for shepherd review and camera-ready

@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
+%\documentclass[sigplan,review]{acmart}
 \documentclass[sigplan,screen]{acmart}
@@
 %% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
-%%\acmJournal{PACMPL}
-%%\acmVolume{1}
-%%\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
-%%\acmArticle{1}
-%%\acmYear{2022}
-%%\acmMonth{1}
-%%\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-%%\startPage{1}
-\acmConference[PLDI]{ACM Conference}{June 2022}{San Diego, USA}%
-
 %% Copyright information
@@
 %% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
-
+%%\setcopyright{none}
+\copyrightyear{2022} 
+\acmYear{2022} 
+\setcopyright{rightsretained} 
+\acmConference[PLDI '22]{Proceedings of the 43rd ACM SIGPLAN International Conference on Programming Language Design and Implementation}{June 13--17, 2022}{San Diego, CA, USA}
+\acmBooktitle{Proceedings of the 43rd ACM SIGPLAN International Conference on Programming Language Design and Implementation (PLDI '22), June 13--17, 2022, San Diego, CA, USA}
+\acmDOI{10.1145/3519939.3523724}
+\acmISBN{978-1-4503-9265-5/22/06}
 %% Bibliography style
@@
   distinguishing aspect of our approach is that it guarantees
-  convergence \emph{unconditionally} -- without requiring data
-  type operations to satisfy algebraic laws such as commutativity and
+  convergence \emph{unconditionally} -- without requiring data type
+  operations to satisfy algebraic laws such as commutativity and
   idempotence. Consequently, programmers are no longer obligated to
@@
   sequential data types be reused in a distributed setting by
-  extending their implementations rather than refactoring them.
-  The novel component of our approach is a distributed runtime that
+  extending their implementations rather than refactoring them.  The
+  novel component of our approach is a distributed runtime that
   orchestrates \emph{well-formed } executions that are guaranteed to
@@
   case studies. 
+%%We also briefly discuss the utility of our approach in
+%%realizing causal coherency semantics in shared-memory concurrent
+%%systems,
 \end{abstract}
@@
 %% comma separated list
-\keywords{Replication, MRDT, CRDT, Runtime, Convergence, Collaborative
-Editing}  %% \keywords are mandatory in final camera-ready submission
+\keywords{Replication, MRDT, CRDT, Runtime, Convergence, Concurrent
+Revisions, Causal Consistency, decentralized systems}  %% \keywords are mandatory in final camera-ready submission
 
@@
 %% Acknowledgments
-%\begin{acks}                            %% acks environment is optional
-                                        %% contents suppressed with 'anonymous'
-  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-  %% acknowledge financial support and will be used by metadata
-  %% extraction tools.
-%\end{acks}
+\begin{acks}                            %% acks environment is optional
+  We thank the anonymous reviewers and our shepherd Doug Lea for their
+  thoughtful comments. First author has benefited from substantial
+  discussions with Anmol Sahoo and Suresh Jagannathan. Anmol has also
+  developed and maintained the OCaml-Scylla library, which is a key
+  piece of infrastructure underlying \quark.
+\end{acks}
 

### 2022-03-21 related.tex
Edits for shepherd review and camera-ready

@@
 
-The idea of replicated data types with three-way merge functions was
-first proposed in~\cite{mrdt}, where authors demonstrate an approach
-for deriving merge functions from first principles for various data
-types. However, the resultant \emph{mergeable replicated data types}
-(MRDTs) were never proved to be convergent. As we demonstrate for
-the \C{Set} data type in Sec.~\ref{sec:motivation}, simple merge
-functions with intuitive merge semantics can nonetheless lead to
-divergent executions. \quark can thus be used with MRDTs to preempt
-such executions and guarantee their convergence.
+Mergeable Replicated Data Types (MRDTs) were introduced
+in~\cite{mrdt}, where authors also demonstrate an approach for
+deriving merge functions from first principles.  Despite resulting in
+sensible merge semantics for several data types, their approach was
+never shown to guarantee the convergence of the resultant MRDTs.
+Indeed, as we demonstrate in Sec.~\ref{sec:motivation}, convergence
+of unrestricted MRDT executions cannot be guaranteed due to the
+presence of anomalies such as Fig.~\ref{fig:mrdt-exec-2} and
+Fig.~\ref{fig:mrdt-exec-4}. \quark fixes this problem by extending
+MRDTs with a runtime that limits their executions to those that are
+guaranteed to converge.
 
-Concurrent Revisions~\cite{BBL+10} proposes versioned state for
-shared-memory concurrent programs in order to guarantee determinism.
-The fork-join concurrency they consider is however too restrictive for
-the asynchronous distributed setting where replicas do not have a
-single coherent view of the system state.  Furthermore the
-hierarchical organization of the concurrent actors with a designated
-\emph{server} is inconsistent with the fully-decentralized model of
-execution adopted by several collaborative applications.
+% Concurrent Revisions~\cite{BBL+10} proposes versioned state for
+% shared-memory concurrent programs in order to guarantee determinism.
+% The fork-join concurrency they consider is however too restrictive for
+% the asynchronous distributed setting where replicas do not have a
+% single coherent view of the system state.  Furthermore the
+% hierarchical organization of the concurrent actors with a designated
+% \emph{server} is inconsistent with the fully-decentralized model of
+% execution adopted by several collaborative applications.
 
@@
 
+% Type systems
+The meta-theory of \quark abstract machine
+(Sec.~\ref{sec:abstract-sem} bears resemblance to the type safety
+proofs carried out in Wright and Felleisen's style~\cite{WF92}.
+Theorem~\ref{thm:progress} (Progress) is analogous to the Progress
+lemma of type safety, and Theorem~\ref{thm:convergence} (Convergence)
+is analogous to Preservation.  However, unlike the type-based
+approaches, safety in \quark is enforced through run-time monitoring.
+Static enforcement of convergence in MRDTs using, for e.g., Session
+Types~\cite{HNMSession16}, is an interesting direction for future
+work.
+
+% Git
+\quark's programming model is inspired by distributed version control
+systems in general, and Git in particular. To the best of our
+knowledge, operational semantics of Git has never been formalized.
+This is not a serious concern considering that humans are heavily
+involved in Git workflows, and the complexity of version histories
+manifest by Git is bound by the limits of human cognition.
+Nonetheless, it is possible to construct Git version histories that
+result in non-convergent and counter-intuitive final
+states~\cite{tycon,russell}. Such anomalies are more likely to occur
+if Git version histories are manifest commensurately with distributed
+executions of computer programs. \quark's run-time monitoring
+pre-empts this possibility in our case.
+
 Finally, the implementation of \quark bears resemblance to

### 2022-04-25 abstract-semantics.tex
Committing edits

@@
 \paragraph{Properties} We now formalize the notable properties of the
-abstract machine and its executions.
+abstract machine and its executions\footnote{Proofs can be found in
+the Appendix}.
 
@@
   \begin{itemize}
-    \item A ``quiescent'' state, where: $\forall b_1,b_2 \in
+    \item A ``quiescent'' state, where: $\\\forall b_1,b_2 \in
       dom(H).~C(H(b_1)) = C(H(b_2))$, Or
@@
 
-\noindent\paragraph{Ivy formalization} Appendix contains the manual
-proofs of the aforementioned theorems. In addition we also formalized
-the \quark abstract machine in Ivy~\cite{ivy}, and verified its
-properties automatically with help of Z3~\cite{z3}. Ivy is a
-multi-modal verification tool, whose primary purpose is to verify the
-design and implementations of distributed protocols. The key
-differentiating aspect of Ivy is its onus on predictable automation
-and decidable reasoning.  Automated proofs in Ivy are restricted to
-logical fragments for which the tool is a decision procedure. %Towards
+\noindent\paragraph{Ivy formalization}  In addition to the manual
+proofs, we also formalized the \quark abstract machine in
+Ivy~\cite{ivy} and verified its properties automatically with help of
+Z3~\cite{z3}. Ivy is a multi-modal verification tool, whose primary
+purpose is to verify the design and implementations of distributed
+protocols. The key differentiating aspect of Ivy is its onus on
+predictable automation and decidable reasoning.  Automated proofs in
+Ivy are restricted to logical fragments for which the tool is a
+decision procedure. %Towards
 % this end, Ivy checks the verification conditions for syntactic

### 2022-04-25 abstract.tex
Committing edits

@@
 Despite the utilization of a runtime, our approach comes at no additional cost of latency and availability.
-We implement our approach in a system called \quark and conduct a thorough evaluation of its performance over a large-scale collaborative editing application.
+Instead, we introduce a novel tradeoff against a metric called \emph{staleness}, which roughly corresponds to the time taken for replicas to converge.
+We implement our approach in a system called Quark and conduct a thorough evaluation of its tradeoffs.

### 2022-04-25 concrete-semantics.tex
Committing edits

@@
 
-\paragraph{Refinement} The following theorem asserts that the \quark
-distributed machine refines the abstract machine described in
-Sec.~\ref{sec:abstract-sem}:
+\paragraph{Refinement} The relationship between the \quark abstract
+(Sec.~\ref{sec:abstract-sem}) and distributed
+(Fig.~\ref{fig:quark-semantics}) machines is stated thus:
 \begin{theorem}[{\bf Refinement}]
@@
   \begin{itemize}
-    \item $R$ relates initial abstract and concrete states, i.e.,
+    \item $R$ relates the initial abstract and concrete states, i.e.,
       $R(\Delta_{\odot},\delta_{\odot})$.

### 2022-04-25 evaluation.tex
Committing edits

@@
 \label{fig:latency}
-\vspace*{-0.2in}
+\vspace*{-0.1in}
 \end{figure}
@@
 \label{fig:monkey-staleness}
-% \vspace*{-0.2in}
+\vspace*{-0.2in}
 \end{figure}
@@
 around in a round-robin fashion. Despite the steady increase, the 90th
-percentile staleness values remain low -- in the order of 100ms with
-the number of replicas under 30, and in the order of 1s with number of
-replicas under 60. While further optimizations might reduce staleness,
-a non-trivial staleness overhead is inevitable in \quark due to our
-use of synchronized merges to guarantee convergence.
+percentile staleness values remain low -- less than 1s with the
+number of replicas under 30, and less than 2s for number of replicas
+under 60. While further optimizations might reduce staleness, a
+non-trivial staleness overhead is inevitable in \quark due to our use
+of synchronized merges to guarantee convergence.
 
@@
 \label{fig:rb-staleness}
-% \vspace*{-0.2in}
+%\vspace*{-0.2in}
 \end{figure}

### 2022-04-25 quark.tex
Committing edits

@@
 \usepackage{microtype}
+\usepackage{balance} % To balance columns on last page
 
@@
 %% Title information
-\title[]{Runtime-Assisted Convergence in Replicated Data Types}
+\title[]{RunTime-Assisted Convergence in Replicated Data Types}
 %% [Short Title] is optional;
@@
                                         %% can be repeated if necessary
-%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
+\orcid{0000-0002-4189-3189}             %% \orcid is optional
 \affiliation{
@@
 %  \streetaddress{Street1 Address1}
-%  \city{City1}
+  \city{Boulder}
 %  \state{State1}
@@
 }
-%\email{first1.last1@inst1.edu}          %% \email is recommended
+\email{gowtham.kaki@colorado.edu}          %% \email is recommended
 
@@
 %  \streetaddress{Street2a Address2a}
-%  \city{City2a}
+  \city{Boulder}
 %  \state{State2a}
@@
 }
-%\email{first2.last2@inst2a.com}         %% \email is recommended
+\email{prasanth.prahladan@colorado.edu}         %% \email is recommended
 
 %% Third Author
-\author{Nicholas Lewchenko}
+\author{Nicholas V. Lewchenko}
 %\authornote{with author2 note}          %% \authornote is optional;
@@
 }
+\email{nile1033@colorado.edu}         %% \email is recommended
 %
@@
   converge. We implement our approach in a system called \quark and
-  conduct a thorough evaluation of its tradeoffs in the context of two
-  case studies. 
+  conduct a thorough evaluation of its tradeoffs. 
 %%We also briefly discuss the utility of our approach in
@@
 \keywords{Replication, MRDT, CRDT, Runtime, Convergence, Concurrent
-Revisions, Causal Consistency, decentralized systems}  %% \keywords are mandatory in final camera-ready submission
+Revisions, Causal Consistency, Decentralized Systems}  %% \keywords are mandatory in final camera-ready submission
 
@@
 
-\bibliography{all}
-
 %% Acknowledgments
@@
   We thank the anonymous reviewers and our shepherd Doug Lea for their
-  thoughtful comments. First author has benefited from substantial
-  discussions with Anmol Sahoo and Suresh Jagannathan. Anmol has also
-  developed and maintained the OCaml-Scylla library, which is a key
-  piece of infrastructure underlying \quark.
+  thoughtful comments. First author has benefited from discussions
+  with Anmol Sahoo and Suresh Jagannathan. Anmol has also developed
+  and maintained the OCaml-Scylla library, which is a key piece of
+  infrastructure underlying \quark.
 \end{acks}
 
+%% Appendix
+\appendix
+\section{Proofs}
+\input{proofs}
+
 %% Bibliography
-%\bibliography{bibfile}
-
-%% Appendix
-% \appendix
-% \section{Appendix}
+\bibliography{all}
 

### 2022-04-25 related.tex
Committing edits

@@
 
+\balance
 Mergeable Replicated Data Types (MRDTs) were introduced
@@
 distributed setting~\cite{bailis-vldb, alvaro-calm,
-gotsman-popl16,redblue-atc, redblue-osdi, ecinec}. These techniques
+gotsman-popl16, pldi15, redblue-osdi}. These techniques
 treat replicated storage as a black box with a fixed pre-defined
@@
 Finally, the implementation of \quark bears resemblance to
-Quelea~\cite{pldi15}. Both these systems are implemented as shim
-layers on top of off-the-shelf distributed stores, and both offload
-low-level concerns to the underlying store. Quelea however enforces
-high-level invariants on applications built with CRDTs. Operations are
-assumed to be commutative and applications are assumed to be
-convergent. In contrast, \quark enforces convergence of data types
-that are otherwise not guaranteed to converge.
+Quelea~\cite{pldi15} as both systems offload low-level concerns to an
+underlying data store. While Quelea enforces high-level invariants on
+applications built with CRDTs that are assumed to be convergent,
+\quark enforces convergence of data types that are otherwise not
+guaranteed to converge.

### 2022-04-25 proofs.tex
Added proofs

%\section{Proofs}

\begin{remark}
  The following algebraic properties over sets are referred in the
  proofs:
  \begin{itemize}
    \item {\bf Law1}: Set difference distributes over union. $\\\forall
      (A,B,C).~ A - (B \cup C) ~=~ (A - B) \cap (A - C)$
    \item {\bf Law2}: $\forall (A,B,C,D).~ (A - C) ~\cap~ (B - C) =
      \emptyset \\
      \hspace*{0.5in}\conj C \subseteq D ~\Rightarrow~ (A - D) ~\cap~ (B -
      D) = \emptyset$
    \item {\bf Law3}: Set union distributes over difference. $\\\forall
      (A,B,C).~ (A \cup B) - C ~=~ (A - C) \cup (B-C)$
    \item {\bf Law4}: Set union distributes over intersection.
      $\\\forall (A,B,C).~ (A \cup B) \cap C ~=~ (A \cap C) \cup
      (B \cap C)$
    \item {\bf Law5}: $\forall (A,B,C,D).~ (A - C) ~\cap~ (B - C) =
      \emptyset \\
      \hspace*{0.5in}\conj A \subseteq B ~\Rightarrow~ A \subseteq C$
  \end{itemize}
\end{remark}

% \begin{theorem}[{\bf Uniqueness of LCA}]
%   Assuming that we start with the initial state $\Delta_{\odot}$, the
%   transition rules in Fig.~\ref{fig:sem-head-pull} will always lead to
%   a state $\Delta = ((V,E),B,H,L)$ where every distinct pair of
%   branches, $b_1 \in dom(H)$ and $b_2 \in dom (H)$, has a single LCA given by
%   $L(b_1,b_2)$, which is equal to $L(b_2,b_1)$.
% \end{theorem}
\begin{proof}[Proof of Lemma~\ref{lem:lca-uniqueness}]

  By induction on $\Delta$. The initial state $\Delta_{\odot}$
  vacuously satisfies the unique LCA condition. The inductive step has
  three cases:
  \begin{itemize}
    \item Case {\sc Commit}: LCA function $L$ is not updated in this
      case. Neither are any new branches created. Hence the proof
      follows from the induction hypothesis (IH).

    \item Case {\sc Fork}: We fork off a new branch $b'$ from $b$ in
      this case. The LCA of $b$ and $b'$ in this case is clearly
      $H(b)$ -- the version from which $b'$ is forked. Indeed, this is
      the updated value of $L(b,b')$ in the conclusion.

      For every other branch $b''$, the LCA of $b'$ and $b'$ (i.e.,
      $L(b',b'')$) is same as the LCA of $b$ and $b''$ (i.e.,
      $L(b,b'')$), which IH gurantees to be unique. This is because
      every ancestor of $H(b')$ is either equal of $H(b)$ or is an
      ancestor of $H(b)$, the hence lowest common ancestor of $H(b'')$
      and $H(b)$ should also be the lowest common ancestor of $H(b'')$
      and $H(b')$, which is indeed the updated value of $L(b,b'')$ in
      the conclusion.

    \item Case {\sc Merge}: The proof follows from the premise
    of \rulelabel{Merge}, which requires that, for every branch $b''$ in
    the system, the LCA of $b''$ with merging branches $b$ and $b'$ be
    causally related, i.e, $L(b,b'') \reaches L(b',b'') ~\disj$
    $L(b',b'') \reaches L(b,b'')$. Fig.~\ref{fig:merge-precondition}
    helps visualize this condition in the most general case when
    (\rom{1}). Branches $b$, $b'$, and $b''$ are distinct, and
    (\rom{2}). Their LCAs $L(b,b'')$ and $L(b',b'')$ lie on a distinct
    pair of branches not equal to $b$, $b'$, and $b''$. In the figure,
    once you merge $b'$ into $b$, every version $v$ that is an
    ancestor of $L(b',b'')$, i.e., $v \reaches L(b',b'')$, will be  a
    common ancestor of $H(b)$ and $H(b'')$. Clearly, $L(b',b'')$ is
    the lowest among such common ancestors. But the current lowest
    common ancestor of $b$ and $b''$ is $L(b,b'')$. This leads to
    the condition where there are two lowest common ancestors --
    $L(b',b'')$ and $L(b,b'')$, \emph{unless} both are ancestrally
    related. The premise
    $L(b,b'') \reaches L(b',b'') \disj L(b',b'') \reaches L(b,b'')$
    guarantees that they are causally-related, ensuring the uniqueness
    of LCA.

    \item Case {\sc FastFwd}: Similar to {\sc Merge}

  \end{itemize}

\end{proof}

% \begin{lemma}[Commit sets grow monotonically]
%   \label{lem:commit-monotonicity}
%   Let $\Delta = ((V,E),N,C,H,L)$ be the state of a version-control
%   system. Forall $v_1,v_2 \in V$, if $v_1 \reaches v_2$ then $C(v_1)
%   \subseteq C(v_2)$.
% \end{lemma}
\begin{proof}[Proof of Lemma~\ref{lem:commit-monotonicity}]
  Straightforward induction on $\Delta$.
\end{proof}

% \begin{lemma}[{\bf Commit sets modulo LCA are disjoint}]
%   \label{lem:commit-disjointness}
%   Let $\Delta = ((V,E),N,C,H,L)$ be the state of a version-control
%   system. Forall distinct $b_1, b_2 \in dom(H)$, and $v_0, v_1, v_2
%   \in V$ s.t.  $v_1 = H(b_1)$ and $v_2 = H(b_2)$ and $v_0 =
%   L(b_1,b_2)$, the following is true:
%   $(C(v_2) - C(v_0)) ~\cap~ (C(v_1) - C(v_0)) = \emptyset$.
% \end{lemma}
\begin{proof}[Proof of Lemma~\ref{lem:commit-disjointness}]
  Proof by induction on $\Delta$. We consider the cases below starting
  with then {\sc Merge} case.
  \begin{itemize}
    \item Case {\sc Merge}: In all the cases where $b_1 \neq b$ and
      $b_2 \neq b$, proof follows from IH because nothing changes for
      these branches in $\Delta'$. We shall therefore focus on the
      case when one of $b_1$ or $b_2$ is $b$. Without loss of
      generality, let's assume $b_1 = b$. Two cases now depending on
      what $b_2$ is:
      \begin{itemize}
        \item Case $b_2 = b'$: From the {\sc Merge} rule, it's clear
          that $v_1 = H'(b) = v$ and $v_2 = H'(b') = H(b')$ and $v_0 =
          L'(b,b') = H(b') = v_2$. Since $v_2 = v_0$, $C(v_2) - C(v_0)
          = \emptyset$, hence the proof.
        \item Case $b_2 \neq b'$. Let $b_2 = b'' \neq b'$. Hence $v_1 =
          H'(b)$, $v_2 = H'(b'') = H(b'')$, and $v_0 = L'(b,b'')$,
          which would either be $L(b,b'')$ or $L(b',b'')$
          (see the premise of {\sc Merge}). For notational
          convenience let's use $v'$ to denote $H'(b') = H(b')$, $v''$
          to denote $H'(b'') = H(b'')$, and $v_{old}$ to denote
          $H(b)$. The hypotheses are listed below:
          \begin{smathpar}
          \begin{array}{lr}
            C(v_{old}) - C(L(b,b'')) ~\cap~ C(v'') - C(L(b,b'')) =
              \emptyset & IH1\\
            C(v') - C(L(b',b'')) ~\cap~ C(v'') - C(L(b',b'')) =
              \emptyset & IH2\\
            L(b,b'') \reaches L(b',b'') \disj L(b',b'') \reaches
              L(b,b'') & H1\\
            C(H(b')) \supset C(L(b,b')) & H2\\
            C(H(b)) \supset C(L(b,b')) & H3\\
          \end{array}
          \end{smathpar}
          We will now destruct $H1$ and consider each disjunct
          separately.
          \begin{itemize}
            \item Case $L(b,b'') \reaches L(b',b'')$: From
              Lemma~\ref{lem:commit-monotonicity} we know:
              \begin{smathpar}
              \begin{array}{lr}
                C(L(b,b'')) \subseteq C(L(b',b'')) & H4\\
              \end{array}
              \end{smathpar}
              The Goal is to prove the following:
              \begin{smathpar}
              \begin{array}{c}
                C'(v) - C'(L'(b,b'')) ~\cap~ C'(v'') - C'(L'(b,b''))
                ~=~ \emptyset\\
              \end{array}
              \end{smathpar}
              We shall now rewrite the goal through a series of
              equalities. First, since $C'(v) = C(v_{old}) \cup
              C(v')$, and $L'(b,b'') = L(b',b'')$ (since $L(b,b'')
              \reaches L(b',b'')$ as per our assumption in this case),
              and $C'(v'') = C(v'')$:
              \begin{smathpar}
              \begin{array}{p{\columnwidth}}
                $(C(v_{old}) \cup C(v')) - C(L(b',b'')) ~\cap~ C(v'') - C(L(b',b''))
                ~=~ \emptyset$\\
              \end{array}
              \end{smathpar}
              Rewrite using Algebraic Law 3:
              \begin{smathpar}
              \begin{array}{c}
                (C(v_{old}) - C(L(b',b'')) ~\cup~ C(v') -
                C(L(b',b''))) \\ ~\cap~ C(v'') - C(L(b',b'')) ~=~ \emptyset\\
              \end{array}
              \end{smathpar}
              Rewrite using Algebraic Law 4:
              \begin{smathpar}
              \begin{array}{c}
                C(v_{old}) - C(L(b',b'')) ~\cap~ C(v'') -
                C(L(b',b'')) \\
                \hspace*{0.1in}
                \cup~~ C(v') - C(L(b',b'')) ~\cap~ C(v'') - C(L(b',b''))
                ~~=~~ \emptyset\\
              \end{array}
              \end{smathpar}
              Rewrite using IH2:
              \begin{smathpar}
              \begin{array}{c}
                C(v_{old}) - C(L(b',b'')) ~\cap~ C(v'') -
                C(L(b',b'')) ~=~ \emptyset\\
              \end{array}
              \end{smathpar}
              Applying Algebraic Law2 with $C := L(b,b'')$:
              \begin{smathpar}
              \begin{array}{c}
                C(v_{old}) - C(L(b,b'')) ~\cap~ C(v'') - C(L(b,b''))
                ~=~ \emptyset \\~\conj~ L(b,b'') \subseteq L(b',b'')\\
              \end{array}
              \end{smathpar}
              Now both conjuncts of the goal follow from hypotheses
              $IH1$ and $H4$, respectively.

            \item Case $L(b',b'') \reaches L(b,b'')$: From
              Lemma~\ref{lem:commit-monotonicity} we know:
              \begin{smathpar}
              \begin{array}{lr}
                C(L(b',b'')) \subseteq C(L(b,b'')) & H5\\
              \end{array}
              \end{smathpar}
              The Goal is to prove the following:
              \begin{smathpar}
              \begin{array}{c}
                C'(v) - C'(L'(b,b'')) ~\cap~ C'(v'') - C'(L'(b,b''))
                ~=~ \emptyset\\
              \end{array}
              \end{smathpar}
              We shall now rewrite the goal through a series of
              equalities. First, since $C'(v) = C(v_{old}) \cup
              C(v')$, and $L'(b,b'') = L(b,b'')$ (since $L(b',b'')
              \reaches L(b,b'')$ as per our assumption in this case),
              and $C'(v'') = C(v'')$:
              \begin{smathpar}
              \begin{array}{c}
                (C(v_{old}) \cup C(v')) - C(L(b,b'')) \\
                ~~\cap~ C(v'') - C(L(b,b''))
                ~=~ \emptyset\\
              \end{array}
              \end{smathpar}
              Rewrite using Algebraic Law 3:
              \begin{smathpar}
              \begin{array}{c}
                (C(v_{old}) - C(L(b,b'')) ~\cup~ C(v') - C(L(b,b'')))\\
                ~~\cap~ C(v'') - C(L(b,b'')) ~=~ \emptyset\\
              \end{array}
              \end{smathpar}
              Rewrite using Algebraic Law 4:
              \begin{smathpar}
              \begin{array}{c}
                C(v_{old}) - C(L(b,b'')) ~\cap~ C(v'') -
                C(L(b,b'')) \\
                \hspace*{0.1in}
                \cup~~ C(v') - C(L(b,b'')) ~\cap~ C(v'') - C(L(b,b''))
                ~~=~~ \emptyset\\
              \end{array}
              \end{smathpar}
              Rewrite using IH1:
              \begin{smathpar}
              \begin{array}{c}
                C(v') - C(L(b,b'')) ~\cap~ C(v'') -
                C(L(b,b'')) ~=~ \emptyset\\
              \end{array}
              \end{smathpar}
              Apply Algebraic Law2 with $C := L(b',b'')$:
              \begin{smathpar}
              \begin{array}{c}
                C(v') - C(L(b',b'')) ~\cap~ C(v'') - C(L(b',b''))
                ~=~ \emptyset \\
                ~~\conj~ L(b',b'') \subseteq L(b,b'')\\
              \end{array}
              \end{smathpar}
              Now both conjuncts of the goal follow from hypotheses
              $IH2$ and $H5$, respectively.
          \end{itemize}
      \end{itemize}

    \item Case {\sc FastFwd}: Fast forwarding is a special case of
      merge, hence the proof is going to be very similar to the {\sc
      Merge} case. To reduce the {\sc FastFwd} case to {\sc Merge},
      we make the following observations about the diferences between
      both the rules:
      \begin{itemize}
        \item The subsumption relation $C(H(b)) \supset C(L(b,b'))$ in
          the premise of {\sc Merge} is replaced with equality in {\sc
          FastFwd}, but this is irrelevant is the subsumption premise
          is never used in the proof of {\sc Merge} case.

        \item $N'(v) = N(H(b'))$ in {\sc FastFwd} whereas it is
          computed using {\sf merge} in {\sc Merge}. However, the
          current theorem says nothing about $N$, and the rest of
          $\Delta$ is independent of $N$, hence this is irrelevant.

        \item $C'(H'(b)) = C(H(b'))$ in the conclusion of {\sc
          FastFwd} whereas it is equal to $C(H(b)) \cup C(H(b'))$ in
          {\sc Merge}. However since $C(H(b)) = C(L(b,b')) \subset
          C(H(b'))$ in the premise of {\sc FastFwd}, $C'(H'(b))$ in
          the conclusion is indeed equal to $C(H(b)) \cup C(H(b'))$
          like in {\sc Merge}.
      \end{itemize}
      Consequently, we can reuse the proof of {\sc Merge} as a proof
      of {\sc FastFwd}.

    \item Case {\sc Fork}: Branch $b'$ is forked from branch $b$. In
      all the cases where $b_1 \neq b'$ and $b_2 \neq b'$, proof
      follows from IH because nothing changes for these branches in
      $\Delta'$. We shall therefore focus on the case when one of
      $b_1$ or $b_2$ is $b'$. Without loss of generality, let's assume
      $b_2 = b$. We now have two cases depending on what values $b_1$
      can take:
      \begin{itemize}
        \item Case $b_1 = b$: In this case $L'(b_1,b_2) = L'(b,b') =
          H(b) = H'(b)$. Hence, $C'(H'(b_1)) - L'(b_1,b_2)$ is equal
          to $C'(H'(b)) - C'(H'(b)) = \emptyset$. Proof follows.

        \item Case $b_1 = b'' \neq b$: In this case, the proof follows
          straightforwardly from the inductive hypothesis that asserts
          the lemma for $b$ and $b''$ in $\Delta$ because $L'(b',b'')$
          $=$ $L(b,b'')$ and $C(H(b')) = C(H(b))$.
      \end{itemize}

    \item Case {\sc Commit}: Proof is trivial because $c$ is a fresh
      commit id that does not exist in $\Delta$.

  \end{itemize}
\end{proof}

% \begin{theorem}[{\bf Convergence}]
%   Let $\Delta = ((V,E),N,C,H,L)$ be the state of a version-control
%   system. Forall distinct $b_1, b_2 \in dom(H)$, and $v_1, v_2 \in V$ s.t.
%   $v_1 = H(b_1)$ and $v_2 = H(b_2)$, the following is true:
%   $C(v_1) = C(v_2) \Rightarrow N(v_1) = N(v_2)$.
% \end{theorem}
\begin{proof}[Proof of Theorem~\ref{thm:convergence}]
  Proof by induction on $\Delta$. Base case trivially satisfies as
  there are no two branches in $\Delta_{\odot}$. Inductive proof has
  four cases corresponding to the four rules in
  Fig.~\ref{fig:git-semantics}:
  \begin{itemize}
    \item Case {\sc Commit}: Branch $b$ is extended with a fresh
      commit $c$, hence there does not exist another branch whose head
      has same commit set as $H'(b)$.

    \item Case {\sc Fork}: In all the cases where $b_1 \neq b'$ and
      $b_2 \neq b'$, proof follows from IH because nothing changes for
      these branches in $\Delta'$. We shall therefore focus on the
      case when one of $b_1$ or $b_2$ is $b'$. Without loss of
      generality, let's assume $b_2 = b$. Two cases now depending on
      what $b_1$ is:
      \begin{itemize}
        \item  Case $b_1 = b$: Proof by definition as {\sc Fork}
          defines $C'(H'(b')) = C(H(b)) = C'(H'(b))$ and $N'(H'(b')) =
          N(H(b)) = N'(H'(b))$. 
        \item Case $b_1 = b'' \neq b$: Proof follows from IH relating
          $b_1$ and $b$ in $\Delta$ because $C'(H'(b')) = C(H(b))$ and $N'(H'(b')) =
          N(H(b))$.
      \end{itemize}

    \item Case {\sc FastFwd}: The proof is similar to the {\sc Fork}
      case since $C'(H'(b)) = C(H(b')) = C'(H'(b'))$ and $N'(H'(b)) =
      N(H(b')) = N'(H'(b'))$.

    \item Case {\sc Merge}:  In all the cases where $b_1 \neq b$ and
      $b_2 \neq b$, proof follows from IH because nothing changes for
      these branches in $\Delta'$. We shall therefore focus on the
      case when one of $b_1$ or $b_2$ is $b$. Without loss of
      generality, let's assume $b_1 = b$. Two cases now depending on
      what $b_2$ is:
      \begin{itemize}
        \item Case $b_2 = b'$: The relevant premise of {\sc Merge} is
          reproduced below:
          \begin{smathpar}
          \begin{array}{lr}
            C(H(b)) \supset C(L(b,b')) & H0\\
          \end{array}
          \end{smathpar}
          We shall prove that $C'(H'(b_1))$ cannot be equal to
          $C'(H'(b_2))$ by first assuming that they are equal and then
          deriving a contradiction.
          \begin{smathpar}
          \begin{array}{lr}
            C'(H'(b_1)) = C'(H'(b_2)) & H1\\
          \end{array}
          \end{smathpar}
          Since $b_1 = b$, and $b_2 = b'$, and $C'(H'(b')) = C(H(b'))$:
          \begin{smathpar}
          \begin{array}{lr}
            C'(H'(b)) = C(H(b')) & \\
          \end{array}
          \end{smathpar}
          Since $C'(H'(b')) = C(H(b)) \cup C(H(b'))$:
          \begin{smathpar}
          \begin{array}{lr}
            C(H(b)) \cup C(H(b'))= C(H(b')) & H2\\
          \end{array}
          \end{smathpar}
          From Lemma~\ref{lem:commit-disjointness} on $\Delta$, we
          know the following:
          \begin{smathpar}
          \begin{array}{lr}
            C(H(b)) - C(L(b,b'))\\
            \hspace*{0.5in}~\cap~ C(H(b')) - C(L(b,b')) ~=~
            \emptyset & H3\\
          \end{array}
          \end{smathpar}
          Rewriting H3 using H2:
          \begin{smathpar}
          \begin{array}{lr}
            C(H(b)) - C(L(b,b'))\\
            \hspace*{0.25in}\cap~~ (C(H(b)) \cup C(H(b')) - C(L(b,b')) ~=~
            \emptyset & \\
          \end{array}
          \end{smathpar}
          Rewriting using Algebraic Law3:
          \begin{smathpar}
          \begin{array}{lr}
            C(H(b)) - C(L(b,b')) ~~\cap\\
            (C(H(b)) - C(L(b,b')) ~\cup~ C(H(b')) - C(L(b,b'))) ~=~ \emptyset & \\
          \end{array}
          \end{smathpar}
          Rewriting using Algebraic Law4:
          \begin{smathpar}
          \begin{array}{lr}
            (C(H(b)) - C(L(b,b')) ~\cap~ C(H(b)) - C(L(b,b')))\\
            ~~\cup~~ (C(H(b)) - C(L(b,b')) ~\cap~ C(H(b')) - C(L(b,b'))) 
            ~=~ \emptyset & \\
          \end{array}
          \end{smathpar}
          Simplifying, and Rewriting using $H3$:
          \begin{smathpar}
          \begin{array}{lr}
            C(H(b)) - C(L(b,b')) ~=~ \emptyset & \\
          \end{array}
          \end{smathpar}
          Which is possible if and only if:
          \begin{smathpar}
          \begin{array}{lr}
            C(H(b)) \subseteq C(L(b,b')) & H4\\
          \end{array}
          \end{smathpar}
          $H0$ and $H4$ now result in a contradiction.

        \item Case $b_2 = b'' \neq b'$: Premises of {\sc Merge}:
          \begin{smathpar}
          \begin{array}{lr}
            C(H(b)) \supset C(L(b,b'))& H1\\
            C(H(b')) \supset C(L(b,b'))& H2\\
            L(b,b'') \reaches L(b',b'') \disj L(b',b'') \reaches
            L(b,b'') & H3\\
          \end{array}
          \end{smathpar}
          From the conclusion of {\sc Merge}:
          \begin{smathpar}
          \begin{array}{lr}
            C'(H'(b)) = C(H(b)) \cup C(H(b')) & H4\\
            C'(H'(b'')) = C(H(b'')) & H5\\
          \end{array}
          \end{smathpar}
          And the premise of the theorem:
          \begin{smathpar}
          \begin{array}{lr}
            C'(H'(b)) = C'(H'(b'')) & \\
          \end{array}
          \end{smathpar}
          Due to $H5$, this is equivalent to:
          \begin{smathpar}
          \begin{array}{lr}
            C'(H'(b)) = C(H(b'')) & H6\\
          \end{array}
          \end{smathpar}
          From Lemma~\ref{lem:commit-disjointness} on $H'(b')$ and
          $H'(b'') = H(b'')$:
          \begin{smathpar}
          \begin{array}{lr}
            C'(H'(b)) - C'(L'(b,b'')) ~\cap\\
            \hspace*{0.5in}C(H(b'')) - C'(L'(b,b''))
            ~=~ \emptyset & \\
          \end{array}
          \end{smathpar}
          Rewriting using $H6$:
          \begin{smathpar}
          \begin{array}{lr}
            C'(H'(b)) - C'(L'(b,b'')) ~=~ \emptyset & \\
          \end{array}
          \end{smathpar}
          Hence:
          \begin{smathpar}
          \begin{array}{lr}
            C'(H'(b)) \subseteq C'(L'(b,b''))
          \end{array}
          \end{smathpar}
          Rewriting using $H4$:
          \begin{smathpar}
          \begin{array}{lr}
            C(H(b)) \cup C(H(b')) ~\subseteq~ C'(L'(b,b'')) & H7\\
          \end{array}
          \end{smathpar}
          Let us distruct $H3$ and consider the two disjuncts
          separately:
          \begin{itemize}
            \item Case $L(b,b'') \reaches L(b',b'')$: From the
              conclusion of {\sc Merge}:
              \begin{smathpar}
              \begin{array}{lr}
                L'(b,b'') = L(b',b'') & H8\\
              \end{array}
              \end{smathpar}
              Rewriting $H7$ using $H8$:
              \begin{smathpar}
              \begin{array}{lr}
                C(H(b)) \cup C(H(b')) ~\subseteq~ C(L(b',b'')) & H7\\
              \end{array}
              \end{smathpar}
              Which tells us:
              \begin{smathpar}
              \begin{array}{lr}
                C(H(b)) \subseteq C(L(b',b'')) & H9 \\
              \end{array}
              \end{smathpar}
              Note that $L(b',b'')$ is LCA of $H(b')$ and $H(b'')$,
              hence $L(b',b'') \reaches H(b')$. From
              Lemma~\ref{lem:commit-monotonicity}, we get:
              \begin{smathpar}
              \begin{array}{lr}
                C(L(b',b'')) \subseteq C(H(b')) & H10\\
              \end{array}
              \end{smathpar}
              From $H9$ and $H10$:
              \begin{smathpar}
              \begin{array}{lr}
                C(H(b)) \subseteq C(H(b')) &  H11\\
              \end{array}
              \end{smathpar}
              Lemma~\ref{lem:commit-disjointness} on $H(b)$ and
              $H(b')$ tells us:
              \begin{smathpar}
              \begin{array}{lr}
                C(H(b)) - C(L(b,b')) ~\cap\\
                \hspace*{0.5in}C(H(b')) - C(L(b,b')) ~=~
                \emptyset & H12\\
              \end{array}
              \end{smathpar}
              Applying Algebraic Law 5 on $H11$ and $H12$:
              \begin{smathpar}
              \begin{array}{lr}
                C(H(b)) \subseteq C(L(b,b'))
              \end{array}
              \end{smathpar}
              Which contradicts $H1$. Hence $C'(H'(b))$ cannot be
              equal to $C'(H'(b''))$.

            \item Case $L(b',b'') \reaches L(b,b'')$: Very similar to
              the above case. Using the similar approach as above, we
              derive $C(H(b')) \subseteq C(H(b))$ (the opposite of
              $H11$), which leads us to $C(H(b')) \subseteq
              C(L(b,b'))$, thus contradicting $H2$.
          \end{itemize}
      \end{itemize}
  \end{itemize}
\end{proof}

%%\begin{proof}[Proof of Theorem~\ref{thm:progress}]
%%  By induction on $\Delta$. Non-trivial cases are discussed below.
%%  \begin{itemize}
%%    \item \rulelabel{Commit}: Let $b$ be the branch on which a
%%      new version $v$ is committed. Destructing IH gives us two
%%      subcases:
%%      \begin{itemize}
%%        \item Case 1 - System is in queiscent state before commit: If
%%          $b$ is the only branch, then system is again in queiscent
%%          state after commit. Else, there exists a branch $b'$ s.t.
%%          $L(b,b')$ exists and is unique
%%          (Lemma~\ref{lem:lca-uniqueness}). Linearity of merges
%%          guarantees that in queiscent state all LCAs are linearly
%%          ordered. Since \rulelabel{commit} doesn't change $L$, and
%%          $v$ is a new version, $v$ can be committed onto $b'$.

%%        \item Case 2 - System is mergeable state before commit: Since
%%          \rulelabel{Commit} doesn't alter $L$, whatever branches are
%%          mergeable before the commit, remain mergeable after the commit.
%%      \end{itemize}

%%    \item \rulelabel{Fork}: Let $b'$ is the new branch forked from
%%      $b$. If system was in queiscent state before, then it is again
%%      in the queiscent state as $H(b') = H(b)$. Else, if the system
%%      was in a mergeable state before, then it is again in the
%%      mergeable state as \rulelabel{Fork} doesn't change the LCA
%%      relationships among existing branches.
%%  \end{itemize}
%%\end{proof}

### 2022-04-25 proofs.tex
Move appendix (proofs) to separate document
Build it with:

latexmk -pdf appendix

@@
 % \end{theorem}
-\begin{proof}[Proof of Lemma~\ref{lem:lca-uniqueness}]
+\begin{proof}[Proof of Lemma~3.4]
 
@@
     causally related, i.e, $L(b,b'') \reaches L(b',b'') ~\disj$
-    $L(b',b'') \reaches L(b,b'')$. Fig.~\ref{fig:merge-precondition}
+    $L(b',b'') \reaches L(b,b'')$. Fig.~7
     helps visualize this condition in the most general case when
@@
 % \end{lemma}
-\begin{proof}[Proof of Lemma~\ref{lem:commit-monotonicity}]
+\begin{proof}[Proof of Lemma~3.5]
   Straightforward induction on $\Delta$.
@@
 % \end{lemma}
-\begin{proof}[Proof of Lemma~\ref{lem:commit-disjointness}]
+\begin{proof}[Proof of Cor.~3.6]
   Proof by induction on $\Delta$. We consider the cases below starting
@@
             \item Case $L(b,b'') \reaches L(b',b'')$: From
-              Lemma~\ref{lem:commit-monotonicity} we know:
+              Lemma~3.5 we know:
               \begin{smathpar}
@@
             \item Case $L(b',b'') \reaches L(b,b'')$: From
-              Lemma~\ref{lem:commit-monotonicity} we know:
+              Lemma~3.5 we know:
               \begin{smathpar}
@@
 % \end{theorem}
-\begin{proof}[Proof of Theorem~\ref{thm:convergence}]
+\begin{proof}[Proof of Theorem~3.7]
   Proof by induction on $\Delta$. Base case trivially satisfies as
@@
   four cases corresponding to the four rules in
-  Fig.~\ref{fig:git-semantics}:
+  Fig.~6:
   \begin{itemize}
@@
           \end{smathpar}
-          From Lemma~\ref{lem:commit-disjointness} on $\Delta$, we
+          From Cor.~3.6 on $\Delta$, we
           know the following:
@@
           \end{smathpar}
-          From Lemma~\ref{lem:commit-disjointness} on $H'(b')$ and
+          From Cor.~3.6 on $H'(b')$ and
           $H'(b'') = H(b'')$:
@@
               hence $L(b',b'') \reaches H(b')$. From
-              Lemma~\ref{lem:commit-monotonicity}, we get:
+              Lemma~3.5, we get:
               \begin{smathpar}
@@
               \end{smathpar}
-              Lemma~\ref{lem:commit-disjointness} on $H(b)$ and
+              Cor.~3.6 on $H(b)$ and
               $H(b')$ tells us:

### 2022-04-25 motivation.tex
Shorten to 14 pages by rewording

@@
 \end{ocaml}
-\caption{An OCaml interface to \C{Set} abstract data type}
+\caption{An OCaml interface to \C{Set} abstract data type.}
 \label{fig:ocaml-set}
@@
   \end{subfigure}
-\caption{Anamolous executions resulting from the Asynchronous
+\caption{Anamolous executions from asynchronous
   replication of \C{Set}. Dashed lines denote effect propagation.}
@@
 asynchronously applied at remote replicas as shown in the figure,
-resulting in the divergent states at $R_1$ and $R_2$.
+causing divergent states at $R_1$ and $R_2$.
 
@@
 \C{Set} RDT where concurrent (re-)adds consistently \emph{win}
-over removes. Such an \emph{add-wins} set is useful, for instance, to
-implement a shared shopping cart where two users can concurrently
-remove an item from the cart, but if one of them re-adds it then the
-item should be present in the final cart\footnote{This is in fact the
-semantics of Amazon's shopping cart~\cite{dynamo}}.
+over removes. Such an \emph{add-wins} set is useful, for instance, in
+an online shopping cart where two users can concurrently
+remove an item, but if one of them re-adds it, then the
+item is present in the final cart\footnote{This is in fact the
+semantics of Amazon's online shopping cart~\cite{dynamo}.}.
 
@@
 \emph{add-wins} \C{Set} interface is similar to Fig.~\ref{fig:cc-set},
-except that it requires an additional data structure to track
+except that it requires an additional data structure for
 element-wise vector clocks: 
@@
 their (lowest) common ancestor version. In our running example, the
-state is a value of type \C{Set.t}, hence \C{Set.merge} function would
-have the type signature:
+state is a value of type \C{Set.t}, hence \C{Set.merge} would
+have the signature:
 \begin{center}
